Single-cell spatial mapping reveals reproducible cell type organization and spatially-dependent gene expression in gastruloids

Version 1

Abstract

Gastruloids are three-dimensional stem-cell-based models that recapitulate key aspects of mammalian gastrulation, including formation of an anterior-posterior (AP) axis. However, we do not have detailed spatial information about gene expression and cell type organization, particularly at the level of individual gastruloids. Here, we report a spatially resolved, single-cell molecular catalog of the transcriptomes of 26 individual gastruloids. We found that cell type composition and spatial organization were remarkably consistent across gastruloids. Posterior cell types formed distinct, organized clusters, while anterior cell types were more disorganized. To distinguish progressive differentiation from cell type differences, we developed the L-metric, a parameter-free quantification of mutually exclusive gene expression. This analysis revealed spatial organization without explicit encoding, recapitulated known cell type relationships, and identified novel gene expression states and spatial subclusters within cell types. We confirmed that in gastruloids, NMP differentiation occurred through a continuous, spatially-coordinated process. We also showed that endothelial precursors exhibited unique spatial organization and had distinct gene expression profiles dependent on their association with anterior somitic or posterior endodermal tissues. This work enables the rigorous use of gastruloids as models for studying the molecular mechanisms underlying mammalian development and tissue organization, and introduces novel computational tools for analyzing spatially-resolved single-cell datasets.

Introduction

Embryogenesis involves not only the creation but, crucially, the organization of new cells and new cell identities. 3D gastruloids model an important stage of embryogenesis: gastrulation. Gastruloids break symmetry, elongate along an anterior-posterior (AP) axis, and form many of the cell types found in embryos at a similar developmental stage, as shown by single-cell RNA-seq analyses [1–6]. These characteristics mean they could hold enormous potential to enable us to deeply characterize, perturb, and test developmental hypotheses, circumventing many practical and ethical constraints imposed by using in vivo models. Yet despite extensive single-cell characterization of gastruloids, we still lack detailed information about the arrangement of cell types within individual gastruloids.

Sequencing-based gene expression analyses of gastruloids have at most been able to generate coarse-grained maps of the spatial arrangement of the constituent cells via sectioning [2,7]. These methods cannot resolve the position of individual cells, nor can they measure how tissues are organized other than along the AP axis. Computational methods that measure gene expression and identity of individual cells can only estimate AP axis location through inference methods and do not measure organization in any other dimension [8,9]. Additionally, detailed comparative spatial analyses of individual gastruloids have not been performed, as the experiments either require pooling gastruloids or are low-throughput, such that there are insufficient samples to make meaningful comparisons. While some studies point to stringent reproducibility of gene expression along the AP axis in gastruloids [10,11], others show that there is great variation, gastruloid-to-gastruloid, in the organization of specific tissues [12]. Without high-resolution spatial characterization across multiple gastruloids, it’s unclear what aspects of tissue development can be effectively modeled with this system.

The cell types present in gastruloids have largely been inferred from single-cell sequencing analyses. As is standard for these experiments, cells are clustered on the basis of their overall gene expression profiles, and types are assigned based on a qualitative assessment of the genes differentially expressed by the cluster [2,7,13–15]. However, it is unclear whether discrete binning cells into categories is the only, or even the most useful, approach to single-cell studies of gastruloids, a system nearly entirely composed of cells in the process of differentiating and transitioning between identities. A more gene-centric analysis, especially one coupled to a spatial measurement, could more naturally represent these transitions and how they are coupled to the arrangement and physical location of cells. Whether the diversity of cell types observed in sequencing experiments truly maps onto spatially organized collections of cells, and if this organization is consistent between gastruloids, remains an open question.

To address these gaps, we developed a spatially resolved, single-cell molecular map of the location, identity, and gene expression of cells within 26 individual gastruloids. We found that despite some morphological variability, elongated, polarized gastruloids had largely reproducible cell type composition. The order of types and expression of individual genes along the AP axis was also quite consistent. Leveraging our spatial single-cell expression data, we found that the posterior of the gastruloids was characteristically patterned with distinct clusters of progenitors and differentiated cell types. We present novel computational techniques useful for identifying genes with mutually exclusive expression patterns, which illuminated the spatial component of a bipotent progenitor differentiation decision and revealed a spatial hierarchy of gene expression patterns. Using evidence generated from this metric, we show evidence of spatial and functional control of endothelial cells, suggesting that 3D gastruloids may be used in the future to model early stages of vasculogenesis and organization.

Results

Cell types’ locations and relative proportions are consistent across morphologically normal gastruloids

To measure the spatial distribution of gene expression, we prepared gastruloids using mouse E14TG2a cells and a standard protocol (see Methods). We harvested mature gastruloids after 120 hours of growth and assayed with seqFISH [16]. There was substantial variation in the length, width, and relative amounts of anterior and posterior tissues in the final dataset, although all were within the range of what would be considered a ‘morphologically normal’ gastruloid [1,10]. We assigned an AP axis to each gastruloid using the expression of T, a canonical marker for the posterior (Figure 1a). When we compared how gene expression varies along the AP axis, we saw good agreement at a coarse-grained level with a previous study that sectioned gastruloids along the axis and analyzed gene expression in each section [2] (Figure S1.3a).

Cell types’ locations and relative proportions are consistent across morphologically normal gastruloids

a. Representative gastruloid with the expression of T shown in magenta. Each spot is one transcript (count), nuclei are shown in gray. Thick dashed line denotes the AP axis, thin lines outline the anterior and posterior halves of the gastruloid. Scale bar is 200 μm. b. Gallery of representative gastruloids. Scale bar is the same as a. c. Proportion of each cell type across all gastruloids, ordered by median proportion. d. Correlation between the ratio of anterior area to posterior area across gastruloids and the proportion of that gastruloid typed as either somite (left) or paraxial mesoderm (right). e. Smoothed density of each cell type along the AP axis. Nuclei in each gastruloid were projected onto the AP axis, and their position was normalized to the total length. Traces from individual gastruloids are shown by individual curves. f. Distribution of mixing coefficients across all gastruloids and examples of low, medium, and high mixing. Scale bar is 200 μm.

We assigned a cell type to each nucleus using a cell type scoring method with known marker genes. Unsupervised clustering produced similar results (Figure S4.4a), but we achieved greater cell type distinction when we scored using a marker gene panel; see Methods for additional details. A representative gallery of typed gastruloids is shown in Figure 1b; the full dataset is in Figure S1.2. Because each cell received a cell type score for each type, we could use the entropy of the cell type score probability distribution to assess confidence of our cell type assignment: a cell that received a similar score for multiple cell types would have a high entropy distribution, while one which scored highly for one type and low for the rest would have low entropy. On average, the cell type entropies for most cells in a given cell type were low, with the exception of paraxial mesoderm and cardiac mesoderm cells, which had intermediate values (see additional discussion below). The generally low entropies indicate that most of our cell type assignments were high-confidence (Figure S1.1a).

Once we had the cell type identity and spatial location of each cell in all the gastruloids, we characterized the organization of each by mapping where each cell type was found relative to other types and overall morphology. The approximate location of each cell type relative to gastruloid morphology was quite consistent: the posterior region, although highly variable in size (Figure S1.2a,b,c), mainly consisted of neuromesodermal precursors (NMP, turquoise), a bipotent cell type that contributes to both neural and mesodermal tissues [17–19]. This localization is consistent with previous observations of the expression of NMP genes in the gastruloid posterior [2,7,20]. However, unlike these previous studies, which inferred NMPs from the expression of single genes and could not resolve other cell types, we were able to see that NMPs were intermixed in a patchy fashion with mesodermally-fated cells that are on the path to forming somitic tissues (presomitic mesoderm, PSM, red). In some instances, we saw PSM cells forming symmetric patches on either side of a central neurally-fated region (Figure 1d, center gastruloid, Figure S1.2a v., xi., xii, b iv), mimicking the organization found in embryos [21–23] and thought to occur in gastruloids [8,24]. Generally, we saw 1-2 distinct clusters of neurally fated cells (spinal cord precursors, black), often adjacent to or opposite the region associated with somitic fate acquisition (“differentiation front”, light blue), although occasionally found in the center of the gastruloid (Figure S1.2a ii., vi., xvii). Although the distinct organization of neurally fated cells has been observed in other embryo models [3], the organization of spinal cord precursors in gastruloids in the context of other cell types has not yet been systematically characterized; staining for spinal cord-associated genes showed compelling patterning, but the patterns varied by gene, complicating interpretation of the organization of the tissue [4]. Consistent with [2], the differentiation front was nearly always found in one distinct cluster marking the boundary between the anterior and posterior (Figure S1.2a i., iii., vi., b iii., vi). The anterior regions, although clearly not organized into somites, consisted mainly of somitic cells, presomitic cells, and cardiac mesoderm (yellow, green, and purple), with highly variable mixing (Figure S1.2 a iii., v., x., xv., xvi) . Also present in the posterior of almost all the gastruloids were small clusters of endothelial precursors (pink) (Figure S1.2 a iii., vi., vii., xvii., c), in some cases associated with a cluster of endodermal cells (blue) (Figure S1.2 a x., xi., xiv., xv). In general, if an endodermal region was present, it was located in the center of the gastruloid (Figure S1.2 a v., viii., xii., xvii., b ii., iv., c), consistent with observations in [12]. Although there was clearly variation in the details of cell type arrangement, there were obvious organizational principles that were conserved between gastruloids independent of size, consistent with the fidelity of gene expression along the AP axis reported in reference [10]. This simultaneous mapping of all cell types is consistent with many observations in the literature about the expression of individual genes and integrates them into one consistent picture of cell type organization.

We sought to quantify variability in cell type composition between gastruloids. Early single-cell datasets relied on pooling multiple gastruloids, thus obscuring the degree to which the overall cell type distribution was reflected in each individual gastruloid. However, recent single-cell measurements of individual gastruloids have suggested substantial gastruloid-to-gastruloid variation in cell type proportions [13]. Figure 1c shows distributions of cell type proportions across samples. Figure S1.1b shows the per-gastruloid distributions. In general, we found that cell type proportions did not vary significantly between gastruloids, with the exception of endoderm, paraxial mesoderm, and somite. Unlike other cell types that were consistently present but varied in proportion, endoderm variation largely resulted from the absence of this tissue in some gastruloids, reflecting the developmental variability of endodermal tissues previously documented in gastruloids [12]. The variation in the proportion of somite cells could be at least partially explained by variability in the amount of anterior tissue relative to posterior tissue (Figure 1d, left), and given that somitic tissue forms progressively over the course of gastruloid development, could potentially be explained by desynchronization of somite formation such that some gastruloids had, at 120 hours, formed more of this cell type which emerges later in gastruloid development [13]. We did not observe gastruloids that were as strongly neurally-biased as those reported in [13], but we did see some gastruloids with a relatively high proportion of spinal cord precursor cells (Figure S1.2a ii., xv., b vii), and overall the proportion of spinal cord had a negative covariation with the mesodermally-derived cell types, consistent with the anticorrelation also reported in [13] (Figure S1.1c).

The proportion of somite cells was strongly positively correlated with the proportion of presomitic mesoderm cells (covariation = 0.63, Figure S1.1c). Presomitic mesoderm cells progressively take on somitic fate at and after the differentiation front; a positive correlation between these two would suggest that the amount of presomitic mesoderm is linked to how much somite is ultimately formed. How much presomitic mesoderm is formed depends on fate decisions made by NMPs: NMP progeny can become either mesodermal or neural, depending on their local signaling environment. This chain of correlation suggests that fate decisions made by NMPs may drive the overall composition of individual gastruloids and that, as in embryos, a balance between the amount of somitic and mesodermal tissue is achieved, although likely to a less stringent degree [25]. Further experiments could delineate whether, as in embryos, Wnt3a drives this balance [25], and the degree to which growth of differentiated progeny affects compositional variation [20].

Finally, paraxial mesoderm variation could only be weakly explained by morphological characteristics (Figure 1d, right), and couldn’t be explained by covariation with any other cell type (Figure S1.1e). Given that paraxial mesoderm also has a relatively higher average cell type entropy score, it suggests either that our marker genes for this cell type were not sufficient to confidently assign a score, that their gene expression is more variable than other types, or that this type itself is transient in gastruloids, as has been suggested in the literature [26].

Cell types’ arrangement is consistent across gastruloids, but varies by type

Given that the proportions of cell types within each gastruloid were fairly consistent, to what degree did their spatial organization vary? We first projected each cell’s expression onto the AP axis and looked at the distribution of AP axis locations across gastruloids (Figure 1e). The most posterior cell types (NMP and presomitic mesoderm) showed consistent, wide distributions from 0-30% of the axis. Centered around 30%, spinal cord precursors and endoderm had distinctive peaks (clusters), the exact location of which varied between gastruloids. The differentiation front was consistently located in one peak between 30-40% of the AP axis. The anterior cell types tended to be more broadly distributed without distinct peaks; however, endothelial precursors in particular had wide, multimodal distributions that varied substantially between gastruloids despite a comparably small proportional variation (Figure 1c). Generally speaking, variation in the proportion of a cell type across gastruloids was not directly predictive of the variation in location of that cell type relative to the AP axis. From these observations, we concluded that the amount of any given cell type is not the sole determining factor in where it is located along the AP axis.

Several studies of gene expression in gastruloids have used pooled measurements to infer the AP axis-location of genes and cell types [3,7,13,24,27] and our data are largely consistent with these lower-resolution findings (Figure S1.3a). Yet it is obvious from individual gene staining [1,4,8,28] and our detailed 2D maps of cell identity and location that gastruloid organization is much more complex than the average order of cells along the AP axis. We thus needed an analytical method for quantifying spatial organization beyond distributions along the AP axis. To characterize spatial organization, we quantified the overall mixing of cell types within each gastruloid using a mixing coefficient that quantifies how often each cell is found next to a cell of a different type or the same type. These neighbor relationships are summarized into a single score that varies from 0 (completely mixed) to 1 (completely segregated) (Figure 1f). The mixing coefficient values range from 0.29 to 0.58 (Figure 1g). With the exception of cardiac mesoderm, the amount of any given cell type did not significantly correlate with the mixing coefficient (Figure S1.1d). The proportion of cardiac mesoderm negatively correlates with mixing, implying that this cell type is particularly likely to be found as a singlet.

Our spatial transcriptomics is imaging-based, and the fact that we image transcripts in a single plane admits the possibility that, in any individual gastruloid, we may collect data from a different part of the gastruloid. We controlled for this to the extent possible within experimental limitations by imaging multiple gastruloids across several experiments and keeping our imaging parameters, particularly the instrument z-depth relative to the coverslip, nearly identical across experiments. The relatively wide distribution of mixing coefficients demonstrates that even within gastruloids with broadly similar morphologies and cell type proportions, the underlying organization of cell types can vary substantially.

Pairwise cell type interactions suggest more mixing of cell types in the gastruloid anterior than posterior

Our simultaneous observation of variability in mixing coefficients and consistent cell type organization along the AP axis led us to ask whether mixing coefficients were driven largely by local topology and whether this effect was disproportionately due to the arrangement of a subset of cell types. The interaction between long-range signaling gradients and short-range cell-cell interactions is thought to drive gastruloid symmetry breaking and patterning [29–31], and our spatial data allowed us to query the cell-cell level organization and interaction between cells. First, we asked to what degree each cell type is exposed to all other cell types (including those that could not be typed). We calculated an “exposure index”, which quantifies the degree to which each cell type (source) is exposed to another cell type (neighbors). It is similar to the mixing coefficient used to score the overall organization of a gastruloid, but differs in that the mixing coefficient only measures differences in self vs. non-self interactions between cell types, whereas the exposure index tracks interactions between all pairs of cell types. The measure is asymmetric in that the exposure of cell type A to B may not be the same as the exposure of cell type B to A; for example, a cell type found in singlets within a larger tissue will be more frequently exposed to cells of that common tissue type, while the common type will see itself much more frequently than the intermixed cell. Thus, this asymmetry accounts for the fact that local patterning can influence the degree to which any one cell type is found next to another. Asymmetry in off-diagonal entries in the exposure matrix (Figure 2a), therefore, can reflect local patterning. For instance, somite and endothelial precursors have an asymmetric pattern: although endothelial cells are often exposed to somite cells (exposure index 0.207), somite cells are comparatively infrequently exposed to endothelial cells (0.07). This observation, coupled with a fairly strong endothelial-endothelial exposure index (Figure S2a), suggests small clusters of endothelial cells are found among somite cells, reflecting our visual impressions. In contrast, cardiac mesoderm cells were often exposed to all other anterior cell types, but their self-exposure index was low, matching our observation that they are found in inconsistent and disorganized clumps or individually. Like paraxial mesoderm, this cell type also has a high average cell type score entropy (Figure S1a).

Pairwise cell type interactions quantify cell type mixing and interaction motifs

a. Segregation index, which represents how frequently the source cell type (row) is found next to the target cell type (column). The maximum possible value is 1, self-interactions excluded for readability. b. Significantly enriched and depleted triplet combinations of cell types, ordered by effect size from left to right. Bars are colored by the composition of the triplet they represent. Only triplets found in over half of the gastruloids are shown.

The overall interaction matrix (Figure 2a) revealed that there were more inter-type exposures between the anterior cell types of the gastruloid than between the posterior types. This quantification matches the impression from visual inspection that the gastruloids contain both somitic cells and paraxial mesoderm cells that properly localize to the anterior (as per [2,8]) but are arranged in a disorganized fashion. Cells of the posterior, by contrast, were much less mixed, with individual cell types forming large, distinct clusters. These results demonstrate that our analytical methods were able to capture and quantify the degree of spatial disorder in these two compartments. The contrast between organized posterior clustering and disorganized anterior mixing matches expectations based on literature that shows that self-organization mechanisms in gastruloids in the anterior vs. posterior are differentially sensitive to culture conditions, with somitic patterning requiring external matrix support [1,4,8,24], distinguishing it from the seemingly more autonomous organization observed in posterior cell types.

Local topology of cell type interactions is preserved across gastruloids

Our analyses have been focused on pairwise interactions, but these data also have the potential to reveal higher-order interactions between cell types. To measure such interactions, we first determined all the possible combinations of three cells (triplet motif) from the 9 cell types we had identified (with replacement). Then, for each cell in the gastruloid, we identified the two nearest neighbors and determined which triplet type it was. We calculated the enrichment of each triplet motif relative to its expected frequency based on the amount of each of the constituent cell types. Figure 2b shows this distribution of triplet motifs that were found in over half the gastruloids and in each were significantly enriched or depleted based on overall cell type frequency. Figure S2b shows the same distribution but for triplets that were found in any gastruloid. We found that motifs formed from 3 of the same cell type were overwhelmingly the most significantly enriched, and all were found in over half of the gastruloids. The enriched 2+1 (two same) motifs were composed of all posterior cell types or all anterior cell types. 3/4 of the most common significantly enriched motifs that were of three different cell types consisted of cardiac mesoderm with two other anterior cell types or differentiation front, again suggesting that the cardiac mesoderm cells are uniquely spread out as individuals.

We also identified motifs that were highly underrepresented compared to what would be expected given the frequency of occurrence of each of the individual cell types: nearly all of the significantly underrepresented motifs were of 3 different cell types, with the rarest being somite+NMP+presomitic mesoderm. This observation was initially surprising given that these cells are all thought (in gastruloids) to derive from the same lineage; however, the early mesodermal precursors are spatially segregated from the anterior by the differentiation front, which is where somite determination happens [2,3,24] The relative paucity of this triplet indicates that the spatial organization of the differentiation process occurs with very high fidelity, and that less differentiated precursors rarely make it into the anterior, where they would be exposed to somite and paraxial mesoderm cells.

Overall, these data speak to the consistency of gastruloid organization, although broad-scale morphological differences can explain some differences in cell type proportion and rare individual gastruloids may have unique cell type arrangements. In general, morphologically normal gastruloids consistently patterned posterior cell types and moved more differentiated cell types to the anterior. Along the way, the granular interactions between individual cells were reproduced with remarkable fidelity. It is notable that this arrangement across length scales was achieved with only minimal external chemical cues.

Co-expression of cell type marker genes delineates progressive NMP differentiation in gastruloids

One of the central goals of studying developmental processes is to determine the means by which cells make fate decisions and when these decisions are driven by the physical location of the cell as opposed to cell-intrinsic programs. A large part of gastruloid development is thought to be driven by the differentiation, growth, and migration of NMPs and their progeny, which are fated towards either a mesodermal or posterior neural (spinal cord) fate [3,28,32]. In embryos, this process is thought to happen in a continuous fashion; NMP progeny simultaneously move anteriorly and acquire either a mesodermal or spinal cord fate [18,25,33,34]. The differently fated cells are driven to different parts of the developing spinal column; lineage tracing evidence suggests that fate commitment is at least partially driven by physical location and the local signaling environment [25], and is progressive in nature [33]. We wondered how this coupling between space and differentiation was resolved in gastruloids, and whether the continuous transition from progenitor to more differentiated cell type was reflected in spatially continuous patterns of gene expression.

Our data represent a single slice of time during which NMPs, presomitic mesoderm, and spinal cord precursors exist simultaneously; if this process happened continuously, we would also expect to see a subset of cells somewhere on an identity continuum between these types (Figure 3a). For simplicity, when initially assigning cell types, we used differential expression of panels of marker genes to assign identity. However, many of the cells in the posterior had non-zero cell type scores in all three categories (Figure S3.1a). This observation motivated us to leverage the advantages of our dataset to delineate where individual genes were expressed in space with single-cell resolution and determine whether that additional level of resolution gave additional information about NMP differentiation trajectories.

Progressive NMP differentiation is revealed by the single cell L-metric

a. Illustration of NMP differentiation with predictions for where along the trajectory gene expression is shared or unique. b. Co-expression plot showing the relative amount of Eogt vs Pax6 (left) and Eogt vs Rfx4 (right) per cell in a representative gastruloid. Pax6 and Rfx4 are both annotated as spinal cord-associated genes. c. Per cell expression of all NMP exclusive genes summed versus all other gene categories summed, one plot per type. n=25514 cells typed as NMP, presomitic mesoderm, or spinal cord in n=26 gastruloids d. Per cell expression of all spinal cord exclusive genes summed versus expression of all presomitic mesoderm genes summed. n=25514 cells typed as NMP, presomitic mesoderm, or spinal cord in n=26 gastruloids. e. Per cell expression scatterplots of the two pairs of genes shown in b). The y axis of each is the per-cell expression of Eogt. The x-axis is the per-cell expression of Pax6 (left) or Rfx4 (right). R is Pearson’s r, scL is scL-metric. f. Distribution of scL-metric values for pairs of spinal cord and presomitic mesoderm genes (left), NMP and presomitic mesoderm genes (center), and NMP and spinal cord genes (right). g. Hierarchical clustering heatmap of scL-metric vectors for NMP, presomitic mesoderm, spinal cord, and combined category genes. Blocks highlighted are discussed in the text. h. Tree of the hierarchical relationships between genes resulting from the clustering shown in e). Color indicating the gene type is shown at the bottom, legend is the same as in g. The density plots shown are the summed, averaged gene expression of all genes in the leaves up until that node, smoothed with a 2D density kernel estimate (see Methods for details about smoothing).

Since both presomitic mesoderm and spinal cord cells derive from NMPs, we reasoned that it was possible, especially if the cells were still in the process of differentiating, for cells typed as presomitic mesoderm or spinal cord to residually express NMP genes. Indeed, there were genes in our panel that were shared between each type and the precursor (referred to here as NMP+presomitic mesoderm genes and NMP+spinal cord genes or “shared genes”). Figure 3c shows the total expression of each gene group versus the NMP genes for all cells in all gastruloids that we typed as NMP, presomitic mesoderm, or spinal cord. As expected, there was a clear correlation between the mixed categories and NMP genes, supporting the notion of a continuous differentiation process.

In an effort to locate cells that had fully committed to one lineage or the other, we looked at the expression of genes thought to be exclusively associated with presomitic mesoderm or spinal cord. Our expectation was that the expression of these genes would be mutually exclusive, and several genes did display this pattern (Figure 3b, right). However, we were surprised to find many examples where these supposed marker genes for different cell types were expressed in the same cell (Figure 3b, left). Spot assignment in imaging data has the potential to mis-assign transcripts, which could, in principle, lead to artifactual mixed states. However, we don’t believe misassignment explained our observations. We used stringent nuclear dilation (small enough that ∼30% of transcripts were not assigned to a nucleus). Moreover, when we calculated the exposure index, we found very little exposure between presomitic mesoderm and spinal cord cells (0.07 and 0.04, Figure 2a), meaning the majority of the time these cells were not nearest neighbors, and thus had little potential for overlap and transcript mis-assignment.

Was this trend of co-expression true for the two groups of marker genes as a whole? When we compared the total per-cell expression of the genes in both categories (spinal cord and presomitic mesoderm), we noted that while they did show more mutually exclusive (i.e. L-shaped) expression than when compared to the NMP genes, there were clearly many cells that had substantial counts from both PSM and spinal cord gene categories (Figure 3d). One possible explanation for the co-expression of genes associated with both arms of the differentiation trajectory is that these cells had not yet fully committed to either lineage. It is also possible that a subset of the genes were better (or at least more exclusive) markers than others (as our observation of the heterogeneity in expression patterns suggested), and that there may in fact be a continuum of markers, some of which may distinguish certain cell types and not others. To determine if any of these possibilities were true, we needed a way to identify genes that were expressed in a mutually exclusive manner.

The L-metric quantifies mutually exclusive gene expression

Marker genes for distinct lineages are often assumed to be mutually exclusive in their expression. However, the above results show that such assumptions may not hold. We were curious whether there were genes that were truly exclusively associated with each cell type, and if such genes in general would prove to be more effective marker genes than those that were merely associated with the cell type.

To this end, we developed a pairwise metric between genes that reported the degree of mutually exclusive expression. It is calculated by rank ordering cells by the expression of one gene and measuring the degree to which the expression of the other gene is anti-rank-ordered (see Methods for details). We call this metric the “single-cell L-metric” (scL-metric) because when the per-cell expression of mutually exclusive genes was plotted against one another, the data made an L shape (Figure 3e, right). A value of -1 represents perfectly mutually exclusive expression, which only happens when the genes are never found in the same cell. Higher values indicate more co-expression. The maximum possible value is 1, which is obtained when both genes express in exactly the same cells, independent of the total amount. The scL-metric is inherently asymmetric: if the two pair orderings produce very different L values, this indicates that there is a corresponding asymmetry in their expression. For example, all counts of the first gene may be found in cells expressing the second gene, yielding an scL-metric closer to 1. However, the second gene may be expressed in many cells with no expression of the first gene, and, thus, swapping the pair ordering results in an scL-metric value closer to 0. In most cases, the values were similar, so we averaged them for simplicity, thereby symmetrizing the metric. The scL-metric is parameter- and threshold-free, which is an advantage over methods that require thresholds to binarize data before calculating an odds ratio or mutual information.

Our expectation is that ubiquitously expressed genes like cell cycle and housekeeping genes will have relatively high scL-metric values no matter which genes they are compared with, whereas genes that are specifically associated with a single cell type will have an scL-metric value close to -1 when compared with genes specific to other types, but higher values when compared with genes associated with the same cell type. An example with two NMP genes (Cdx2 and Cdx4) with a high degree of co-expression (scL-metric=0.41) is in Figure S3.1e. In the case of NMP differentiation, we could use this metric to determine whether there are pairs of genes that definitively delineate whether a cell has spinal cord or mesodermal identity, and furthermore, if there are genes that distinguish these more differentiated cells from the parental cell type.

We compared the distribution of L-metric values between spinal cord and presomitic mesoderm genes; as our previous results suggested, there were many genes that had very strong mutually exclusive expression, while others were expressed in the same cells. For example, Pax6 (spinal cord) and Eogt (presomitic mesoderm) are often expressed in the same cells; they have a mild positive correlation as measured by Pearson’s R (0.24), and an scL-metric of -0.55 (Figure 3e, left). However, Rfx4 (spinal cord) and Eogt (presomitic mesoderm) were expressed mostly in distinct cells; their expression was nearly uncorrelated when measured by Pearson’s R (-0.02), but they have an scL-metric of -0.96 (Figure 3e, right). Thus, Rfx4 appears to be a good marker for differentiated spinal cord cells, while Pax6 might mark cells that are neurally fated but not yet committed. We also note that it is not possible to distinguish from these analyses whether all differentiated spinal cord cells express Rfx4, merely that it is exclusively associated with this type.

Overall, the distribution of spinal cord and presomitic mesoderm gene sc-L-metric values was strongly skewed towards -1, with at least 20 pairs of genes strongly distinguishing between the two types (Figure 3f, left). From the 54 possible pairs of spinal cord and presomitic mesoderm genes, we found that 24 (44.4%) had scL-metric values < -0.9, likely indicating that they mark different populations of cells.

We predicted that there would be more cell-wise co-expression (i.e., higher scL-metric) between NMP genes and the other two categories (presomitic mesoderm and spinal cord), since these two types of cells derive from NMPs. When we compared the distribution of scL-metric values of NMP and spinal cord or NMP and presomitic mesoderm gene pairs, we did in fact see a shift of the distribution to higher values compared to the presomitic mesoderm and spinal cord pairs, indicating more co-expression (Figure 3f, right).

Despite this shift to higher values, some pairs still seemed to delineate the cell types. For example, Fgf9 (presomitic mesoderm) and Wnt3a (NMP) have a very low (L = -0.84). There were even more low-L pairs when comparing NMP and spinal cord genes, such as Sox21 and Wnt3a (L=-0.98) and Scube2 (spinal cord) and Cdx1 (NMP) (L=-0.91). Notably, the average scL-metric of presomitic mesoderm and NMP gene pairs was much higher than spinal cord and NMP gene pairs, which suggests that perhaps commitment to a neural pathway involves a more systematic reorganization of gene expression.

Analysis using this metric allowed us to distinguish genes such as Eogt and Rfx4 that clearly mark distinct subsets of cells, from those like Eogt and Pax6 that partially overlap and may mark intermediates as well as more differentiated cells. We also observed a unique pattern when we compared the expression of Rfx4, which we identified as a good marker of spinal cord genes, with Nkx1-2, a shared gene annotated as being associated with both NMPs and spinal cord cells. The expression of Nkx1-2 was strongest in the posterior, but extended into the region marked by Rfx4. It seemed to be progressively downregulated in this region from posterior to anterior, producing a graded expression pattern with the anterior-most cells only expressing Rfx4 (Figure S3.1c, d). This observation suggests that shared genes can mark cells on their way to becoming spinal cord cells, and at least some are turned off as the cells differentiate, suggesting that the continuous differentiation pathway observed in embryos is also found in gastruloids.

These graded expression patterns led us to be more broadly interested in the patterns of spatial expression and scL-metric values of genes that were annotated as belonging to more than one cell type. Up until now, we had excluded such genes from our analyses, but we reasoned that their distribution and co-expression might reveal more about the path that NMPs take on their way to becoming presomitic mesoderm or spinal cord cells. Shared NMP+presomitic mesoderm genes included T, Fgf17, and Hoxaas3, while shared NMP+spinal cord genes included Sox2, Gbx2, and Nkx1-2. When we compared both shared categories to their terminal cell type, we found on average that the distributions of L-metric values were positively skewed compared to the NMP and terminal cell type distributions, indicating that, as expected, there was even more co-expression among these categories (Figure S3.1b). Interestingly, we found that both distributions were bimodal: a subset of genes were clustered around 0, indicating high co-expression with the terminal cell type genes (this included Dll1 (presomitic mesoderm) and T (NMP+presomitic mesoderm), and Cdh6 (spinal cord) and Nkx1-2 (spinal cord+NMP). Others were closer to -1, meaning the gene in the shared category is either more associated with the parental type (NMPs) or potentially is specific to intermediates. Examples of these pairs include Mesp2 (presomitic mesoderm) and Msgn1 (presomitic mesoderm+NMP), and Rfx4 (spinal cord) and Hoxb6 (spinal cord+NMP).

We wondered whether we could use the scL-metric scores to refine our marker gene panel. We reasoned that ‘good’ marker genes would have very low scL-metric scores with most other genes and have high scL-metric values with a small subset of genes (presumably those belonging to the same type). However, filtering on these criteria did not markedly improve our cell type scoring, likely because the genes in our panel were hand-selected to be good type representatives, so most fit this criterion (Figure S4.3a, b and Methods). However, we found that scL-metric analysis could be used to quantify how related cell types were in terms of marker gene expression (Figure S4.3c), and moreover was extremely effective at assessing the outcome of unsupervised clustering, including quantifying overlap in gene expression between clusters and identifying gene pairs that were mutually exclusively expressed between clusters, even when they were highly related (Figure S4.4).

This analysis demonstrates that we could use the scL-metric to delineate, in an automated fashion, whether genes reported in the literature as associated with multiple clusters in single-cell data were truly associated with multiple cell types, without parameter tuning and using only the cell-by-gene table.

The L-metric captures the spatial distribution of gene expression despite being calculated without spatial information

We noticed that sometimes, even genes that had been annotated as belonging to the same cell type showed mutually exclusive expression (scL-metric = -1) (Figure S3.1b). We noticed that the spatial distribution of expression of the genes in these pairs was often quite distinct, and often did not match the spatial cell type distribution as well. Thus, we wondered whether an overall spatial organization, compatible with but independent of cell typing, would emerge from clustering genes based on the degree of mutual exclusivity of their expression.

For each gene, we calculated a vector of scL-metric values with all other genes, effectively cataloging the extent of mutual expression across our dataset. We clustered these vectors; a heatmap of this clustering is shown in Figure 3g. Broadly speaking, the genes associated with each cell type clustered together, and the shared category genes generally clustered closely to the NMP genes. Interestingly, although both the NMP and spinal-cord exclusive genes were each found in one large type-specific cluster (Figure 3g ii., v.), the presomitic mesoderm exclusive genes were split into two groups (Figure 3g i., iv.). There was also a subset of NMP-exclusive and spinal cord-exclusive genes that clustered together and separately from the rest of their group (Figure 3g iii.).

To determine what could be driving these differences, we plotted where the genes were expressed in space. Since the clustering was hierarchical, we first made a tree representing this hierarchical relationship, and then at each node of the tree, we made a summed density plot of all the genes contained below that node. The density plots at the top of the tree show broad distributions of expression, but as we followed the branches down closer to the leaves, we saw fascinating spatial patterns begin to emerge. We were surprised to find that the L-metric, which doesn’t inherently contain any spatial information, was able to capture a spatial tree of gene expression.

Most of the NMP-exclusive genes were found very close to the posterior and were diffusely expressed. However, in the mixed cluster, which contained both NMP-exclusive and spinal cord-exclusive genes, we saw a different pattern — here the NMP genes were more patchily expressed and extended far more anteriorly (Figure 3h i..), suggesting that genes in this group may mark cells destined to become part of the spinal cord. Indeed, one such gene was Epha5, an ephrin receptor differentially expressed in a cluster labeled as NMPs by single-cell sequencing [2] but which has separately been annotated to be involved in neural development [35] We also saw a clear division between more anteriorly and more posteriorly expressed spinal cord genes (Figure 3h ii..) Thus, clustering L-metric values allowed us to delineate differences in behavior between cell type-associated genes: one cluster of spinal cord and NMP genes appears to mark neurally fated but uncommitted cells (Figure 3h i..), while another marks committed spinal-cord fated cells (Figure 3h ii..).

We also looked more deeply into the split presomitic mesoderm exclusive genes. Not only were they expressed in different parts of the gastruloid (compare Figure 3h iii., iv.), we realized that these genes were associated with presomitic mesoderm cells in different states: Mesp2 is turned on in a specific and spatially constrained location at the onset of somitogenesis [3], a process which has only just started in this gastruloid, while Dll1 more generally marks the presomitic mesoderm. Thus, using only differences in scL-metric values, we were able to identify both general presomitic mesoderm versus presomitic mesoderm that was starting to undergo somitogenesis.

These analyses demonstrate that information contained within the hierarchical relationships between genes determined by scL-metric reveals novel information about cell states within cell types and can identify distinct spatial locations of cells in this cell state, all without explicit encoding of spatial information, but rather quantifying and clustering the degree to which genes are mutually exclusively expressed with one another.

Clustering scL-metric vectors clearly resolves cell types and reveals novel genetic interactions

Given the amount of spatial and state information that was encoded in the scL-metric heatmap for a subset of our gene panel, we expanded our analyses to all genes, hoping to discover new genetic interactions or refine existing ones. We first calculated the scL-metric value for all genes in all gastruloids, then averaged across gastruloids and clustered the resulting interaction vectors (see Methods for details). The heatmap is shown in Figure 4a (heatmap including cell cycle genes is shown in Figure S4.1a). We noted that just as when we clustered genes associated with NMPs and their direct descendants, genes associated with cell types tended to cluster together. Specifically, NMP, spinal cord, endoderm, and endothelial genes clustered very strongly together, while presomitic mesoderm genes again were split into two groups, one of which was more closely associated with genes involved in early somitogenesis.

Clustering L-metric vectors clearly resolves cell types and reveals novel genetic interactions

a. Heatmap of all scL-metric values for all genes in the panel (excluding poorly detected and cell cycle genes, see Methods, n=166). Colored bars on the top and right hand sides indicate if a gene is associated with a particular cell type. Heatmap was hierarchically clustered by row. b. Expansion of a presomitic mesoderm cluster (i.). Clustering relationships are indicated with the dendrogram, and summed densities for all genes in an example gastruloid show where the genes in each cluster are expressed spatially. Clustering relationships determined from the full gene set shown in a. c. Expansion of the posterior cell type cluster (ii.). Clustering relationships are indicated with the dendrogram, and summed densities for all genes in an example gastruloid show where the genes in each cluster are expressed spatially. Clustering relationships determined from the full gene set shown in a. d. Expansion of the endothelial cluster (iii.). Clustering relationships are indicated with the dendrogram, and summed densities for all genes in an example gastruloid show where the genes in each cluster are expressed spatially. Clustering relationships determined from the full gene set shown in a. e. Expression of two example pairs of genes from the endothelial cluster: Cldn5 and Tgfb1, and Cldn5 and Sox17.

We examined the largest cluster of presomitic mesoderm genes (Figure 4a, i..). Just as before, the clustering reflected distinct spatial distributions; a highlight of one diagonal block with two subclusters is shown in Figure 4b, along with the expression patterns in a representative gastruloid. The right subcluster has most of the presomitic mesoderm genes, and is particularly enriched in the left side of this gastruloid, despite the fact that there is a cluster of presomitic mesoderm genes on both sides of the central axis (Figure S1.2a). The other subcluster has genes associated with differentiation into somites, and contains Snai1, which is a gene associated with EMT and cells in a more migratory state, consistent with how somitogenesis is thought to occur in gastruloids [2,3,24]. These genes cluster together because both sets are strongly expressed in a central region of presomitic mesoderm, one is more posteriorly biased, while the other is more anteriorly biased.

The genes associated with more anterior cell types, like somite, paraxial mesoderm, and the differentiation front (which marks the boundary between posterior and anterior), also clustered together in one large block, although the genes associated with the different types tended to be mixed (Figure 4c). When we plotted their spatial distributions, the reason for this separation became clear — one block was genes more strongly expressed in the most anterior of the gastruloid, while the other, including almost all of the differentiation front genes, was expressed more posteriorly; these genes were associated with the onset of somitogenesis. We saw similar spatial patterning in the cluster of NMP genes. One sub-cluster, which contained genes like Wnt3a, Hoxc10, and Evx1, was very specifically far-posteriorly localized, while the other, which contained genes like Fgf8, Cdx4, and some of the shared NMP+presomitic mesoderm genes, extended further into the anterior (Figure S4.1b).

We also observed interesting differences between the subclusters of endothelial genes (Figure 4d). Cldn5 had the strongest average association with other endothelial genes and strong negative associations with genes associated with other cell types, indicating that it is likely the best marker in our panel for generically marking all endothelial precursors. The cluster that contained Cldn5 showed strong intra-cluster associations (high scL-metric values), and although the other cluster showed relatively high association with Cldn5, the average association within the cluster and with the genes in the other cluster was weaker. However, this cluster contained some interesting genes: Sox17, an endoderm marker, and Tgfb1 (Tgfβ), which is associated with reduction in inflammation, pro-growth signals, and differentiation [36]. Plotting the expression of these genes in representative gastruloids confirmed their co-expression with Cldn5, a consistent marker for endothelial cells (Figure 4e).

Tgfβ appeared to be exclusively associated with Cldn5-expressing cells, but only in some gastruloids (it was extremely lowly expressed in others and not clearly associated with any specific cell type). Tgfβ loss causes embryonic lethality due to hematopoiesis and vascular development failures [37], specifically in extraembryonic tissues. Extraembryonic endoderm has been inconsistently reported in gastruloids; some studies find evidence of this tissue [2], while others do not [13]. Moreover, Tgfβ can drive the transition from endothelial precursors to hematopoietic stem cells [38], so it is particularly exciting to see it specifically expressed in endothelial tissue. While we note that the regions expressing Tgfβ have unique morphology (Figure S1.2a, xv), further work is needed to determine the role this gene plays in the development of these tissues in gastruloids. These results demonstrate that the scL-metric can be used to identify novel cell states in subsets of cell types.

To validate the clustering produced by the L-metric, we compared our results to a state-of-the-art method for identifying gene programs in an unbiased fashion from single-cell data: consensus non-negative matrix factorization (cNMF) [39]. We pooled nuclei from all individual gastruloids and ran cNMF. We found that many of the resulting clusters (Figure S4.2a) corresponded to the clusters identified when the scL-metric tree was truncated to produce exactly the same number of clusters (Figure S4.2b). The similarities were even greater when the scL-metric clusters were hand-selected based on visual inspection of the tree and density of marker genes (Figure S4.2c). From these observations, we conclude that the two methods are capable of producing similar results at a high-level, but are different in their application. Individual cells receive component scores for cNMF gene programs, yielding more per-cell information, while the tree produced by L-metric clustering reveals hierarchical information about gene programs, which quantifies their similarity in expression on a more global scale.

Thus, by quantifying the similarity between genes based on the mutual exclusivity of their expression with all other genes, we were able to extract similarity in the expression of cell type genes without explicitly specifying them. Moreover, the spatial relationship between genes, including among genes associated with the same cell type, was encoded in the tree of relationships produced by clustering. Finally, we discovered novel functions of genes in the panel through their location in the scL-metric tree: although Tgfβ was initially included in our panel to generally detect inflammatory and growth signaling, clustering by expression patterns revealed its unique association with endothelial precursors. Together, these data demonstrate the power of mutually exclusive expression relationships to uniquely quantify relationships between genes and the cells that express them.

Calculating the L-metric on smoothed gene expression spatial profiles reveals higher-order tissue organization and unique cell type arrangement in certain gastruloids

Our analysis of mutually exclusive gene expression on a per-cell basis both yielded the tools to identify highly specific marker genes and revealed gene expression states that were characteristic of subsets of our annotated cell types. However, there are examples where graded gene expression in tissue, independent of the individual cells expressing the gene, is biologically meaningful: zonation in tissues such as intestine [40], liver [41] and kidney [42], and hot and cold areas of tumors [43]. Furthermore, many spatial analysis techniques are not resolved to a single cell. Therefore, we wondered whether we could adapt the scL-metric to identify patterns of gene expression at different length scales independent of cellular assignment.

To do so, we first noted that in a spatially-resolved dataset, the assignment of spots to nuclei is essentially a non-linear partitioning of genes in space — by being assigned to a nucleus, an individual expression spot loses its specific spatial location and is binned into a region defined by the cell boundary. An example of this binning is shown on the left side of Figure 5a. We reasoned that we could instead capture relationships between the purely spatial expression of genes by mapping each transcript onto an evenly spaced grid and then smoothing the resulting 2-dimensional distribution using kernel density estimation. The bandwidth of smoothing can be set in any units, but we found the average (per gastruloid) nearest-neighbor distance to be an intuitive and useful length scale to use. Figure 5a shows an example of how choosing different distances for smoothing captures different scales of patterning—any amount of smoothing that causes gene expression in adjacent cells to be averaged would increase the L-metric by increasing the degree of co-expression, and more smoothing yields a further increase in L. We chose to use density estimations with a distance value of 2 times the nearest-neighbor average distance, as this spacing seemed likely to capture meso-scale organizational patterns in gastroids while preserving finer structures of spatial variance (Figure 5a).

Spatial L-metric reveals tissue-level patterns of gene expression

a. Illustration of how the density-based L-metric (spatial L-metric) is calculated, and how the value changes as a function of how much the density estimate is smoothed. b. Clustered heatmap of the spatial L-metric for a representative gastruloid. Purple box indicates a mixed cluster of endothelial and endoderm genes. c. An image of the gastruloid used to generate the heatmap in b. Cell types are indicated by color, legend is the same as in b.

The scL-metric is calculated by comparing gene expression on a cell-by-cell basis. To adapt this measurement to spatial data, we calculated the L-metric using the density estimation at each spatial bin across the entire gastruloid (i.e., each bin is treated the way a cell is treated when the scL-metric is calculated). We denote this value as the “spatial L-metric”.

We calculated the spatial L-metric for all the consistently-detected genes in our panel across all gastruloids (202/220 genes, see Methods). When we averaged across gastruloids and clustered the values, the resulting heatmap (Figure S5.1a) had many properties similar to the scL-metric heatmap shown in Figure 5a: For the most part, cell types clustered together; however, we noted a few unique patterns. The order of clustering seemed to more closely reflect overall position in the gastruloid: the more posterior cell types (NMP, presomitic mesoderm, spinal cord) were found close together, then endoderm, differentiation front, and a mixed cluster of paraxial mesoderm, somite, and endothelial genes. This ordering suggests that when the spatially smoothed relationships (spatial L-metrics) of all genes in the panel are clustered, the resulting relationships reflect the overall spatial organization of the underlying cell types within the larger tissue structure. Individual cell types appeared to cluster more strongly when using scL-metric (compare Figure 4a with Figure S5.1a), but the order and mixing of cell type genes with the spatial L-metric reflects tissue-level organization—NMP and presomitic mesoderm genes were close together and mixed, reflecting the mixing we see of the cell types associated with these genes.

The other difference of note was that, unlike in the scL-metric heatmap, the endothelial genes were much less strongly clustered, likely reflecting the diversity of locations they were found in within individual gastruloids (Figure S1.2). We looked at their patterns in the heatmaps for individual gastruloids and noticed that there was strong co-clustering of endothelial and endoderm genes in some of the gastruloids. An example is shown in Figure 5b, which is the clustered heatmap for an individual gastruloid (cell types and morphology shown in Figure 5c). A strong block of interaction on the diagonal (“endoderm/endothelium cluster”, outlined) indicates a strong association of these genes on a length scale beyond individual cells. When we looked at the cell type arrangement of this gastruloid, there were two clear populations of endothelial cells: those that were located in the anterior and those that were more posteriorly located and strongly intermixed with the endoderm tissue. Thus, the unique clustering behavior of genes in this gastruloid is reflective of a unique, tissue-level pattern that is only found in some of our samples. When we calculate the scL-metric heatmap for the same gastruloid, although endothelial and endoderm genes cluster close together, they form two distinct diagonal clusters, compared to the single, high-L block found in the spatial-L heatmap (compare the outlined clusters in Figure 5b and Figure S5.2a).

These comparisons suggested that the spatial L-metric could capture patterns on a tunable length scale. When we directly compared the intra-cell type L-metric values calculated using both methods (Figure S5.1b), we saw that while the values for some cell types, like NMP, were quite similar, others, such as spinal cord, showed a dramatic shift towards +1 when calculated using a smoothing of 2 nearest neighbors. This shift is consistent with our observation of consistent expression of NMP genes and a more “salt-and-pepper” expression of spinal cord genes, and suggests that by calculating the spatial L-metric over many length scales, one could learn the effective length scale of interaction for specific gene pairs.

Endothelial precursors show unique organization and distinct, spatially-dependent cell states

Throughout our analyses of both spatial organization and gene expression, endothelial cells emerged as displaying particularly heterogeneous patterning and gene expression that depended on their physical location and neighboring cells, which varied between gastruloids. The relative rarity of this type has precluded in-depth analysis of their location and function in other studies [2,13,14,44], and despite strong interest in modeling hematopoiesis and the development of vasculature [45], it is still not clear to what degree gastruloids produced by standard protocols create and organize endothelial tissue. Therefore, we more specifically examined the endothelial cells identified in our gastruloids, profiling their spatial location, organization, and gene expression.

We first used density-based spatial scanning and clustering [46] to, in an unbiased fashion, detect clusters of like cells in individual gastruloids that by eye appeared to contain small clusters of endothelial precursors in the anterior. These clusters are highlighted in Figure 6a. In some cases, we observed that the cells in the cluster formed a ring-like structure (Figure 6a highlight). We applied the same spatial clustering method to all other cell types and found that, on average, the clusters of endothelial cells were the most circular (Figure 6b, Figure S6.1a). This clustering behavior is consistent with the peri-gastrulation behavior of endothelial precursors in real embryos, which have specific migratory activity [47] and form aggregated clusters that go on to form endothelial and hematopoietic cells [48]. We detected endothelial clusters in all of the gastruloids (Figure S6.1c), and found that they formed more clusters than all other cell types except paraxial mesoderm (Figure S6.1b). Importantly, the relative location of these clusters along the AP axis varied substantially between gastruloids (Figure 6c, Figure S6.2a), highlighting the necessity of studying individual gastruloids to observe the unique behavior of this rare cell type.

Endothelial precursors show unique organization and distinct, spatially-dependent cell states

a. Example of clusters of endothelial cells. Representative gastruloid shown. Highlight shows cluster morphology, color is cell type (endothelial in pink, paraxial mesoderm in mint green, somite in orange, differentiation front in light blue, and untyped cells in gray). Number of clusters ranges from 23 (cardiac mesoderm) to 185 (paraxial mesoderm). b. Circularity of clusters of different cell types. See Methods for a description of how circularity was calculated. c. Location and extent of endothelial clusters projected onto the AP axis and normalized by total length. 10 representative gastruloids are shown; the average of all gastruloids (n=26, total clusters = 155) is shown as a smoothed kernel density estimate at the top of the plot. d. Representative gastruloid showing two distinct morphologies of endothelial cells: somite-associated (anterior) and endoderm-associated (posterior). e. Genes differentially expressed in somite-associated or endoderm-associated endothelial cells. Bar color represents the cell type associated with the gene (if any); genes that were not previously known to be associated with a cell type in gastruloids are shown in gray. Bars are ordered by significance (adjusted p value) from greatest (Nepn p=5.95e-23) to least (Igfbp5 p=1.10e-5). f. Spatial distribution of expression for example genes shown in e). Each plot shows one gene, each dot is a single transcript, and cells typed as endothelial are outlined in black. The top row shows genes enriched in somite-associated endothelium, the bottom row shows genes enriched in endoderm-associated endothelium.

We also observed that in 5 out of our 26 samples, there was a large central patch of endoderm cells intermixed with endothelial precursors; these samples also had unique spatial L-metric clustering of endothelial and endoderm genes (Figure 5b). An example of one such gastruloid is shown in Figure 6d. Migration to and association with the endoderm is also a hallmark of endothelial development [47,48], and we were curious whether there were differences between these cells and the cells we observed forming anterior, somite-associated clusters. The endoderm-associated endothelial cells were generally in larger clusters, and when we computed the cell type exposure index for just this gastruloid, we found that, consistent with our visual observations, in this particular sample, endothelial and endoderm cells were much more frequently found next to one another than on average (Figure S6.1d). To determine whether these spatial and organizational differences reflected gene expression differences, we divided the gastruloid normal to the anterior-posterior axis to separate the endothelial cells into endoderm-associated and somite-associated and looked for differentially expressed genes between the two groups. The significantly differentially expressed genes we found are shown in Figure 6e.

We found that some endoderm genes were more abundant in endoderm-associated endothelial cells and that some somite genes were more abundant in somite-associated endothelial cells, likely reflecting the mis-association of genes from neighboring cells, which is common in nuclear-based spot assignment in highly packed tissues like gastruloids. However, we also found two endothelial genes that were differentially expressed in endothelial cells from the two different regions: Pecam1 and Cdh5, both of which are associated with angiogenesis, were more enriched in endoderm-associated endothelial cells, and Cxcr4, which is thought to aid haematopoiesis, angiogenesis, and cardiac cell organization by responding to homing signals. Pecam1 also clustered uniquely in our L-metric analysis (Figure 4c), suggesting this differential expression is conserved across gastruloids. Spatial expression of these genes is shown in the bottom row of Figure 6f. Although most endothelial cells are thought to be of mesodermal origin, some evidence suggests that, in the organogenesis of specific tissues like the liver, the endoderm can give rise to endothelial cells [49]. Our data demonstrate a molecularly driven organization distinct from the clustering we observed in the anterior and suggest that multiple mechanisms of endothelial specification could be modeled in gastruloids, even simultaneously within the same structure, although further characterization is needed to determine exactly what processes these unique endodermal/endothelial structures model.

The genes enriched in the anterior have less obvious functional distinctions; for example, it isn’t immediately clear how a gene like Apoe, which has been annotated to be associated with primordial germ cells, specifically functions in anterior endothelial precursors (Figure 6f, top row). Gadd45g, a stress-associated protein, is also differentially expressed, possibly reflecting the fact that cells in the anterior are exposed to more rapid growth, compression, and cell death.

Although this cell type has been consistently observed in single-cell measurements of gastruloids, its relative rarity has precluded in-depth analysis of subtypes or inference of spatial location. Our results strongly suggest that endothelial precursor formation, migration, and organization may all be modeled in 3D gastruloids; recent advances in 2D gastruloids have allowed modeling of cardiac and hepatic vascularization [45], and our data suggest that 3D gastruloids may similarly be adapted to model more specific aspects of hematopoiesis and vascularization. Early specification from a pool of mesodermal precursors is a hallmark of the endothelial lineage [47]; given the consistency with which we observe endothelial precursors, we speculate that this behavior is recapitulated in gastruloids, but further epigenetic measurements are required to validate this hypothesis.

Discussion

One of the central goals of development is to map how the identity and arrangement of cells are determined and coordinated. Although precise scaling of gene expression in gastruloids had been previously reported [10], the relationship between this scaling and the proportions of specific cell types or more complex tissue organization was not known. Towards that end, we have performed detailed measurements and analyses of the expression of hundreds of genes in tens of individual gastruloids, revealing how cell types are organized within and across gastruloids. Our results highlight that many aspects of embryonic development, such as the arrangement and differentiation trajectories of posterior cell types, are faithfully recreated on a per-gastruloid basis. Others, however, can vary, and conditions may need to be tuned to increase reproducibility if a specific aspect of their development needs to be modeled. For example, we reproduced observed variation in endodermal structures [12] and reported both the distinct organization of endothelial cells and variation in that organization between gastruloids. Although we do not fully reproduce the level of per-gastruloid cell type proportion variation reported in a previous single-cell sequencing dataset [13], we do find evidence of a tradeoff between neural and mesoderm tissues; differences in the number of cells analyzed per gastruloid could potentially explain this discrepancy. Moreover, we observe many of the same cell types as in this study, and show that their organization is quite consistent.

We introduce the L-metric, which quantifies mutually exclusive gene expression, and highlight two ways of using it that can be employed on different types of data. The cell-based L-metric, scL-metric, can be applied to any technique that generates a cell by gene table. Hierarchical clustering analysis of the resulting vectors can reveal how similar each gene is to others in terms of how it is co-expressed with all genes, and may reveal spatial information if the sample was collected from a spatially-organized tissue. We were able to verify the spatial information because our data were spatial, but the scL-metric is calculated only using the cell-by-gene table, a standard output for any single-cell analysis technique.

The spatial L-metric, which operates spatial bin-wise rather than cell-wise, can be used on spatial data even when cell assignment or nuclear segmentation isn’t possible — in principle, it can reveal which genes are patterned similarly at many different length scales, with the fundamental resolution limit set by the measurement technique. It was able to identify co-expression, even amongst lowly-expressed genes, such as the spinal cord markers, that had a more ‘salt-and-pepper’ expression pattern and were only strong markers when considered in aggregate. It is possible that some genes with bursty gene expression may appear not to express together, but spatial smoothing may allow for the identification of co-expression of these genes, which may be useful for cell type identification.

Although the identification of cell types has utility for high-level analyses of organization, our in-depth dissection using the L-metric revealed that, particularly for cells in the process of differentiating, a gene-centric view may yield a more holistic picture of tissue organization. It is particularly exciting to imagine these analyses performed in a time-resolved fashion, as dynamic changes in gene expression across a tissue in response to physical rearrangement and broad signaling patterns could reveal mechanisms or spatial aspects of differentiation not visible when cells are merely binned into predetermined categories.

Methods

Cell culture

All gastruloids were generated from E14TG2a cells (a generous gift from the Toettcher lab). Cells were cultured in 2i+LIF media at 37 °C, 5% CO2 and grown on plates coated in 0.1% gelatin (Millipore/Sigma ES006B). 2i+LIF media consisted of GMEM (Gibco 11710-035), 10% stem-cell qualified FBS (R&D Systems S10250), 1x GlutaMAX (Gibco 35050–061), 1x MEM non-essential amino acids (Gibco 11140050), 1 mM sodium pyruvate (Gibco 11360–070), 100 μM 2-mercaptoethanol (Gibco 471 21985–023), and 100 units/mL penicillin/streptomycin (Gibco 15140–122), 1000 units/mL LIF (Millipore Sigma ESG1107), 2 μM PD0325901 (Tocris 4192), and 3 μM CHIR99021 (Tocris 4423). Cells were passaged every 2 days: cells were trypsinized (Life Technologies 12563011), counted using an automated cell counter (Invitrogen Countess II), and 5x10-5 — 8x10-5 cells were transferred to a 25cm2 dish; cells were passaged at least once after thawing before gastruloid aggregation.

Gastruloid aggregation

The aggregation protocol was as described in [2] (protocol here). Briefly, cells were trypsinized, rinsed twice with 1X PBS to remove media, and then resuspended in N2B27 medium. N2B27 consisted of equal parts neurobasal medium (Gibco 21103049) and DMEM/F12 (Gibco 11320033) supplemented with 1X N2 (Gibco 17502048) and B27 (Gibco 17504-044), 100 units/mL penicillin/streptomycin (Gibco 15140–122), and 100 μM 2-mercaptoethanol. Cells were resuspended at a concentration of 2500-7500 cells/mL; 40 μL of this mixture were placed in each well of round-bottom ultra non-adherent U-bottom 96-well plates (Costar 7007), yielding 100-300 cells/well. The first two seqFISH runs used 300-cell gastruloids (n=8 total) and the last used 100-cell gastruloids (n=18). Cells were allowed to grow at 37 °C, 5% CO2 for 48 hours and then supplemented with 150 μL fresh N2B27+3 μM CHIR99021 (Chiron, Wnt agonist). After 24 hours, 150 μL of media was removed and replaced with N2B27 without Chiron; this media change was repeated daily until the gastruloids reached maturity at 120 hours. All wells were scored for elongation at 120 hours just before fixation. The gastruloids selected for fixation were elongated and uniaxial; the experiments that generated them had an 80-94% success rate (see Figure S1.2d for full distribution).

Sample fixation

After gastruloids had grown for 120 hours, they were harvested by pooling together all successful (i.e., elongated and uniaxial) wells in a plate, washed with 1X PBS, and then fixed for 10 minutes in 4% PFA in PBS. After fixation, gastruloids were washed with 1X PBS and stored in methanol for up to a month before preparation for seqFISH. 3 separate seqFISH runs were performed on 6 different aggregations (one experiment pooled 4 plates of gastruloids), producing a total of 26 individual samples.

Embedding and sample preparation for seqFISH

∼30 gastruloids were placed in a Lo-Bind tube (Eppendorf 0030108523). After removing as much methanol as possible, the gastruloids were washed with 2XSSC, then incubated in 2XSSC + 1 μg/mL DAPI overnight at 4°C. Stained gastruloids were permeabilized with 1X PBS + 0.2% Tween20, arrayed on a gelatin-coated slide (Spatial Genomics 00200001) with an eyelash tool (Electron Microscopy Sciences 11852), covered in liquid acrylamide gel, and allowed to polymerize for 2 hours. Polymerized samples were washed with 8% SDS, then treated with 8x10-3 units/mL Proteinase K (NEB P8107S). After this wash, the primary probes were hybridized for 16-40 hours, after which the samples were washed and loaded onto a GenePS machine (Spatial Genomics).

Detailed protocol available here.

seqFISH and raw data handling

Sequential rounds of secondary probe addition and imaging were performed in an automated fashion by the instrument [16]. The GenePS software that controlled the instrument was v1.6.0. The instrument was run with a 1 or 2 μm z-offset on 5-18 individual regions of interest, highlighting each gastruloid. The genes chosen for the seqFISH panel were based on differential expression in single-cell RNA sequencing experiments in references [2,13] or by consensus in the literature. 200 of the genes were detected via a sequential barcoding scheme (see [16] for details), and 20 were serially detected as in traditional smRNA-FISH [50]. The serial genes tended to vary in quality between gastruloids; genes with high background or in which expression was too high for the spot detection software to confidently assign spots were excluded. The list of excluded genes for each gastruloid is included with the raw data.

Gastruloid annotation and spot assignment

Detected spots were decoded and exported without nuclear assignment by SG Navigator software version 0.9.0. At this stage, gastruloids without even and consistent expression of the housekeeping gene Eef2, which indicated focus or technical issues or sample variability, were discarded. 5 out of 6 gastruloids were discarded from the first run, and 3 out of 11 in the second run. None of the samples from the third run were discarded. Raw TIFF files of the DAPI images were annotated using the NimbusImage platform, which is available here: https://github.com/arjunrajlaboratory/NimbusImage/ and hosted here: https://www.nimbusimage.com/. Expression of the gene T was used to locate the posterior end and corroborated by consistent differences in nuclear morphology between the posterior and anterior. An AP axis was inferred by drawing a line between that pole and the extreme opposite end of the gastruloid along the center of the gastruloid parallel to the edges and connecting with the furthest extended point to define the anterior pole. Segmentation was performed using cellposeSAM [51] on the high-resolution .tiff files of the DAPI-stained nuclei from the 3rd round of hybridization. Once the nuclear masks had been created, spots were assigned to nuclei using a custom software package called SGAnalysis (available here). Nuclei were dilated by 8 pixels (0.86 μm), and all spots that fell within this dilated nuclear area were assigned to that nucleus. In cases where the dilated regions overlapped, spots were assigned to the nucleus for which the edge of the dilated area was furthest away. On average, we were able to resolve 131,112 nuclei for each gastruloid, and of those nuclei, 79,607 had enough mRNA counts to be typed. Nuclei without enough mRNA were generally small, uniformly bright (in contrast to the usual mottled appearance of mouse nuclei), and lacked expression of the highly-expressed housekeeping gene Eef2, indicating the corresponding cell was likely dead or fragmented. We primarily found these nuclei in the anterior tissues and classified them as “none”-type, excluding them from our analyses unless noted. The average reads/nucleus per gastruloid ranged from 47-80; for typed nuclei the value was 105-265.

Cell type scoring

To assign cell type, we used an panel of marker genes for 9 consensus cell types thought to exist in gastruloids at this stage of development: Neuromesodermal precursor (NMP), presomitic mesoderm (PSM), spinal cord precursor (spinal cord), endoderm, differentiation front, cardiac mesoderm, paraxial mesoderm, somitic, and endothelial precursor (endothelium). Genes were chosen based on differential expression in references [2,13], or consensus in the literature. Each nucleus received a score for each cell type using the score_genes function in scanpy [52], and the highest-scoring cell type category was assigned to that cell. If none of the scores met a universally applied cutoff (0.1), then the cell was assigned ‘none’.

Cell type entropy score

The degree of uncertainty in cell typing was quantified by calculating the entropy of the cell type score distribution. The scores can be negative or zero, so they were first converted to probabilities using a softmax function with temperature scaling: P(i)=exp⁡(s′i)/∑exp⁡(s′j)where s′i=(si−max⁡(s))/τ A temperature of 0.1 was used. Once the scores had been converted, the Shannon entropy of the distribution was calculated: H(P)=−∑ipilog2⁡(pi)

Mixing score calculation

The mixing score for a gastruloid was calculated by finding the nearest neighbor of each cell and determining how many cells had a nearest neighbor of the same type. This value was normalized by dividing by the expected number of same-type nearest neighbors based only on the frequency of each type in that gastruloid. Values range from 0 (perfectly mixed) to 1 (perfectly segregated).

Exposure index

The exposure of each cell type was calculated by finding the k nearest neighbors (“target”) of every cell (“source”) in every gastruloid (k=5 was used), counting the frequency of each cell type among these neighbors, and then normalizing based on the overall frequency of occurrence of the neighbor (“target”) cell type. This generated a pair-wise exposure score of each cell type to self (symmetric) and all other cell types (asymmetric) that ranged from 0 (never found next to one another) to 1 (always found next to one another).

Triplet identification and enrichment

Spatial enrichment of cell type triplets was assessed using a computational framework based on Delaunay triangulation. Each gastruloid was processed independently to account for potential inter-sample variability in cell density and spatial organization. Spatial neighbors were identified using Delaunay triangulation, which partitions the coordinate space into triangular tessellations connecting nearest neighbors. A distance cutoff of a maximum of 200 pixels was used; neighbors with greater than this distance were not considered. All groups of 3 nearest neighbors were used to evaluate triplet frequency. Statistical significance was evaluated through a permutation testing framework. For each sample, the observed triplet frequencies were compared against null distributions generated through random reassignment of cell type labels while preserving spatial coordinates. The number of permutations was set to 1,000 per sample. The data for all gastruloids were compiled together such that the significance of each triplet in each sample, and how many samples each triplet was found in, could be assessed. Only significantly enriched or depleted triplets were analyzed; significant triplets found in at least half of the gastruloids are shown in Figure 2b, and all are shown in Figure S2b.

Kernel density estimation for generating smoothed spatial profiles of gene expression

We performed kernel density estimation (KDE) to generate smoothed spatial profiles of gene expression patterns. Using the nonparametric.KDEmultivariate method from the statsmodels package in Python, a given gene’s KDE of gene expression in a given gastruloid was calculated using a two-dimensional Gaussian kernel for each RNA spot fitted to a bandwidth equal to the average nearest neighbor distance between cell centroids in that gastruloid multiplied by 2 (Table S1) in both dimensions. To produce an expression profile for a gene from its KDE on a given gastruloid, we first partitioned the gastruloid’s annotated hull to produce a grid containing square elements, each with side length equal to half of the average cell diameter, and then evaluated the KDE on the center of each grid element. Visualizations of multiple genes’ combined expression in a given gastruloid were obtained by summing individual genes’ normalized KDE values for each grid element.

Calculating the single-cell L-metric

The scL-metric for a given gene pair was calculated per-gastruloid using a cell-by-gene matrix filtered to exclude cells with less than 2 transcripts for both genes and sorted to place cells in descending order according to the first gene’s expression values. The scL-metric is asymmetric with respect to pair order in that the first gene provides the reference expression distribution to which observed and simulated expression distributions of the second gene are compared. On the ordered and filtered cell-by-gene matrix, the following three simulated expression distributions were created for the second gene: perfect coexpression, in which the second gene’s expression values were sorted in descending order, like the first gene; perfect anti-coexpression, in which the second gene’s expression values were sorted in ascending order; and, uniform coexpression, in which all cells were assigned the mean expression value of the second gene. Cumulative sums were calculated for the observed expression distributions of the first gene and second gene and the simulated expression distributions of the second gene. The cumulative sums of each expression distribution associated with the second gene were plotted against the cumulative sums of the first gene, and each of the resulting curves’ areas were obtained using trapezoidal integration and normalized by the product of the first gene’s total expression and the second gene’s total expression, producing Aobserved, Apositive, Anegative, and Auniform (Figure S3.2a,b). The scL-metric value for this gene pair, specific to its ordering, was then determined by comparing Aobserved to each of the areas under the simulated distributions’ curves: IfAobserved>Auniform,then scL-metric=(Aobserved−Auniform)/(Apositive−Auniform), resulting in 0<scL-metric<1.IfAobserved=Auniform,then scL-metric=0.IfAobserved<Auniform,then scL-metric=−(Aobserved−Auniform)/(Anegative−Auniform), resulting in −1<scL-metric<0. We computed the scL-metric across all gene pairs for each gastruloid in our dataset from 05/07/2025 (n=18 gastruloids). We did not integrate this dataset with our other two datasets (n=8 gastruloids), because relatively poorer detection quality in those datasets limited our ability to estimate the distributions necessary to calculate the scL-metric. Because some of the serial genes in the seqFISH panel were inconsistently detected between samples (see seqFISH section of Methods), the number of gene pairs varied slightly between samples. When an scL-metric matrix for an individual gastruloid is shown, the number of genes for that gastruloid is indicated. We averaged together the scL-metric matrices for the 18 individual gastruloids. For this analysis, a common set of 202 genes with good detection in all gastruloids was used.

All scL-metric analyses, with the exception of hierarchical clustering analysis, use a symmetrized form of the scL-metric obtained by averaging the two directional scL-metric values for a given gene pair. This symmetrization was motivated by the aim to summarize the coexpression relationship between two genes with a single value and the lack of significant differences between directional L-metric values.

Cell type analysis via unsupervised clustering

We used scanpy [52] to process the dataset from 05/07/2025. All nuclei were pooled together, nuclei with less than 40 total gene counts were filtered out, and then PCA (50 PCs) and UMAP embedding were performed using the standard pipeline. We performed Leiden clustering using scanpy.tl.leiden, and used visual inspection to choose a resolution at which to cluster. A resolution of 0.8 produced fairly well-separated clusters and roughly corresponded to the expected number of cell types. Clustering results and spatial projection are shown in Figure S4.4a.

Filtering marker genes

We re-scored nuclei using a refined marker gene panel. We retained only genes that met the following criteria: more than 10% of all scL-metric values < -0.8 (73% of all genes) and more than 12% of all scL-metric values > -0.3 (90% of all genes). 70% of the genes in our panel met both. Scores with this filtered gene panel are shown in Figure S4.3b.

Assessing marker gene panels and cluster-associated genes

To visually represent the relatedness of the genes either differentially expressed in clusters identified with unsupervised clustering, or the groups of genes associated with cell types in a marker gene panel, we calculated the average L-metric score of all possible combinations where the first gene in the L-metric score belonged to the first type or cluster, and the second gene belonged to the second group or cluster. In the case of the marker gene panel, we first filtered genes on expression level, considering only genes that had > 2.3 counts / nucleus in nuclei where they were expressed; this filtering retained 84% of the genes in the panel.

Using cNMF to find gene clusters

All nuclei were pooled together, and cNMF [39] was run on all nuclei and all non cycle-genes. The default parameters were used and 5-10 components were calculated. 7 components produced the highest stability, and the top 24 most positively differential (by ordered z-score) genes were used for visualization.

Calculating the spatial L-metric

The spatial L-metric was computed analogously to the scL-metric except that rather than rank ordering the expression of genes in individual cells, we instead fitted a KDE over the detected spots for a given gene and evaluated it on a square grid of ’spatial bins’ that spanned the entire gastruloid (as described above), then rank-ordered these spatial bins by the resulting probability densities of expression. We treated these bins the same way we had treated cells in the previous calculation, so that the resulting L value reflected mutually exclusive expression in space rather than per-nucleus. The length scale of the interactions considered was determined by the chosen size of the spatial bin. Throughout, we report values for bins at 2 times the nearest-neighbor distance (which was calculated on a per-gastruloid basis). We did not filter low-expressing bins the way we had for cells for 2 reasons: 1) we calculated the KDE on all spots detected, not just those that fell within a nuclear area, so there were fewer areas of low expression, and 2) smoothing resulted in a non-zero density throughout the imaging area. Just as for the sc-L-metric, we symmetrized the spatial L-metric by averaging the 2 values calculated for each gene pair, except when displaying the values as a heatmap.

Hierarchical clustering of L-metric matrices

We performed hierarchical clustering on L-metric matrices to group genes based on their pairwise coexpression relationships. Hierarchical clustering of a given L-metric matrix was done by applying the cluster.hierarchy.linkage method from the SciPy package in Python with parameters method and distance set to “ward” and “euclidean,” respectively, on the rows of the matrix, which represented interactions between a given gene (as the first gene in the pair) and all other genes. We performed this hierarchical clustering analysis on the following gene sets: 1) A subset of only NMP, presomitic mesoderm, and spinal cord genes, evaluated for 1 gastruloid (n=36 genes, Figure 3g); 2) a subset of all genes well-detected in all gastruloids, but not including cell cycle genes (n=166 genes, evaluated on the average value across all gastruloids in Figure 4); 3) a common set of 202 genes well-detected in all gastruloids (evaluated on the average L-metric values across all gastruloids in Figure S4.1a and Figure S5.1); and, 4) a subset of genes in one an example gastruloid excluding poorly-detected genes and cell cycle genes (n=171 genes, Figure 5b and Figure S5.2a).

Cell cluster identification

Cell clusters were identified using DBSCAN [46] implemented in sklearn.cluster. A maximum distance of 300 pixels was used as a cutoff for neighbor identification, and a minimum of 8 cells was required to seed a cluster.

Cluster circularity

Cluster circularity was calculated by taking the convex hull of the identified cluster, then comparing the shape of that convex hull to a circle. The area (A) and perimeter (P) of the convex hull were measured, and then the following formula was used to calculate circularity: C=4πA/P2 If C=1, the shape was a perfect circle. Values less than 1 indicated deviations from circularity.

eLife Assessment

This work presents important findings on quantifying gene coexpression from spatial omics. These quantification methods have been applied to gastruloid to describe how genes are spatialised. The description of the quantifying tools might be incomplete, which also weakens the biological message. Clearer formalization and justification of quantification will improve the study.

Reviewer #1 (Public review):

Summary:

The authors performed seqFISH in 26 gastruloids and performed a variety of computational analyses on these novel spatial data sets. Whilst the data is valuable and the computational concepts useful (exposure index, L-metric, ... ), the article falls short on novelty and is written using a very clunky language, often with contradictory conclusions.

Major issues:

(1) The authors did well in explaining and detailing the provenance of data and the individual experiments performed. However, their 26 gastruloid data still constitute a very limited sampling from their total organoids: one experiment pooled 4 plates at an 80-94% success rate; 6 different aggregation experiments were done, making a total of 1843 gastruloids, sampled 26 (~1-2%). A simple IF stain of 2-3 markers in a bigger sample could have given a more accurate picture of specific domains of interest and their proximity. Regardless, more information should be given about the existing samples: variation across experimental batches, differences between 300-cell vs 100-cell gastruloids that were used.

(2) Language in the manuscript should be revised. Overall the manuscript is very long, descriptive and written "impressions and beliefs" are often not adequately justified and indeed can be contradictory, e.g. in Section 1: the title states "cell types' locations ...are consistent", a few sentences down we find "there was substantial variation" and "within range of what would be considered a 'morphologically normal' gastruloid". "quite consistent", "compelling patterning", "we don't believe"... these types of expressions are best avoided and replaced with data or used and bolstered with quantitative numbers such as percentages when a given cutoff is used. Another example: "location of each cell type relative to gastruloid morphology was quite consistent the posterior region ... mainly consisted in NMPs." Given T expression in the posterior, this result phrased as such appears quite inflated, in fact, looking at cell types in Figures S1, 2a/b/c, this reviewer would state they are all but consistent and indeed it takes sophisticated analyses to find a pattern (of sorts) beyond the coarse domains expected!

(3) Figure 6 is one of the most valuable parts of the work, as the authors use the battery of analyses developed to investigate the variable and not-so-robust endothelial clusters in gastruloids. However, this investigation is still very preliminary, and it should be further linked with known biology. It is still unclear what the unique organization of this cell type is (circularity isn't convincing) and whether any signalling cues of adjacent cells could explain it. Is there any evidence that more mature endodermal cell types are generated (like the suggested "liver") to give rise to endothelial cells? It would certainly be interesting to perform IF for this cell type together with mesodermal and endodermal markers to validate seqFISH predictions on a bigger sample.

(4) Figures 1c and 6b need statistical significance assessments.

(5) The article should include an analysis of Hox colinearity expression in these gastruloids as a validation of the system.

Reviewer #2 (Public review):

Summary:

This manuscript presents an ambitious and technically challenging spatial-transcriptomic atlas of 26 gastruloids using seqFISH. The authors introduce quantitative metrics (mixing score, exposure index, L-metric / scL-metric, spatial L-metric, triplets) to characterize spatial organization at multiple scales. The dataset is valuable, and several analyses are original, particularly the rank-based L-metric family for mutual exclusivity.

Strengths:

The authors generate one of the most detailed spatial transcriptomic datasets of gastruloids to date. They propose creative computational metrics (L-metric/scL-metric) to quantify mutual exclusivity of gene expression without predefined thresholds, and they explore organizational principles from single-cell topology to cluster-level structure. Many observations align well with known gastruloid biology, such as posterior robustness and anterior variability. The writing is generally clear, and the figures are rich.

Weaknesses:

Several central claims rely on metrics whose computation and justification are insufficiently explained, making it difficult to assess how robust or interpretable the results are. Many choices in the analysis appear arbitrary or are insufficiently motivated (normalization schemes, choice of parameters such as the number of neighbors, the distance cutoffs, hierarchical clustering setup, and so on). The interpretations of spatial consistency, gene-program inference, and endothelial heterogeneity are plausible but might be stronger than the evidence currently supports.

The manuscript would benefit from stronger benchmarking, quantification of uncertainty, and explicit controls for known artifacts in spatial transcriptomics (e.g., spillover, 2D slicing, cell type assignment entropy). The biological insights are promising, but since several depend on methodological assumptions that have not yet been demonstrated to be stable, they would benefit from clearer methodological explanation.

The work is rich and could become a reference dataset. Then, clarifying and validating the quantitative methods will considerably strengthen the impact and reliability of the conclusions.

Reviewer #3 (Public review):

Summary:

Triandafillou and colleagues report a single-cell resolved spatial atlas of gene expression of 26 gastruloids. While previous work had analyzed either single-cell gene expression or spatially coarse-grained patterns of gene expression (van den Brink et al, 2020), the authors here use multiplexed sequential RNA FISH (seqFISH) to create the first gastruloid atlas, which is simultaneously spatially and cellularly resolved. This atlas adds to a growing list of resources cataloging gastruloid development (see also Suppinger et al 2023).

To analyze this dataset, the authors also describe a novel analytical framework. Their analysis centers around the 'L-metric', which measures the degree to which pairs of genes are either coexpressed or mutually exclusive. While this metric is similar to calculating correlations in gene expressions, it has important differences (including that it can, in principle, be asymmetric; although the authors symmetrize much of their analysis). In addition to the gene-centric L-metric analysis, the authors also analyze cells in their dataset according to the cell type entropy (an information-theoretical measure of confidence in cell type assignment) and the 'exposure index' (a measure of the similarity of nearest cellular neighbors).

Using this framework, the authors focus their analysis on two major features of development. The first is the differentiation of the bipotent neuromesodermal progenitor (NMP) cells in the posterior of the gastruloid into either presomitic mesoderm (PSM) or spinal cord SC lineages. They use L-metric analysis to compare overlap in marker genes used to separate NMP, PSM, and SC fates. They highlight that L-metric analysis can recover spatial patterns of gene expression (without explicit spatial information) and discern subtle features of marker genes beyond simple binning of cell types (e.g., that Epha5 expression in anterior NMPs may predict future SC differentiation).

The second is the formation of endothelial (spatial) clusters within the gastruloid. The authors highlight two subtypes of endothelial clusters: (1) smaller clusters within the somitic anterior region, and (2) larger clusters associated with endoderm. While the authors discern some subtle differences in gene expression between these two clusters, their different spatial patterns suggest a potential physiological difference that would not be captured in traditional droplet microfluidic-based scRNAseq pipelines.

Overall, this manuscript is a sophisticated and technically sound study that will provide a valuable beachhead for future studies of developmental patterning in gastruloids and organoids.

Strengths:

The major strengths of this study are the overall technical sophistication of the data set and analysis, as well as its potential generalizability to other developmental systems (both in vitro and in vivo). The data are extensively analyzed and reasonably interpreted, and this atlas makes good use of the variability in gastruloid development to extract the statistical structure of developmental processes. The L-metric offers a parameter-free tool to analyze transcriptomic datasets that could overcome the pitfalls of other approaches.

Weaknesses:

The major limitations of this study are the depth and novelty of the developmental processes studied. The authors provide very convincing proof-of-concept that their data set can recover known features of gastruloid development, including NMP differentiation and endothelial development. However, further analysis and/or investigation would be required to discover new principles of gastruloid development and patterning.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

The authors performed seqFISH in 26 gastruloids and performed a variety of computational analyses on these novel spatial data sets. Whilst the data is valuable and the computational concepts useful (exposure index, L-metric, ...), the article falls short on novelty and is written using a very clunky language, often with contradictory conclusions.

We thank the reviewer for their comments about the value of our data and computational concepts. We agree with the reviewer’s critical comments and have endeavored to address them all. We believe the resulting manuscript is greatly clarified and improved.

Major issues:

(1) The authors did well in explaining and detailing the provenance of data and the individual experiments performed. However, their 26 gastruloid data still constitute a very limited sampling from their total organoids: one experiment pooled 4 plates at an 80-94% success rate; 6 different aggregation experiments were done, making a total of 1843 gastruloids, sampled 26 (~1-2%). A simple IF stain of 2-3 markers in a bigger sample could have given a more accurate picture of specific domains of interest and their proximity. Regardless, more information should be given about the existing samples: variation across experimental batches, differences between 300-cell vs 100-cell gastruloids that were used.

This omission was an oversight on our part and we thank the reviewer for catching it. We did the following to address this point:

(1) Added date labels to Figure S1.2d (now S1.1a) so that the proportion correct for each separate experiment is clear:

(2) We added the raw images of the gastruloids used in the study taken before fixation.

(3) We segmented these images and quantified metrics of the masks to address differences across samples in morphology. We note that the samples collected on 9/1/2024 were on average smaller than the other two experiments, but spanned the same range of elongation. Elongation was measured as 1-(the ratio of the width and the height of the segmented gastruloid area); the code can be found here:

https://github.com/arjunrajlaboratory/ImageAnalysisProject/blob/1b2f2119f77083c27f58f8c36b14 c48d97ea706c/workers/properties/blobs/blob_metrics_worker/entrypoint.py

Interestingly, the final size as measured by cross-sectional area of a brightfield image of the gastruloid did not correlate with the initial seeding number (the experiment on 4/4/2025 used 100 starting cells and the other two experiments used 300).

The literature also supports our assumption that combining experiments with different starting numbers of cells would not dramatically affect the results (Bennabi et al. 2024). In this paper they show that only 35 genes were differentially expressed between gastruloids formed from 100 cells and those formed from 300 cells (compared to 319 for those formed from 1200 cells and 336 for those formed from 50 cells, both compared to 300 cells). The same paper also demonstrates that the positioning of gene expression (as measured by IF staining) for several representative genes (Bra and Foxc1) does not significantly differ between gastruloids formed from 100 and 300 cells when normalized for overall size and AP axis length (as we’ve done in this paper as well).

We have updated the text and revised Figure S1.1 to reflect these changes.

“To measure the spatial distribution of gene expression, we prepared gastruloids using mouse E14TG2a cells and a standard protocol (see Methods). We harvested mature gastruloids after 120 hours of growth. The experiment was performed 3 times on different days, so to ensure consistency we checked that the proportion of the gastruloids that formed correctly was the same or greater than the median of all experiments (Figure S1.1a). Although there was variation in the length, width, and relative amounts of anterior and posterior tissues in the gastruloids considered, they were within the range of what would be qualitatively considered a ‘morphologically normal’ gastruloid [1,10].

To address potential batch effects due to biological differences between runs, we examined brightfield images of all the gastruloids generated for each experiment (529 total gastruloids across 6 plates on 3 different days), segmented them, and quantified morphological characteristics. When we embedded all 529 gastruloids into PCA space, there was near-complete overlap between all groups, with the exception of one plate from 9/1/2024, which was slightly higher in PC1. Figure S1.1b shows this embedding, and examples of gastruloids at the extreme ends of PCs 1 and 2. We note that the samples collected on 9/1/2024 were on average smaller than the other two experiments, but spanned the same range of elongation (Figure S1.1c). Interestingly, the final size as measured by cross-sectional area of a brightfield image of the gastruloid did not correlate with the initial seeding number (the experiment on 4/4/2025 used 100 starting cells and the other two experiments used 300). Previous studies have demonstrated that the gene expression differences between gastruloids seeded with 100 and 300 cells is extremely small [Bennabi 2025]” (see also Revised Figure S1.1a-c).

(2) Language in the manuscript should be revised. Overall the manuscript is very long, descriptive and written "impressions and beliefs" are often not adequately justified and indeed can be contradictory, e.g. in Section 1: the title states "cell types' locations ...are consistent", a few sentences down we find "there was substantial variation" and "within range of what would be considered a 'morphologically normal' gastruloid". "quite consistent", "compelling patterning", "we don't believe"... these types of expressions are best avoided and replaced with data or used and bolstered with quantitative numbers such as percentages when a given cutoff is used. Another example: "location of each cell type relative to gastruloid morphology was quite consistent the posterior region ... mainly consisted in NMPs." Given T expression in the posterior, this result phrased as such appears quite inflated, in fact, looking at cell types in Figures S1, 2a/b/c, this reviewer would state they are all but consistent and indeed it takes sophisticated analyses to find a pattern (of sorts) beyond the coarse domains expected!

We thank the reviewer for their careful reading of our paper and appreciate that the work would be strengthened by increasing the degree to which quantitative measures are used to justify the statement we make. We have made the following changes to the manuscript to address this criticism:

(1) We more clearly delineate where we are making qualitative descriptions and have removed summary language (like ‘consistent’, ‘normal’, ‘variable’ etc.) from these sections. For example, the section the reviewer refers to originally read:

“Once we had the cell type identity and spatial location of each cell in all the gastruloids, we characterized the organization of each by mapping where each cell type was found relative to other types and overall morphology. The approximate location of each cell type relative to gastruloid morphology was quite consistent: the posterior region, although highly variable in size (Figure S1.2a,b,c), mainly consisted of neuromesodermal precursors (NMP, turquoise), a bipotent cell type that contributes to both neural and mesodermal tissues [17–19]....”

And now reads:

“Once we had the cell type identity and spatial location of each cell in all the gastruloids, we first qualitatively examined where each cell type was found relative to other types and overall morphology. The posterior region, although variable in size (Figure S1.3a,b,c), mainly consisted of neuromesodermal precursors (NMP, turquoise), a bipotent cell type that contributes to both neural and mesodermal tissues [17–19]...”

(2) We follow this qualitative description with a quantitative analysis of cell type proportion where we clearly state which variable aspects are statistically significant:

“We sought to quantify variability in cell type composition between the 26 morphologically normal gastruloids. Previous single-cell datasets relied on pooling multiple gastruloids, thus obscuring the degree to which the overall cell type distribution was reflected in each individual gastruloid. However, recent single-cell measurements of individual gastruloids have suggested substantial gastruloid-to-gastruloid variation in cell type proportions [13]. Figure 1c shows distributions of cell type proportions across samples, and Figure 1d shows the coefficient of variation of these proportions. Individual gastruloid cell type distributions, including the proportion of cells that had insufficient reads to be confidently assigned a type, are shown in Figures S1.4b and c. We found that cardiac mesoderm, endoderm, and spinal cord cells had the greatest coefficient of variation in proportion between gastruloids (Figure 1d). To calculate statistical significance, we first performed a centered log-ratio (CLR) transform on the proportions, then looked for covariation between cell types across gastruloids. We found there was a statistically significant inverse correlation between the proportion of endoderm and NMP, presomitic mesoderm, and differentiation front (Figure S1.4d).”

(3) We added a summary paragraph at the conclusion of the results from the first two figures which clearly states which aspects of gastruloid organization we find to be variable and which are consistent, with statistical testing:

“Variation in cell type abundance and organization is structured and concentrated in specific cell types

We have demonstrated that some aspects of gastruloid composition and spatial organization are consistent across gastruloids, while others are more variable. Consistent features include proportions for NMP, presomitic mesoderm, somite, and paraxial mesoderm, whose coefficients of variation were lower than other cell types (Figure 1d). Organizationally, all cell types across gastruloids are more physically clustered than random (Figure 2a), and the order in which cell types are found along the AP axis has statistically significant high agreement between gastruloids as measured by Kendall’s W (Figure S1.5c). At the local neighbourhood scale, we found that most cell type interactions were conserved across gastruloids (Figure S2.1c). At the local scale, across individual gastruloids, we found many motifs of three cells that were statistically enriched over random, suggesting a conserved local order (Figure 2c). While the normalized distance along the AP-axis of all cell types significantly varied compared to a bootstrapped null (Figure S1.5a), the effect size was small, and decreased in almost all cases when normalized to gene expression (of T) in addition to morphology (Figure S1.5b).

However, there were also variable features. The proportion of cardiac mesoderm, endoderm, and spinal cord had the highest coefficient of variation between gastruloids (Figure 1d). Because proportions must sum to one, a change in the proportion of one cell type is necessarily linked to changes in others; we performed centred log transformation and looked for statistically significant covariation. Of all possible pairings, the following proportions had a significantly negative correlation across samples: endoderm/differentiation front, NMP/endoderm, presomitic mesoderm/endoderm, none/endothelial, and spinal cord/endothelium. This result shows that the proportions of these cell types predictably co-vary between samples, potentially suggesting some kind of biological trade-off in cell type specification or organization (Figure S1.4d).

Across gastruloids, intra-cell type interactions (degree of clustering) of spinal cord, endoderm, and differentiation front vary (Figure S2.1b). This variation suggests that these cell types may be patterned differently between gastruloids. For example, the local motif of 3 endoderm cells found next to one another was statistically enriched within some but not all individual gastruloids, and by definition is completely absent from gastruloids lacking endoderm (Figure 2c). We interpret this contrast to mean that when endoderm is found in a gastruloid, it is consistently patterned at a local level, but may vary more at a global level. This interpretation is concordant with the findings from [Farag 2024], which demonstrates several distinct classes of endoderm organization in gastruloids.

To summarize, while changes in the amount of individual cell types can vary, these changes are in most cases explained by variations in morphology and molecular characteristics (such as anterior: posterior ratio and the expression of morphogens like T). For patterning, we found that, in most cases, global patterns were conserved, but there were small variations in local patterning that may lead to variable meso-scale organization of specific cell types, particularly those found in the middle of the anterior-posterior axis.”

(3) Figure 6 is one of the most valuable parts of the work, as the authors use the battery of analyses developed to investigate the variable and not-so-robust endothelial clusters in gastruloids. However, this investigation is still very preliminary, and it should be further linked with known biology. It is still unclear what the unique organization of this cell type is (circularity isn't convincing) and whether any signalling cues of adjacent cells could explain it. Is there any evidence that more mature endodermal cell types are generated (like the suggested "liver") to give rise to endothelial cells? It would certainly be interesting to perform IF for this cell type together with mesodermal and endodermal markers to validate seqFISH predictions on a bigger sample.

We appreciate the reviewer pointing out that the comparisons between different endothelial cell types was interesting, and agree that the clustering methods were insufficiently justified and that a more explicit consideration of the signaling context of the gastruloid could strengthen our findings.

We have re-evaluated how we calculate differentially expressed genes. We restricted our analysis to only consider genes that are expressed at > 2 counts/cell in at least 50% of the subsets considered. The results are in shown in the revised Figure 6.

We find that, as the reviewer suggested, some signaling genes are significantly differentially expressed. Specifically, Notch1 is more expressed in endoderm-associated endothelial cells, and this could reflect an increase in notch signaling in the posterior of the gastruloid. Tek, on the other hand, is more expressed in the somite-associated endothelial cells, and Tek has been annotated to be involved in retinoic acid signaling. These findings align with the reviewer’s observation that signaling from adjacent cells could explain or relate to differentially expressed genes.

We also did a more thorough review of the literature, and found several papers that reported unique subsets of endothelial precursors, albeit in related systems. In [Rossi 2022] and [Rossi 2021] the authors find a population of endoderm-associated endothelial cells in gastruloids grown with a different protocol that involves Matrigel embedding, treatment with factors that promote blood development, and growth for 168 hours. In [Veenlveit 2020] the authors find a unique somite-associated population of endothelial cells in Trunk-Like Structures, which are similar to gastruloids but model later in development and have more physical organization with discrete somites. To address the reviewer’s request that we further link with known biology we have added the following to the text:

“We observed that in 5 out of the 26 gastruloids, there was a large central patch of endoderm cells intermixed with endothelial precursors; these samples also had unique spatial L-score clustering of endothelial and endoderm genes (Figure 5b). An example of one such gastruloid is shown in Figure 6a. Migration to and association with the endoderm is also a hallmark of endothelial development [47,48], and we were curious whether there were differences between these cells and the cells we observed forming anterior, somite-associated clusters. When we computed the cell type exposure index for just this gastruloid, we found that, consistent with our visual observations, in this particular sample, endothelial and endoderm cells were much more frequently found next to one another than on average (Figure 6b). To determine whether these spatial and organizational differences reflected gene expression differences, we divided the gastruloid normal to the anterior-posterior axis to separate the endothelial cells into endoderm-associated and somite-associated and looked for differentially expressed genes between the two groups in this gastruloid. To ensure we were focused on genes that truly varied in expression in endothelial cells and were not merely a reflection of spillover from surrounding cells, we pre-filtered genes on expression, so only genes that were present in at least 50% of the cells in either group at a greater than 2 count per cell level were considered. The significantly differentially expressed genes after filtering are shown in Figure 6d. As an additional check on the degree to which transcript mis-assignment affected our analysis of gene expression in these cells in particular, we varied the nuclear dilation in this gastruloid specifically, and calculated cell type score entropy as a function of nuclear dilation (Figure S6.1a). Because cell type score entropy of a cell reflects the degree to which that cell specificity expresses genes associated with a single cell type, our expectation was that if spillover between endoderm and endothelial cells was a significant issue, then decreasing the nuclear dilation should greatly decrease the entropy scores for both groups. Although we saw a slight increase in the spread of the distribution as nuclear dilation increased, the median cell type entropy stayed extremely low for both groups (Figure S6.1a). From this analysis we conclude that the genes we identify as differentially expressed are not due to spillover from surrounding cells, but instead are due to spatially-dependent differences in endothelial cell biology.

The genes with the highest fold-change in expression in endoderm-associated endothelial genes are shown on the left-hand side of Figure 6d. Two are endothelial genes: Pecam1 and Cdh5, both of which are associated with angiogenesis. Spatial expression of these genes is shown in Figure 6e (larger version in Figure S6.1b). Notch1 is more expressed in endoderm-associated endothelial cells, and this could reflect an increase in Notch signaling in the posterior of the gastruloid. [Chan et al 2017] demonstrated that Notch signalling can be sensitive to shear stress, raising the possibility that the differences in cell state we observe may be driven by differences in mechanical forces in the anterior and posterior. Although most endothelial cells are thought to be of mesodermal origin, some evidence suggests that, in the organogenesis of specific tissues like the liver, the endoderm can give rise to endothelial cells [49]. Furthermore, in [Rossi 2022] the authors show that in a gastruloid-like model specifically designed to model blood development, there is strong spatial adjacency between endothelial and endoderm cells. They hypothesize that these may be a subset of endothelial cells, specifically hemogenic endothelial cells (which have the potential to become blood progenitors). Our data demonstrate a molecularly driven organization distinct from the clustering we observed in the anterior and suggest that multiple mechanisms of endothelial specification could be modeled in gastruloids, even simultaneously within the same structure, although further characterization is needed to determine exactly what processes these unique endodermal/endothelial structures model.

Several other endothelial genes are instead differentially expressed in somite-associated endothelial cells: Nrp2, Tek, Apoe, and Cldn5. Although these genes have less obvious functional distinctions than the endoderm-associated genes, Nrp2 enables semaphorin receptor activity, including nervous system development and ventral trunk neural crest cell migration and Tek negatively regulates endothelial cell apoptotic process and response to retinoic acid (RA), which is known to be higher in the gastruloid anterior. Furthermore, a specialized population of endothelial precursors associated with somites was also observed in trunk-like structures, which show more tissue-like organization than gastruloids [Veenvliet et al. 2020].

Although endothelial cells have consistently been observed in single-cell measurements of gastruloids, their relative rarity has precluded in-depth analysis of subtypes or inference of spatial location. Our results strongly suggest that endothelial precursor formation, migration, and organization may all be modeled in 3D gastruloids, even without treatment with additional factors as in [Rossi 2021, 2022]; recent advances in 2D gastruloids have allowed modeling of cardiac and hepatic vascularization [45], and our data suggest that 3D gastruloids may similarly be adapted to model more specific aspects of hematopoiesis and vascularization. Early specification from a pool of mesodermal precursors is a hallmark of the endothelial lineage [47]; given the consistency with which we observe endothelial precursors, we speculate that this behavior is recapitulated in gastruloids, but further epigenetic measurements are required to validate this hypothesis” (See Revised Figure 6).

Finally, we tested several methods of clustering and calculating circularity, and determined that the difference in spatial organization of endothelial cells was not robust to changes in method and parameters, so we have chosen to remove that section of the figure and any conclusions drawn from the text.

(4) Figures 1c and 6b need statistical significance assessments.

We thank the reviewer for pointing out that without significance testing these plots are difficult to interpret. We have removed plot 6b (see response above about removing the circularity assessments). For plot 1c we appreciate that it is difficult to interpret which cell types vary more than others in their occurrence without significance testing. To address this we did two things: we first calculated the coefficient of variation for the proportion of each cell type across samples:

To calculate significance, we first considered that since these values are proportions, they must sum to 1 and changes in one cell type will affect at least one other cell type within the same sample. To properly account for this when applying statistical tests, we calculated the CLR-transformed proportion and tested all pairs of cell types for significant variation. The results are shown in the Author response image 2:

We added a plot to Supplemental Figure 1.4, highlighting the significantly varying pairs.

We also address said variation in the text:

“We sought to quantify variability in cell type composition between gastruloids. Previous single-cell datasets relied on pooling multiple gastruloids, thus obscuring the degree to which the overall cell type distribution was reflected in each individual gastruloid. However, recent single-cell measurements of individual gastruloids have suggested substantial gastruloid-to-gastruloid variation in cell type proportions [13]. Figure 1c shows distributions of cell type proportions across samples, and Figure 1d shows the coefficient of variation of these proportions. Individual gastruloid cell type distributions, including the proportion of cells that had insufficient reads to be confidently assigned a type, are shown in Figures S1.4b and c. We found that cardiac mesoderm, endoderm, and spinal cord cells had the greatest coefficient of variation in proportion between gastruloids (Figure 1d). To calculate statistical significance, we first performed a centred log-ratio (CLR) transform on the proportions, then looked for covariation between cell types across gastruloids. We found there was a statistically significant inverse correlation between the proportion of endoderm and NMP, presomitic mesoderm, and differentiation front (Figure S1.4d). We did not observe gastruloids that were as strongly neurally-biased as those reported in [13], but we did see some gastruloids with a relatively high proportion of spinal cord precursor cells (Figure S1.3a ii., xv., b vii.), and overall the proportion of spinal cord had a negative covariation with the mesodermally-derived cell types, consistent with the anticorrelation also reported in [13] (Figure S1.4).

The proportion of somite cells was significantly positively correlated with the proportion of presomitic mesoderm cells (covariation = 0.63, Figure S1.4d).”

(5) The article should include an analysis of Hox colinearity expression in these gastruloids as a validation of the system.

We thank the reviewer for pointing out the importance of these genes in validating the gastruloid system and agree that assessing their expression specifically would help readers assess data quality.

We analyzed the center of mass of expression along the AP axis for the Hox genes included in our panel (Hoxb6, Hoxc10, Hoxd1, Hoxb9, Hoxc8, Hoxc6, Hoxaas3, and Hoxb1). We highlighted these genes in Figure S1.3: their expression along the AP axis is consistent with previously reported expression in the tomoseq dataset from [van den Brink 2020]. A summary of the correlation coefficients for each individual gastruloid for all genes (blue) and the Hox genes (orange) is shown in the Figure S1.2. The Hox genes have similar correlation coefficients overall, although their variation is higher. This is likely due to differences in gastruloid pseudo-age; in future experiments we plan to include more Hox genes and use their expression to further classify gastruloids (see updated Figure S1.2).

We have updated the text with these new results:

“To assess the quality of our data, we first assigned an AP axis to each gastruloid using the expression of T, a canonical marker for the posterior (Figure 1a). When we compared how gene expression varied along the AP axis, we saw good agreement at a coarse-grained level with a previous study that sectioned gastruloids along the axis and analyzed gene expression in each section [2] (Figure S1.2a). The colinearity of the peak expression of Hox genes in our panel was also consistent with this dataset, with a median Pearson correlation of 0.695 (compared to 0.663 for all genes (Figure S1.2b).”

Reviewer #2 (Public review):

Summary:

This manuscript presents an ambitious and technically challenging spatial-transcriptomic atlas of 26 gastruloids using seqFISH. The authors introduce quantitative metrics (mixing score, exposure index, L-metric / scL-metric, spatial L-metric, triplets) to characterize spatial organization at multiple scales. The dataset is valuable, and several analyses are original, particularly the rank-based L-metric family for mutual exclusivity.

Strengths:

The authors generate one of the most detailed spatial transcriptomic datasets of gastruloids to date. They propose creative computational metrics (L-metric/scL-metric) to quantify mutual exclusivity of gene expression without predefined thresholds, and they explore organizational principles from single-cell topology to cluster-level structure. Many observations align well with known gastruloid biology, such as posterior robustness and anterior variability. The writing is generally clear, and the figures are rich.

We really appreciate the reviewer’s kind comments about the quality of the dataset and figures, and for pointing out the strengths of the new computational methods we developed in the analysis of this dataset.

Weaknesses:

Several central claims rely on metrics whose computation and justification are insufficiently explained, making it difficult to assess how robust or interpretable the results are. Many choices in the analysis appear arbitrary or are insufficiently motivated (normalization schemes, choice of parameters such as the number of neighbors, the distance cutoffs, hierarchical clustering setup, and so on). The interpretations of spatial consistency, gene-program inference, and endothelial heterogeneity are plausible but might be stronger than the evidence currently supports.

The manuscript would benefit from stronger benchmarking, quantification of uncertainty, and explicit controls for known artifacts in spatial transcriptomics (e.g., spillover, 2D slicing, cell type assignment entropy). The biological insights are promising, but since several depend on methodological assumptions that have not yet been demonstrated to be stable, they would benefit from clearer methodological explanation.

We thank the reviewer for spending time to give constructive and actionable comments, and we believe the manuscript is greatly strengthened and more consistent and clear as a result of the changes suggested.

The work is rich and could become a reference dataset. Then, clarifying and validating the quantitative methods will considerably strengthen the impact and reliability of the conclusions.

Reviewer #3 (Public review):

Summary:

Triandafillou and colleagues report a single-cell resolved spatial atlas of gene expression of 26 gastruloids. While previous work had analyzed either single-cell gene expression or spatially coarse-grained patterns of gene expression (van den Brink et al, 2020), the authors here use multiplexed sequential RNA FISH (seqFISH) to create the first gastruloid atlas, which is simultaneously spatially and cellularly resolved. This atlas adds to a growing list of resources cataloging gastruloid development (see also Suppinger et al 2023).

To analyze this dataset, the authors also describe a novel analytical framework. Their analysis centers around the 'L-metric', which measures the degree to which pairs of genes are either coexpressed or mutually exclusive. While this metric is similar to calculating correlations in gene expressions, it has important differences (including that it can, in principle, be asymmetric; although the authors symmetrize much of their analysis). In addition to the gene-centric L-metric analysis, the authors also analyze cells in their dataset according to the cell type entropy (an information-theoretical measure of confidence in cell type assignment) and the 'exposure index' (a measure of the similarity of nearest cellular neighbors).

Using this framework, the authors focus their analysis on two major features of development. The first is the differentiation of the bipotent neuromesodermal progenitor (NMP) cells in the posterior of the gastruloid into either presomitic mesoderm (PSM) or spinal cord SC lineages. They use L-metric analysis to compare overlap in marker genes used to separate NMP, PSM, and SC fates. They highlight that L-metric analysis can recover spatial patterns of gene expression (without explicit spatial information) and discern subtle features of marker genes beyond simple binning of cell types (e.g., that Epha5 expression in anterior NMPs may predict future SC differentiation).

The second is the formation of endothelial (spatial) clusters within the gastruloid. The authors highlight two subtypes of endothelial clusters: (1) smaller clusters within the somitic anterior region, and (2) larger clusters associated with endoderm. While the authors discern some subtle differences in gene expression between these two clusters, their different spatial patterns suggest a potential physiological difference that would not be captured in traditional droplet microfluidic-based scRNAseq pipelines.

Overall, this manuscript is a sophisticated and technically sound study that will provide a valuable beachhead for future studies of developmental patterning in gastruloids and organoids.

Strengths:

The major strengths of this study are the overall technical sophistication of the data set and analysis, as well as its potential generalizability to other developmental systems (both in vitro and in vivo). The data are extensively analyzed and reasonably interpreted, and this atlas makes good use of the variability in gastruloid development to extract the statistical structure of developmental processes. The L-metric offers a parameter-free tool to analyze transcriptomic datasets that could overcome the pitfalls of other approaches.

We really appreciate the reviewer’s kind comments about the quality of the dataset and figures, and for pointing out the strengths of the new computational methods we developed in the analysis of this dataset.

Weaknesses:

The major limitations of this study are the depth and novelty of the developmental processes studied. The authors provide very convincing proof-of-concept that their dataset can recover known features of gastruloid development, including NMP differentiation and endothelial development. However, further analysis and/or investigation would be required to discover new principles of gastruloid development and patterning.

We agree that the developmental processes studied here are not inherently novel, and we hope that by showing sufficient overlap with different, less highly resolved methods we have created a convincing document that highlights the potential for this technique to be used to analyze other systems. We appreciate the reviewer’s comments and that the manuscript is improved after making the suggested changes.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

(1) S1.2 plates shown individually, but unclear from which experiment.

The reviewer was right to point out this oversight — we have updated the figure (now S1.1a) with labels for the individual experiments:

(2) Figure 2 could include a clearer indication of the types of triples/ doublets to make it even more informative.

We thank the reviewer for pointing out that the types of triples were not clear — we’ve added a color key and more explanatory text to this figure in order which explicitly explains the type of triplets considered.

(3) Figures should be presented in order. Figure 3c is before 3b, etc.

We appreciate the reviewer’s attention to detail and have swapped these two panels so that their order in the figure reflects the order they are referenced in the text.

(4) Figure 3 is interesting, and the L-metric appears useful to pinpoint crucial genes that, when expressed, indicate a type transition has occurred. It would be great to test this with another set of cell types besides NMPs/presomitic/spinal cord.

We thank the reviewer for their interest in this biological transition, and agree that testing on another transition would be really interesting. There isn’t another set of cell types expected at this stage of gastruloid development that are predicted to have the same type of bifurcating differentiation. However, in an effort to address the spirit of this comment (that looking at other sets of cell types with the L-score would be interesting), we have used our analytical framework on a non-spatial, single-cell dataset from van den Brink et al 2020.

“scL-score analysis reveals cell type groupings and new transcription factor associations in a single-cell RNA-seq dataset

To test the generality of scL-score analysis, we analyzed a previously published dataset from [van den Brink 2020] where individual gastruloids (at the same stage as those used in this study) were pooled and subjected to single-cell RNA-seq analysis. After filtering for cell quality and common gene detection, we calculated the scL-score values using a cell-by-gene table of 14304 cells x 19075 genes.

To first test whether we could reproduce the results from this study, we performed hierarchical clustering on the scL-score difference vector (as previously described) on the set of 207 well-detected genes that were also present in our seqFISH gene panel. The resulting tree showed clustered cell types, similar to the tree produced with the expression data in this study (compare Figure S4.5a to Figure S4.4d). The cell types were clustered significantly more than expected by chance (Figure S4.5b).

Then, to test whether scL-score analysis would be effective for analyzing the entire dataset, we performed hierarchical clustering on all 19075 genes. We then examined the resulting heatmap (Figure S4.5c) for clusters of interest. We observed a cluster enriched for endothelial genes (Figure S4.5d), which contained some genes in our panel but many others which were not; this finding demonstrates that the clustering in Figure 4a is not solely due to the selection of genes in our seqFISH panel. We also observed a large cluster that contained genes associated with pluripotency or primordial germ cell fate (Figure S4.5e). Although some of the genes in this cluster were in our seqFISH panel, when we performed scL-score analysis we did not see them cluster with each other or with any other cell type genes. This lack of clustering implies that in our dataset cells that co-express these genes may be rare or too poorly detected to cluster strongly; however, the same analysis performed with more cells and genes showed association. This result demonstrates that clustering scL-score difference vectors can identify known cell-type-associated genes, even within transcriptome-scale data. Finally, we also found a small cluster showing strong co-expression of the transcription factor Gata4, a crucial regulator of the development of visceral and parietal endoderm, and two other genes: a predicted gene of unknown function (Gm43715) and Troponin C (Tnnc1) (Figure S4.5f). Intriguingly, Gata4 has been implicated in heart development (albeit in an indirect manner) [Watt 2004], and troponin C is important for cardiac muscle cell contraction and has been implicated in cardiomyopathy, although at a much later stage of development than that modeled by gastruloids [Li 2015].

Together, these results demonstrate that scL-score analysis is reproducible across datasets, even when different numbers of genes are compared. It effectively clusters genes associated with cell types, and can reveal developmental transitions. Moreover, increasing the number of cells and genes can reveal new clusters, some of which may predict novel regulatory interactions or spatial co-occurrence not previously observed.”

(5) Figures 4 and 5 felt exploratory, and I would recommend combining them into a single figure highlighting the usefulness of L-metric and its spatial version.

We thank the reviewer for the suggestion to merge the content of Figures 4 and 5. While we agree that they are thematically related, given the size of the heatmaps generated in the analyses, we were unable to combine them in a way that preserved the readability of the figure and stayed within the space constraints of the page size; thus, we have chosen to keep these as separate figures.

Reviewer #2 (Recommendations for the authors):

(1) Quantification methods require clearer formalization and justification

A key limitation is that the manuscript relies on several spatial metrics whose definitions are not sufficiently formalized.

(a) To evaluate biological interpretations, the reader needs a precise description of:

- How each metric is calculated (mixing score, exposure index, scL-metric, spatial L-metric).

- Why specific choices were made (normalizations, parameter values, distance thresholds).

- What the expected ranges and interpretations are.

We agree that these elements are crucial to interpreting quantitative metrics and thank the reviewer for their close read of the work. The reviewer had many comments on the exposure/mixing values calculated for Figures 1 and 2, and for the L-metric (now called L-score) values calculated for Figures 3, 4, and 5. To address the reviewers concerns we have done the following:

(1) Created a new, unified framework for calculating cell type exposure and mixing.

(2) Rewritten the methods section for this section with a particular emphasis on including elements the reviewer suggested, including specifically outlining normalizations and what they account for, distances chosen and the biological rationale behind them, and the range of values expected for each measure:

“Quantification of Cell Type Spatial Relationships: Exposure index

To characterize the spatial organization of cell types, we computed two related metrics: a pairwise exposure index matrix capturing type-specific spatial relationships, and a scalar mixing index summarizing overall spatial integration. For each cell, we identified neighbours as all cells whose centroids fell within a specified radius r of the focal cell's centroid. We chose a value of r of 16 μm, which gave an average of 5-6 neighbors per cell. We chose this value as it captures local interactions, which was the primary goal of these analyses. For each ordered pair of cell types (s, t), we calculated the exposure index as the proportion of type s cells' neighbors that are type t:

where Ns→t denotes the count of neighbour pairs in which the focal cell is type ‘s’ and the neighbour is type ‘t’, and Ns denotes the total number of neighbours across all type ‘s’ cells. Each row of the resulting exposure matrix sums to unity and represents a probability distribution over neighbour types for a given source type. To account for differences in cell type abundance, we normalized exposure indices relative to the expectation under random spatial arrangement:

Where pt is the proportion of cells that are type ‘t’. Normalized values of zero indicate exposure consistent with random mixing, positive values indicate spatial attraction (co-localization), and negative values indicate spatial avoidance, with a minimum of −1 representing complete exclusion.”

“Quantification of Cell Type Spatial Relationships: Mixing index

To summarize overall spatial integration across all cell types, we computed a mixing index defined as the fraction of neighbour pairs involving different cell types:

where N_cross is the number of neighbour pairs involving cells of different types and N_total is the total number of neighbour pairs. For normalization, we compared the observed mixing to the expectation under random spatial arrangement:

where

is the expected cross-type interaction rate given cell type proportions. Normalized values of zero indicate random spatial mixing, positive values indicate greater integration than expected (hyper-mixing), and negative values indicate spatial segregation.

Exclusion of untyped cells. When computing the mixing index, cells lacking confident type assignments were optionally excluded from both the numerator and denominator, ensuring the metric reflects only spatial relationships among typed cells. These cells were retained in the exposure matrix to quantify how typed cells interact with unclassified cells.

Statistical analysis of variance. To identify cell type pairs whose spatial relationships varied significantly across samples, we computed the variance in exposure indices across samples for each type pair. To account for the expected relationship between mean exposure and variance, we regressed log-variance against log-absolute-mean across all type pairs and computed residuals. Type pairs with residual variance exceeding the 97.5th percentile (two-tailed α = 0.05) were considered significantly variable, indicating spatial relationships that differ across samples beyond what is expected from sampling variation and composition differences.”

(3) We have also rewritten the methods for how the L-score is calculated, adding emphasis to where we normalize, what ranges of values are expected, and what the interpretation of these values are:

“Calculating the single-cell L-score

The single-cell L-score (scL-score) was computed for each ordered gene pair (gene A, gene B) within a single gastruloid. The cell-by-gene expression matrix was filtered to retain only cells with at least 2 detected transcripts for both genes. Cells were sorted in descending order by gene A's expression values; gene A thus serves as the reference distribution, and the score is asymmetric with respect to gene order.

Three reference distributions were constructed for gene B: (1) perfect coexpression, in which gene B's values were sorted in the same descending order as gene A; (2) perfect mutual exclusivity, in which gene B's values were sorted in ascending order; and (3) independence, in which every cell was assigned the mean expression value of gene B.

Cumulative sums of expression values were computed for gene A, for the observed expression of gene B, and for each reference distribution. The cumulative sum of each gene B distribution (observed and reference) was then plotted against the cumulative sum of gene A. This cumulative-sum-versus-cumulative-sum representation captures how gene B's expression accumulates relative to gene A's: if gene B's expression is concentrated in the same high-expressing cells as gene A, gene B's cumulative curve rises steeply at first; if concentrated in opposite cells, the curve rises steeply at the end. The area under each curve was calculated using trapezoidal integration and normalized by the product of gene A's and gene B's total expression, yielding four normalized areas: A_observed (observed relationship), A_positive (perfect coexpression), A_negative (perfect mutual exclusivity), and A_uniform (independence). This normalization ensures that scores are comparable across gene pairs with different overall expression levels (Figure S3.2a,b). The scL-score was then defined as follows:

If A_observed > A_uniform: scL-score = (A_observed − A_uniform) / (A_positive − A_uniform), yielding values in (0, 1].

If A_observed = A_uniform: scL-score = 0.

If A_observed < A_uniform: scL-score = −(A_observed − A_uniform) / (A_negative − A_uniform), yielding values in [−1, 0).

A score of +1 indicates perfect coexpression, −1 indicates perfect mutual exclusivity, and 0 indicates independence.

The scL-score was computed for all gene pairs in each gastruloid from the 05/07/2025 dataset (n = 18 gastruloids). The other two datasets (n = 8 gastruloids) were excluded due to lower transcript detection quality. To generate an average scL-score matrix, the analysis was restricted to a common set of 202 genes well-detected across all 18 gastruloids, and per-gastruloid matrices were averaged. Unless otherwise noted, a symmetrized scL-score was used: scL-score_sym(A, B) = [scL-score(A, B) + scL-score(B, A)] / 2.

Calculating the spatial L-score

The spatial L-score extends the scL-score to spatial regions. For each gene, a kernel density estimate (KDE) was fitted over all detected transcript spots and evaluated on a regular square grid spanning the gastruloid. Bin side length was set to twice the median nearest-neighbour distance between detected spots, calculated separately for each gastruloid. Spatial bins were ranked by KDE-derived density and processed identically to the scL-score calculation. Low-density bins were not filtered, as KDE smoothing produced non-zero density values throughout the imaging area. The spatial L-score was symmetrized as for the scL-score, except when displaying asymmetric heatmaps.

Hierarchical clustering of L-score matrices

Gene-gene distances were defined as Euclidean distances between L-score vectors. Agglomerative hierarchical clustering was performed using Ward's linkage criterion (scipy.cluster.hierarchy.linkage, method='ward', metric='euclidean'). This approach operates on L-score vector differences rather than on pairwise L-score values directly, and therefore does not require the L-score itself to satisfy the properties of a mathematical distance metric; the Euclidean distance between L-score vectors is non-negative and symmetric by construction, satisfying the requirements of Ward's method. Heatmaps display pairwise L-score values, not vector distances.

We applied this clustering procedure to the following gene sets:

(1) A subset of NMP, presomitic mesoderm, and spinal cord marker genes in one gastruloid (n = 36 genes; Figure 3g).

(2) All well-detected genes excluding cell cycle genes, averaged across all gastruloids (n = 166 genes; Figure 4 and Figure S4.3a).

(3) All well-detected genes common to all gastruloids, averaged across gastruloids (n = 202 genes; Figures S4.1a, S4.3b, S5.1a).

(4) All well-detected genes excluding cell cycle genes in one example gastruloid (n = 171 genes; Figures 5b, S5.2a).

(5) All well-detected genes common between our seqFISH panel and those that were detected in > 3 cells in scRNA-seq data from [XXX] (n=207 genes; Figure S4.5a).

(6) All genes detected in > 3 cells in scRNA-seq data from [XXX] (n=19075 genes, Figure S4.5c-f).

In Figure S4.2a,b a transformation of the L-metric values was used to cluster genes. The scL-scores were averaged across gastruloids as described above, and then each pairwise scL-score was transformed to a distance-like value with(1 - scL)/2. The matrix was then symmetrized as described previously. Agglomerative hierarchical clustering was performed directly on this transformed gene-gene distance matrix using Ward's linkage criterion (scipy.cluster.hierarchy.linkage, method='ward', metric='euclidean').”

(4) We have added an illustrative figure about how the L-score is calculated which defines expected behaviour for several cases, gives a visual explanation of the process, and shows several extreme behaviours and their biological interpretation (see Revised Figure S3.2).

(b) For example:

- Mixing score: The normalization is unclear. Why only 1-nearest neighbor instead of k-NN? Why not consider existing spatial-autocorrelation metrics such as Moran's I, which would also apply to gene-level mixing?

- Exposure index: The normalization makes the metric unbounded (e.g., exposure > 1 when local frequency > global frequency). It is unclear whether this behavior is intended. Since exposure to self is meaningful, the same metric could replace the mixing score and simplify the framework. The choice of k = 5 is not justified; parameter-free approaches like Delaunay triangulation could avoid arbitrary cutoffs. If k-nn is preferred, then the robustness of the score to change the k value should be studied.

These issues make it difficult to interpret the magnitude of reported effects or compare them across studies.

The reviewer makes an excellent point — we have completely overhauled this analysis in the following way to address the issues raised:

(1) Created one unified metric (see points 1 and 2 above) which considers for every cell, the identity of its neighbors in a 16 um radius (on average 5 or 6 neighbors for each cell in each gastruloid). This value was chosen so that in most cases, the cells in the immediate vicinity of a cell were considered, but not those further out (i.e. the measure is sensitive to close interactions rather than far ones). We made a matrix of all interaction pairs for a given gastruloid, with diagonal elements representing within-type interactions and off-diagonal elements representing cross-type interactions. We have replaced the previous description with the following:

“Several studies of gene expression in gastruloids have used pooled measurements to infer the AP axis-location of genes and cell types [3,7,13,24,27] and our data are largely consistent with these lower-resolution findings (Figure S1.3a). Yet it is obvious from individual gene staining [1,4,8,28] and our detailed 2D maps of cell identity and location that gastruloid organization is much more complex than the average order of cells along the AP axis. We thus needed an analytical method for quantifying spatial organization beyond distributions along the AP axis. To further characterize spatial organization, we sought to quantify the degree to which cells were mixed in each gastruloid, and how that mixing might vary between gastruloids. For each cell in each gastruloid, we counted the interactions between that cell and all its neighbours within a 16 μm radius (on average 5-6 neighbors per cell), and summarized all these interactions for all cells in the gastruloid in a matrix, normalizing each element by the frequency of the cell type considered to be the ‘neighbour’ in the interaction.”

(2) To quantify overall mixing, we calculate the sum across types of the frequency of self interactions (normalized to the total interactions) and then take the inverse (1-M). We then normalize this value to the expected cross-type interactions predicted from random mixing (i.e. the proportion of that cell type).

Because the density of cells is fairly consistent across the gastruloids, w is very close to p (the proportion of that type).

Mnorm=Mrow−EcrossEcrosswhereEcross =1−∑tpt2 is the expectation of cross-type rate with random mixing.

Mixing ranges from -1 (totally segregated) to +1 (more mixed than random, i.e. there is attraction between unlike types). 0 is completely random, and negative values indicate that cell types within that gastruloid tend to cluster. We added the following to the text to explain this:

“To quantify overall mixing, we calculated the sum (across types) of the frequency of self interactions, normalized to the total interactions) and then took the inverse. We normalized this value to the expected cross-type interactions predicted from random mixing (i.e. the proportion of the neighbouring cell type). This gave us, for each gastruloid, a value that we call the mixing index that ranged from -1 (totally segregated) to +1 (totally mixed with less frequent self-interactions than expected from chance). A mixing index of 0 indicates a random distribution, i.e., neighbour frequency is exactly what would be predicted by that cell type’s frequency alone.”

We also edited the following description of the overall distribution of mixing indices:

“The mixing index values range from -0.50 to -0.22 (Figure 2a). All gastruloids had a negative mixing index, indicating that they all, on average, had more like-cell type interactions than would be expected given random mixing of types. However, we note that there is a ~14% difference in the mixing index across gastruloids, meaning some variation in overall mixing is present.”

(3) The exposure index for a given pair can be found from the off-diagonal elements of the interaction matrix, and the normalization means it represents relative overexposure/clustering (positive values) or underexposure/avoidance (negative values). The minimum value is -1 and the maximum is (1-pt)/pt. We changed the description of how the exposure index is calculated to reflect this unified method of quantification:

We noted that the off-diagonal elements of the matrix we used to calculate the mixing index were informative about cell type-cell type interactions. Specifically they quantify the degree to which each cell type (source) is exposed to another cell type (neighbours). To assess the overall frequency of cell type-cell type interactions, we first pooled the data from all gastruloids together into one interaction matrix (Figure 2b). The measure can range from -1 (no interactions at all), with higher values indicating a greater frequency of being found in close proximity. It is asymmetric in that the exposure of cell type A to B may not be the same as the exposure of cell type B to A.

We have updated all of the quantification in Figures 1 and 2 with these new measures.

(2) L-metric: unclear justification and interoperability

(a) First of all, even if it is not a major issue, the L-metric is not a "metric" at least in the mathematical sense of a metric since a metric is always positive. The L-metric is central to several major conclusions (gene exclusivity, modules, spatial organization), but its conceptual basis and computational steps need more justification.

We thank the reviewer for pointing this out and have changed “L-metric” to “L-score” throughout. We have also endeavoured to clarify the conceptual basis and have fleshed out the various computational steps as outlined in more detail in our responses below.

(b) Several steps (ranking, cumulative curves, area under the curve) are difficult to interpret biologically It is unclear why each transformation is required and how it responds to common scenarios (highly expressed genes, correlated vs mutually exclusive patterns).

Since the metric is rank-based, two genes that are both highly expressed in all cells may show low L-metric despite being truly correlated.

We appreciate the reviewer’s comments about both the interpretation of the scL-score calculation and how it behaves in common expression scenarios, particularly for genes that are broadly expressed across many cells. To address these points, we generated a set of simulated examples spanning five scenarios: ubiquitously expressed genes with similarly high average expression, ubiquitously expressed genes with differing average expression, ubiquitously expressed genes with similarly low average expression, genes coexpressed across a subset of cells rather than all cells, and genes generally expressed in opposite subsets of cells. For the first three simulations, we independently sampled two genes across 50 cells using Poisson distributions with mean expression set to 100 or 50 (to simulate a gene with high or low average expression, respectively), without any expression bias towards any subsets of cells. For the latter two simulations of dependent expression relationships, we first sampled gene 1 across 50 cells using a Poisson distribution with mean expression set to 4, then generated gene 2 from gene 1 by sampling from cell-specific Poisson distributions fitted to either generally match or oppose the transcript count obtained for gene 1 in that cell. These simulations show that genes can independently appear broadly coexpressed at the population level (regardless of average expression) simply by being ubiquitously expressed, yet still receive low scL-score values. The simulations of dependent coexpression or mutually exclusive expression relationships receive scL-score values near +1 and -1, respectively. These results align with the reviewer’s prediction, but they reflect why we designed the L-metric to follow a rank-based methodology since they preserve the specificity of the metric’s upper bound (+1) for detecting non-random coexpression relationships rather than chance coexpression relationships resulting from independently ubiquitous expression. The results of these simulations are depicted (See Revised Figure 3.3).

We have made the following edits to the text to specifically address the case the reviewer raised about highly expressed genes:

“To this end, we developed a pairwise metric between genes that reported the degree of mutually exclusive expression. It is calculated by rank ordering cells by the expression of one gene and measuring the degree to which the expression of the other gene is anti-rank-ordered (see Methods for details and Figure S3.2a for a visual explanation of how the measure is calculated). We call this measure the “single-cell L-score” (scL-score) because when the per-cell expression of mutually exclusive genes was plotted against one another, the data made an L shape (Figure 3e, right). A value of -1 represents perfectly mutually exclusive expression, which only happens when the genes are never found in the same cell. Higher values indicate more co-expression. Genes that are ubiquitously expressed without a strong correlative relationship between them will have a score of ~0. The maximum possible value is 1, which is obtained when both genes are expressed in a subset of all cells, and are only ever found together in those cells. We refer to this as ‘perfect co-expression’. This scale, which ranges from -1 (mutually exclusive) to 1 (perfect co-expression) captures the range of possible relationships between genes. Our expectation is that ubiquitously expressed genes like cell cycle and housekeeping genes will, due to the rank-ordered nature of the L-score calculation, have L-scores consistently close to zero no matter which genes they are compared with, whereas genes that are specifically associated with a single cell type will have an scL-score value close to -1 when compared with genes specific to other types, but higher values when compared with genes associated with the same cell type. The results of our simulations confirmed these hypotheses (Figure S3.3a).

To benchmark this measure against existing exclusivity or coexpression measures, we calculated the Exclusively Expressed Index (EEI) [Nakajima 2021] and Coefficient of Expression (COEX) [Galfrè 2021] for the same simulated datasets (Figure S3.3b) and a subset of NMP/presomitic mesoderm/spinal cord genes (Figure S3.4a). All three methods were able, to some extent, to distinguish mutual exclusivity from coexpression, but the scL-score provided clearer separation between these different relationship types; a more detailed description of the analysis is included with Figure S3.3.”

(c) Interpretation of L-metric values is ambiguous

What does 0 represent? Randoms? Co-expression? Is 1 the strongest exclusivity? The manuscript currently mixes "co-expression" and "mutual exclusivity" scales.

We agree with the reviewer that clearly defining what values of the L-score mean is critical to understanding the text. We have added a more explicit discussion of this in the text (excerpted from the response to 2c):

“A value of -1 represents perfectly mutually exclusive expression, which only happens when the genes are never found in the same cell. Higher values indicate more co-expression. Genes that are ubiquitously expressed without a strong correlative relationship between them will have a score of ~0. The maximum possible value is 1, which is obtained when both genes are expressed in a subset of all cells, and are only ever found together in those cells. We refer to this as ‘perfect co-expression’. This scale, which ranges from -1 (mutually exclusive) to 1 (perfect co-expression) captures the range of possible relationships between genes.”

And made a figure representing visually how the L-score is calculated which shows the behaviour and biological interpretation of several extreme values and an example of how the L-score is calculated. See new Figure S3.2:

We also edited figure captions where we referred to plots as ‘co-expression’ plots, since in some cases the plots showed genes that were mutually exclusive or not expressed together in most cells. Figure S3.1:

“c. Spatial distribution of the expression of Nkx1-2 and Rfx4 in an example gastruloid.”

In all other cases we checked, we used the term “co-expression” to mean the opposite of mutually exclusive, as outlined in the definition above.

(d) Additional issues also limit interpretability - Benchmarking is missing.

- No tests on synthetic datasets, negative controls, or curated examples.

- Prior exclusivity methods (EEI, COTAN) routinely benchmark against ground truth; this is now standard.

- The code for the L-metric seems to be missing in the repository.

We appreciate this suggestion offered by the reviewer as benchmarking against prior exclusivity-oriented methods provides an important comparison for clarifying both where the scL-score agrees with existing approaches and where it offers distinct advantages. To address this, we explicitly compared the scL-score to the Exclusively Expressed Index (EEI), which is bounded below by 0 and increases with mutual exclusivity, and to the signed coefficient of coexpression (COEX) from the COexpression Table ANalysis (COTAN) framework, in which positive values indicate coexpression, negative values indicate mutual exclusivity, and values near 0 indicate little structured relationship. We performed this comparison using seven simulated scenarios as well as four representative gene pairs from one gastruloid sample (2025-05-07_roi2). In the two mutually exclusive simulations, all three methods detected exclusivity. In the three simulations of genes independently expressed in all cells (high_high, high_low, low_low), EEI and COEX were 0, while the scL-score remained close to 0 (0.102, -0.225, and -0.025, respectively), consistent with little structured relationship. In the weak coexpression simulation, the scL-score was positive (0.770), EEI remained low, and COEX was also positive (0.340), indicating detectable but modest coexpression. In the perfect coexpression simulation, the scL-score reached 1.000, EEI was 0, and COEX was strongly positive (1.000). Together, these simulations show that all three methods detect strong mutual exclusivity, and both scL-score and COEX distinguish positive coexpression from exclusivity and from unstructured expression.

We then applied the same comparison to four gene pairs from one gastruloid sample (2025-05-07_roi2). All three methods were able, to some extent, to distinguish mutual exclusivity from coexpression, but the scL-score provided clearer separation between these different relationship types. Pax6-Eogt, Rfx4-Eogt, and Nkx1-2-Rfx4 all showed opposing expression by scL-score, with values of -0.572 and -0.526, -0.970 and -0.955, and -0.537 and -0.615, respectively. EEI detected exclusivity most strongly for Rfx4-Eogt (0.171), but gave values of 0 or approximately 0 for the other two pairs, while COEX was negative for all three pairs (Pax6-Eogt: -0.235, Rfx4-Eogt: -0.350, and Nkx1-2-Rfx4: -0.106), consistent with opposing expression, with strongest signal for Rfx4-Eogt. By contrast, Cdx4-Cdx2 showed moderate levels of coexpression by scL-score (0.402 and 0.424) and EEI (0), but COEX indicated that they were not coexpressed (-0.331).

These comparisons also clarify the practical advantage of the L-metric over the EEI and COTAN frameworks. EEI is based on binary zero/non-zero quantification and is therefore designed specifically to measure exclusivity rather than coexpression. COEX provides a signed value and, in our simulations, tracked both exclusivity and coexpression; however, on representative gene pairs from one gastruloid sample, scL and COEX diverged in magnitude for highly exclusive expression relationships (Rfx4-Eogt) and sign for a coexpression relationship (Cdx4-Cdx2), motivating our introduction of a signed measure based on the mutual exclusivity of expression with the scL-score (see New Figures S3.3 and S3.4).

We have updated the text to address the reviewer’s concerns: we benchmark using simulations of commonly occurring scenarios (such as varying expression levels, degree of mutual exclusivity, and amount of noise present in the relationship between the two genes in question) as the reviewer suggested. We also provided a direct comparison to an earlier exclusivity measure (EEI):

“To this end, we developed a pairwise metric between genes that reported the degree of mutually exclusive expression. It is calculated by rank ordering cells by the expression of one gene and measuring the degree to which the expression of the other gene is anti-rank-ordered (see Methods for details and Figure S3.2a for a visual explanation of how the measure is calculated). We call this measure the “single-cell L-score” (scL-score) because when the per-cell expression of mutually exclusive genes was plotted against one another, the data made an L shape (Figure 3e, right). A value of -1 represents perfectly mutually exclusive expression, which only happens when the genes are never found in the same cell. Higher values indicate more co-expression. Genes that are ubiquitously expressed without a strong correlative relationship between them will have a score of ~0. The maximum possible value is 1, which is obtained when both genes are expressed in a subset of all cells, and are only ever found together in those cells. We refer to this as ‘perfect co-expression’. This scale, which ranges from -1 (mutually exclusive) to 1 (perfect co-expression) captures the range of possible relationships between genes. Our expectation is that ubiquitously expressed genes like cell cycle and housekeeping genes will, due to the rank-ordered nature of the L-score calculation, have L-scores consistently close to zero no matter which genes they are compared with, whereas genes that are specifically associated with a single cell type will have an scL-score value close to -1 when compared with genes specific to other types, but higher values when compared with genes associated with the same cell type. The results of our simulations confirmed these hypotheses (Figure S3.3a).

To benchmark this measure against existing exclusivity or coexpression measures, we calculated the Exclusively Expressed Index (EEI) [Nakajima 2021] and Coefficient of Expression (COEX) [Galfrè 2021] for the same simulated datasets (Figure S3.3b) and a subset of NMP/presomitic mesoderm/spinal cord genes (Figure S3.4a). All three methods were able, to some extent, to distinguish mutual exclusivity from coexpression, but the scL-score provided clearer separation between these different relationship types; a more detailed description of the analysis is included with Figure S3.3 and Figure 3.4.”

We added the following explanatory text to Supplemental Figure S3.4:

“We compared the scL-score to two existing measures of exclusivity. The Exclusively Expressed Index (EEI) (Nakajima et al. 2021) is bounded below by 0 and increases with mutual exclusivity. EEI is computed from binary zero/non-zero quantification and is designed specifically to measure exclusivity but not coexpression. The coefficient of coexpression (COEX) from the COexpression Table ANalysis (COTAN) framework (Galfrè et al. 2021) can also be used to quantify relationships between genes: positive values indicate coexpression, negative values indicate mutual exclusivity, and values near 0 indicate little structured relationship. In the two mutually exclusive simulations, all three methods detected exclusivity. In the three simulations of genes independently expressed in all cells (but with varying relative expression levels), EEI and COEX were 0, while the scL-score remained close to 0 (0.102, -0.225, and -0.025, respectively), consistent with little structured relationship. In the weak coexpression simulation, the scL-score was positive (0.770), EEI remained close to 0, and COEX was positive (0.340), indicating detectable but modest coexpression. In the perfect coexpression simulation, the scL-score reached 1.000, EEI was 0, and COEX was strongly positive (1.000). Together, these simulations show that all three methods detect strong mutual exclusivity, and both scL-score and COEX distinguish positive coexpression from exclusivity and from unstructured expression.”

We added the following explanatory text to Supplemental Figure S3.4:

“We calculated the scL-score, EEI, and COEX for four gene pairs from one gastruloid sample (2025-05-07_roi2). The scL-score consistently delineated gene pairs possessing opposing expression profiles, while EEI was not always able to measure those exclusivity patterns (Pax6-Eogt: scL-score=-0.572 and -0.526, EEI=0; Rfx4-Eogt: scL-score=-0.970 and -0.955, EEI=0.171; Nkx1-2-Rfx4: scL-score=-0.537 and -0.615, EEI~0). Only the scL-score was able to detect the coexpression pattern present between the positively associated expression profiles of Cdx4 and Cdx2 (scL-score=0.413, EEI=0, COEX -0.331) (shown visually in Figure S3.4a). Thus, while EEI was informative for measuring gene expression relationships characterized by mutual exclusivity, the scL-score more clearly separated positive, random, and mutually exclusive relationships on a single signed bounded scale. The COEX value trended in the opposite direction than expected, but this may be due to the fact that it cannot be calculated on single-gene pairs and necessarily uses information from the entire count table, which here only consisted of 6 genes. These comparisons combined with the simulations in Figure S3.3, clarify a conceptual advantage of the L-metric over the EEI and COTAN frameworks. In contrast, by leveraging the ranked structure of transcript counts across cells, the L-metric framework does not binarize expression and does not require fitting a parametric distribution. It can be calculated on single gene pairs, and is more sensitive to mutual exclusivity.”

We also amended the Data and Code Availability section to include a specific reference to the L-metric package that was previously missing:

“All code used to process the raw data and generate figures, as well as the processed data and figures can be found at the following link :

https://www.dropbox.com/scl/fo/bchkqlbcjb8ub9m606did/AIudcWZaC566toXzb2L-jXc?rlkey=u0wgtkq8j erxoqb5ump599oip&dl=0

Additional custom scripts used to process the raw seqFISH data can be found on GitHub:

https://github.com/arjunrajlaboratory/NimbusImage/

The code for calculating the L-score can be found on GitHub:

https://github.com/arjunrajlaboratory/l-metric

Images of all gastruloids generated for this study, as well as single-channel seqFISH images with segmentation and annotations are available here:

https://app.nimbusimage.com/#/project/69d3fa8f1be4701f5fab6359 Raw seqFISH images are available upon request.”

(e) Clustering using L-metric vectors

In this study, the authors use hierarchical clustering to group genes according to the L-metric. This choice is reasonable: hierarchical clustering provides a natural representation of similarity relationships across multiple scales, and the L-metric captures a form of signed dissimilarity between genes. However, this approach raises an important issue. Standard hierarchical clustering methods typically assume a non-negative metric, whereas, as noted earlier, the L-metric can take negative values, meaning it does not strictly satisfy the requirements of a metric in the mathematical sense.

To the best of our understanding from both the text and the source code, the authors address this issue by defining the distance between two genes A and B as the Euclidean distance between two vectors: the L-metric values from A to all other genes, and from B to all other genes. Although this procedure is mentioned in the manuscript, it is neither justified nor accompanied by any discussion of how such a distance should be interpreted. It is not the direct distance between gene A and B, but rather whether A and B have a similar L-metric to all other genes. These two gene distances are not without overlap, but they are not the same.

This choice magnifies the interpretability issue: readers must understand two layers of transformations. If the [-1,1] range poses problems for hierarchical clustering, simple transformations (e.g., 1 − L) or alternative clustering methods could avoid these issues.

Given that the L-metric underlies major biological inferences (novel gene modules, spatial subclusters, endothelial states), clearer justification and benchmarking are essential. We think this can lead to more consistency in the spatial metrics.

We really appreciate that the reviewer took the time to understand our proposed method thoroughly, and apologize for any confusion resulting from a lack of clarity in how it is calculated, and the language used to describe it. The reviewer is absolutely correct that it is not a metric in the mathematical sense; we have replaced the word ‘metric’ with the word ‘score’ throughout the text.

The reviewer also raised concern about how the hierarchical clustering was performed, and they were absolutely correct about what the vectors represent—they are, as the reviewer states, “not the direct distance between gene A and B, but rather whether A and B have a similar L-metric to all other genes”. The heatmaps in figures 3, 4, and 5 are intended to cluster genes that have similar expression patterns, i.e. similar L-score values with all other genes. The reviewer pointed out that this transformation wasn’t clear, so we have added an explicit explanation of what the vectors represent, as well as an explanation of why we were interested in how these vectors, which represent a ‘fingerprint’ of how the gene interacts with all other genes, clustered (because this is in the section discussing NMP differentiation we focus on a specific subset of genes here, but later apply to the entire panel):

“For each gene annotated as belonging to any of the three cell types (NMP, PSM, or spinal cord), we calculated a vector of scL-score values with all other genes. Two genes that play similar regulatory or functional roles would be expected to have similar patterns of coexpression and exclusivity across the full gene panel and thus similar L-score vectors. We reasoned that the Euclidean distance between these vectors could be used instead, as it represents the degree to which A and B have a similar scL-score to all other genes considered and satisfies the requirements of a distance measure for the purposes of clustering. We performed hierarchical clustering using the distance between these vectors; the clustering therefore groups genes by the overall similarity of their coexpression profiles rather than by any single pairwise relationship. A heatmap of this clustering (with the pairwise scL-score values displayed between individual genes displayed for clarity) is shown in Figure 3g.”

We also appreciate the reviewer’s suggestion that alternative transformations of the scL-score may improve clustering interpretability. We tried using the reviewer’s suggestion of doing a simple transform: we averaged the scL-score matrices across the 18 gastruloids using the shared gene panels, transformed each scL-score from the original [-1,1] scale to a [0,1] scale using (1-scL)/2, symmetrized the resulting matrix so that each gene pair was represented by a single value, and then performed hierarchical clustering directly on this gene-by-gene distance matrix using average linkage. We used this analysis to test whether a more direct distance-based approach would change the gene groupings recovered by our original clustering method.

This alternative approach largely recovered the same cell type-associated groupings, but the separation between branches in the dendrogram was smaller, making fine-scale ordering harder to interpret. We measured this by evaluating the average cell type dispersion, measured in terms of additive branch length, which was 0.682 and 0.671 for the 166-gene and 202-gene panels, respectively. Since the branch separation on our original dendrograms was greater and thus representative of more robust groupings, we chose to continue using hierarchical clustering based on Euclidean distances between scL-score vectors.

We comment on this alternative transformation we tried in the next section, when considering the clustering of the entire gene panel. We feel this is appropriate as the motivation for looking at the difference vector was derived from expected behaviour of a smaller set of genes, and as the reviewer pointed out it is not clear that that expectation should or would hold for the entire panel.

“To generate the heatmap shown in Figure 4a and S4.1a, we used the same clustering method as described earlier with the Euclidean distance between scL-score vectors. However, we also tried clustering directly on the scL-scores themselves, by transforming each scL-score from the original [-1,1] scale to a [0,1] distance-like scale using a (1-scL)/2 mapping (Figure S4.3a,b). The results were largely consistent, however the cophenetic distance scale was relatively compressed when the transformed values were used (Figure S4.3a,b). This shallow structure implies that many branches are separated by only modest distances, so fine-scale ordering within the dendrogram should be interpreted more cautiously than the larger-scale cell type block structure. We chose to continue using Euclidean distance-based hierarchical clustering of scL-score profiles, where cell type grouping is observed alongside larger cophenetic separations between clusters” (See New Figure S4.3).

(3) Claims of "remarkably consistent" spatial organization are stronger than the data currently support

(a) The manuscript emphasizes reproducible organization across gastruloids, but several factors complicate this interpretation.

We agree that the distinction between what is reproducible/invariant between gastruloids and what varies was not clear in the original manuscript. To address this, we have updated the text to more explicitly distinguish between the two categories. The other suggestions made by the reviewer to sharpen the quantitative measures to strengthen these claims was very helpful and we appreciate the thought put into them, and have used that framework (emphasizing the statistically significant variations and consistencies) in summarizing our findings:

“Variation in cell type abundance and organization is structured and concentrated in specific cell types

We have demonstrated that some aspects of gastruloid composition and spatial organization are consistent across gastruloids, while others are more variable. Consistent features include proportions for NMP, presomitic mesoderm, somite, and paraxial mesoderm, whose coefficients of variation were lower than other cell types (Figure 1d). Organizationally, all cell types across gastruloids are more physically clustered than random (Figure 2a), and the order in which cell types are found along the AP axis has statistically significant high agreement between gastruloids as measured by Kendall’s W (Figure S1.5c). At the local neighbourhood scale, we found that most cell type interactions were conserved across gastruloids (Figure S2.1c). At the local scale, across individual gastruloids, we found many motifs of three cells that were statistically enriched over random, suggesting a conserved local order (Figure 2c). While the normalized distance along the AP-axis of all cell types significantly varied compared to a bootstrapped null (Figure S1.5a), the effect size was small, and decreased in almost all cases when normalized to gene expression (of T) in addition to morphology (Figure S1.5b).

However, there were also variable features. The proportion of cardiac mesoderm, endoderm, and spinal cord had the highest coefficient of variation between gastruloids (Figure 1d). Because proportions must sum to one, a change in the proportion of one cell type is necessarily linked to changes in others; we performed centred log transformation and looked for statistically significant covariation. Of all possible pairings, the following proportions had a significantly negative correlation across samples: endoderm/differentiation front, NMP/endoderm, presomitic mesoderm/endoderm, none/endothelial, and spinal cord/endothelium. This result shows that the proportions of these cell types predictably co-vary between samples, potentially suggesting some kind of biological trade-off in cell type specification or organization (Figure S1.4d).

Across gastruloids, intra-cell type interactions (degree of clustering) of spinal cord, endoderm, and differentiation front vary (Figure S2.1b). This variation suggests that these cell types may be patterned differently between gastruloids. For example, the local motif of 3 endoderm cells found next to one another was statistically enriched within some but not all individual gastruloids, and by definition is completely absent from gastruloids lacking endoderm (Figure 2c). We interpret this contrast to mean that when endoderm is found in a gastruloid, it is consistently patterned at a local level, but may vary more at a global level. This interpretation is concordant with the findings from [Farag 2024], which demonstrate several distinct classes of endoderm organization in gastruloids.

To summarize, while changes in the amount of individual cell types can vary, these changes are in most cases explained by variations in morphology and molecular characteristics (such as anterior:posterior ratio and the expression of morphogens like T). For patterning, we found that, in most cases, global patterns were conserved, but there were variations in local patterning that may lead to variable meso-scale organization of specific cell types, particularly those found in the middle of the anterior-posterior axis.”

(b) Possible selection bias. Only elongated, QC-passing gastruloids were retained; 18/26 datasets remain. seqFISH runs with uneven housekeeping signals were excluded.

We agree with the reviewer that our data are elongated, QC-passing gastruloids, although these represent two sources of variation (biological and technical respectively). Our goal was to characterize the structures considered to be equivalent and morphologically normal in gastruloid studies, and to characterize gene expression and cell type variation within this category, and we have attempted to signal this to readers by consistently including language like “morphologically normal” and “elongated”. We have further updated the language in the manuscript to emphasize this point:

“To measure the spatial distribution of gene expression, we prepared gastruloids using mouse E14TG2a cells and a standard protocol (see Methods). We harvested mature gastruloids after 120 hours of growth. To ensure consistency we checked that the proportion of the gastruloids that formed correctly was the same or greater than the median of all experiments (Figure S1.1a). Although there was variation in the length, width, and relative amounts of anterior and posterior tissues in the gastruloids considered, they were within the range of what would be qualitatively considered a ‘morphologically normal’ gastruloid [1,10].”

In regards to the exclusion of datasets, the only time 18 out of 26 were used was when calculating the averaged L scores for all genes in Figures 4 and 5. In this case we used all 18 gastruloids from the seqFISH run performed on 4/4/2025; this dataset had the highest spot counts due to protocol improvement between runs, and integrating the datasets with very different spot counts was problematic because a minimum expression level is needed to calculate L scores. We used all 26 samples for the spatial metrics calculated in Figures 1 and 2 (Figures 3 and 6 focus on specific gastruloids). We have added additional labels in Figures 1, 2, 3, and 6 to make clear when all 26 datasets are used and when only a subset is used.

(c) Gastruloids are known to be variable; restricting to morphologically "normal" samples could inflate apparent regularity.

We thank the reviewer for this observation and agree that the degree of variability among gastruloids is an important consideration. As stated in response to b), our goal was to characterize the structures considered to be equivalent and morphologically normal in gastruloid studies, and to characterize gene expression and cell type variation within this category. The rate of occurrence of ‘normal’ gastruloids in our hands is ~80% (Figure S1.1a). We agree that it would be interesting to consider how variations from this baseline affect cell type composition and arrangement, and while we make no claims about it in this paper, we have updated the introduction to highlight this point:

“To address these gaps, and to create a systematic, high-resolution dataset of gene expression in gastruloids considered to be morphologically normal, we developed a spatially resolved, single-cell molecular map of the location, identity, and gene expression of cells within 26 individual gastruloids with normal morphologies. We found that despite some morphological variability within the qualitative category of elongated and polarized, “normal” gastruloids had largely reproducible cell type composition.”

(d) Partial lack of statistical validation. The manuscript shows descriptive consistency but no formal tests across runs or batches (e.g., mixed-effects models, ICCs, leave-one-run-out validation).

We agree with the reviewer that a quantitative comparison between batches is important. We have added the following to the text to address this point:

“To address potential batch effects due to biological differences between runs, we examined brightfield images of all the gastruloids generated for each experiment (529 total gastruloids across 6 plates on 3 different days), segmented them, and quantified morphological characteristics. When we embedded all 529 gastruloids into PCA space, there was near-complete overlap between all groups, with the exception of one plate from 9/1/2024, which was slightly higher in PC1. Figure S1.1b shows this embedding, and examples of gastruloids at the extreme ends of PCs 1 and 2. We note that the samples collected on 9/1/2024 were on average smaller than the other two experiments, but spanned the same range of elongation (Figure S1.1c). Interestingly, the final size as measured by cross-sectional area of a brightfield image of the gastruloid did not correlate with the initial seeding number (the experiment on 4/4/2025 used 100 starting cells and the other two experiments used 300). Previous studies have demonstrated that the gene expression differences between gastruloids seeded with 100 and 300 cells is extremely small [Bennabi 2025]” (See Revised Figure S1.1a-c).

(e) 2D sampling limitations. Spatial metrics rely on a single imaging plane chosen as the "midplane," but z-position varies between gastruloids. AP projections, mixing, and triplet analyses could all be sensitive to z-plane choice. Prior work shows that 2D slices can underestimate distances and contacts by large margins (https://pmc.ncbi.nlm.nih.gov/articles/PMC5522766). These limitations should be acknowledged explicitly.

We agree that sampling in 2D can limit the interpretation of our findings and we thank the reviewer for bringing up this important point. The current version of the manuscript addresses the limitations of 2D sampling in the following paragraph at the end of the section titled “Cell types’ locations and relative proportions are consistent across morphologically normal gastruloids”:

“Our spatial transcriptomics is imaging-based, and the fact that we image transcripts in a single plane admits the possibility that, in any individual gastruloid, we may collect data from a different part of the gastruloid. We controlled for this to the extent possible within experimental limitations by imaging multiple gastruloids across several experiments and keeping our imaging parameters, particularly the instrument z-depth relative to the coverslip, nearly identical across experiments. The relatively wide distribution of mixing coefficients demonstrates that even within gastruloids with broadly similar morphologies and cell type proportions, the underlying organization of cell types can vary substantially.”

To further emphasize the specific issues raised we have amended this paragraph to the following:

“The seqFISH technique is imaging-based, and the fact that we image transcripts in a single plane admits the possibility that due to rotational differences, different parts of the gastruloid are imaged in each sample. We controlled for this to the extent possible within experimental limitations by imaging multiple gastruloids across several experiments and keeping our imaging parameters, particularly the instrument z-depth relative to the coverslip, nearly identical across experiments. We also note that previous analysis of 2D and 3D distances has indicated that in many cases, 2D distances (as we use in this work) are preferable for making comparisons between cells in a sample [Finn 2017].”

The final sentence is derived from the abstract of the paper referenced by the reviewer, which states “We conclude that 2D distances are preferred for comparative analyses between cells, but 3D distances are preferred when comparing to theoretical models in large samples of cells. In general, 2D distance measurements remain preferable for many applications of analysis of spatial genome organization.” We thank the reviewer for bringing this paper to our attention.

(4) Uncertainty in cell-type assignment is not incorporated into spatial metrics

Many spatial measurements depend directly on cell-type calls (exposure, triplets, mixing). However:

(a) Anterior cell types have higher entropy in their marker-based scores (Figure S1.1a).

We thank the reviewer for pointing out that several of the cell types in the anterior have high entropy — specifically cardiac mesoderm and paraxial mesoderm. However, we think there is nuance to this point; two of the other prominent anterior cell types (somite and endothelial) have low entropy scores overall, and that spinal cord/neural precursor cells, which are mostly posterior, have somewhat higher entropy; higher entropy scores are not exclusive to the anterior, nor is low entropy exclusive to the posterior. Inspired by the reviewer’s comments, we have re-analyzed our data to include this nuance (see response to point d) below.

(b) Uncertain labels inflate apparent "mixing" or "disorder," because misclassifications randomly create mixed neighbors and triplets.

We agree that uncertainty in labels could affect the interpretation of mixing. We appreciate these comments and the reviewer’s suggestions, and we have followed them in our response to point d) below.

(c) Posterior cell types have low entropy, so comparisons between anterior vs posterior mixing may partly reflect label uncertainty, not biology.

We thank the reviewer for bringing up this important caveat to our findings. We incorporated discussion of this in our text edits (see point d) below.

(d) The authors should incorporate confidence measures (e.g., probability-weighted neighbors, entropy filtering, bootstrapping) to confirm that patterns hold independently of classification noise.

We appreciate these suggestions and have chosen to use entropy filtering to assess whether the spatial organization we observe is highly sensitive to what values are considered ‘low’ entropy. We have updated the text (see below), and added Figure S2.2 to address the reviewer’s comments:

“The contrast between organized posterior clustering and disorganized anterior mixing matches expectations based on literature that shows that self-organization mechanisms in gastruloids in the anterior vs. posterior are differentially sensitive to culture conditions, with somitic patterning requiring external matrix support [1,4,8,24], distinguishing it from the seemingly more autonomous organization observed in posterior cell types.”

“One potential caveat to this finding is that differences in uncertainty in cell typing could be the primary driver of mixing and cell type interaction differences, both between the anterior and the posterior within an individual gastruloid, or overall between gastruloid. To control for this, we applied an entropy filter to our dataset. We filtered out cells that had entropy > 1.5 (see plot below for cutoff), which was chosen based on the distribution of entropy values for ‘none’ type cells, which effectively describe the upper limit of random transcript assignment (99.7% of ‘none’ type cells are removed with this filter, and about 50% of cardiac mesoderm cells and paraxial mesoderm cells, see Figure S2.2a). We first examined overall mixing; there was strong correlation between the per-gastruloid mixing indices before and after entropy filtering (Pearson r = 0.809, Figure S2.2b). Globally, mixing indices decreased with filtering, meaning that overall the cell types were more clustered. When we compared the absolute value of the change in mixing index pre and post-filtering to the proportion of each cell type, the only significant correlation was with cardiac mesoderm (Figure S2.2c). Exposure indices were overall quite similar after filtering, although the strength of somite-somite and somite-paraxial mesoderm interactions increased (Figure S2.2d).”

“We also examined how entropy filtering might affect the exposure index, given that the mixing index is calculated from the exposure index of across cell types. In general, the magnitude of the exposure index values increased when more uncertain cells were excluded, but the directionality and relative ordering was not affected. Although the magnitude of change in the posterior cells types was less than the anterior cell types, the cross-cell type exposure values, particularly between paraxial mesoderm/endothelium and somites, doubled. From these results we conclude that mixing in the posterior is driven mainly by NMP/presomitic mesoderm interactions, and is overall lower than mixing in the anterior, which is driven by rarer cell types like cardiac mesoderm, paraxial mesoderm, and endothelium, being interspersed within somite cells” (See Figure S2.2).

(e) This leads to reviewing the claim on endothelial heterogeneity, which strongly depend on spatial adjacency and gene exclusivity metrics.

We have extensively considered claims of endothelial cell heterogeneity, and these are discussed in detail in response to the reviewer’s next point. We have also copied them here for the reviewer’s convenience:

We re-assigned transcripts to nuclei at varying levels of nuclear dilation. If, as the reviewer suggests, the differences in gene expression are due to transcript mis-assignment, then reducing the nuclear dilation should reduce the entropy in cell type score. We re-analyzed the gastruloid shown in Figure 6, and assigned spots at various levels of nuclear dilation. Without dilation, all nuclei get 132 transcripts on average, and with dilation of 12 pixels (the maximum we tested) each got 181. We reassigned cell types and calculated the cell type score entropy. The results for endothelial and endodermal cells are shown in the Author response image 3:

While we do see a small increase in entropy score with dilation for endothelial cells, neither cell type comes anywhere near approaching the cell type entropy for non-typed cells at any dilation considered.

Additionally, we took several steps to verify that the cell states we found were a true reflection of endothelial cell biology. First, we pre-filtered genes on expression, so we only considered genes that were present in at least 50% of the cells in either group at a greater than 2 count per cell level. This was to ensure that the genes we detected were unique to that location spatially and that no effects were driven by expression from nearby tissue that could affect some cells more than others. Our list of differentially expressed genes changed — although some of the genes we had originally highlighted were still present, Gadd45g specifically was no longer present. The updated plot is shown in Figure 6.

If we do the same analysis with the nuclear dilation equal to 0, we find similar results, although many of the somite-associated genes are no longer present (likely due to the filtering, since overall counts are lower when the nuclear dilation is 0). See Author response image 4.

(5) Endothelial "spatially dependent" gene expression may reflect spillover rather than intrinsic state

The comparison between anterior-associated and posterior-associated endothelial nuclei suggests two transcriptional states. However, spatial adjacency confounds the interpretation:

(a) seqFISH assigns transcripts to nuclei in dense tissue; partial-volume effects can mix RNA from neighboring endodermal or somitic cells.

We thank the reviewer for their attention to detail and agree that a careful consideration of these points is important. We also note that given that some of our differentially expressed genes in endothelial cells are endoderm or somite genes, there indeed may be some transcript misassignment.

We re-assigned transcripts to nuclei at varying levels of nuclear dilation. If, as the reviewer suggests, the differences in gene expression are due to transcript misassignment, then reducing the nuclear dilation should reduce the entropy in cell type score. We re-analyzed the gastruloid shown in Figure 6, and assigned spots at various levels of nuclear dilation. Without dilation, all nuclei get 132 transcripts on average, and with dilation of 12 pixels (the maximum we tested) each got 181. We reassigned cell types and calculated the cell type score entropy. The results for endothelial and endodermal cells are shown in Author response image 3.

While we do see a small increase in entropy score with dilation for endothelial cells, neither cell type comes anywhere near approaching the cell type entropy for none-typed cells at any dilation considered.

(b) Endoderm and endothelium are closely intermixed (Figure S6.1d), and their gene expression co-localizes in KDE maps (Figure 5b-c).

We agree with the reviewer and thank them for their close reading of the manuscript. We re-assigned transcripts to nuclei at varying levels of nuclear dilation. If, as the reviewer suggests, the differences in gene expression are due to transcript mis-assignment, then reducing the nuclear dilation should reduce the entropy in cell type score. We re-analyzed the gastruloid shown in Figure 6, and assigned spots at various levels of nuclear dilation. Without dilation, all nuclei get 132 transcripts on average, and with dilation of 12 pixels (the maximum we tested) each got 181. We reassigned cell types and calculated the cell type score entropy. The results for endothelial and endodermal cells are shown in Author response image 3.

While we do see a small increase in entropy score with dilation for endothelial cells, neither cell type comes anywhere near approaching the cell type entropy for non-typed cells at any dilation considered.

(c) Without explicitly quantifying spillover, differential expression between these two endothelial subsets cannot be confidently attributed to cell-intrinsic differences.

(d) You may control for spatial proximity with any of the following:

- Include adjacency index as a covariate in DE models.

- Use scL-metric to test the mutual exclusivity of endothelial vs endoderm genes within the same nucleus.

- Apply local permutation nulls: shuffle transcripts within local windows and recompute DE.

- Restrict analysis to gastruloids that contain both endothelial subsets.

We thank the reviewer for bringing up this important point, which we were eager to address. We agree with the reviewer’s point that by only considering gastruloids that contain both subsets of endothelial cells is the correct way to do the analysis. We were already only considering this case (specifically the values calculated in Figure 6 and for 1 gastruloid pictured in Figure 6a). We added an n=1 label to revise Figure 6d to emphasize this point.

Additionally, we took several steps to verify that the cell states we found were a true reflection of endothelial cell biology. First, we pre-filtered genes on expression, so we only considered genes that were present in at least 50% of the cells in either group at a greater than 2 count per cell level. This was to ensure that the genes we detected were unique to that location spatially and that no effects were driven by expression from nearby tissue that could affect some cells more than others. Our list of differentially expressed genes changed — although some of the genes we had originally highlighted were still present, Gadd45g specifically was no longer present. The updated plot is shown in Figure 6.

If we do the same analysis with the nuclear dilation equal to 0, we find similar results, although some genes are no longer present (likely due to the filtering, since overall counts are lower when the nuclear dilation is 0) (See Author response image 4).

We have also included in the supplement larger images of some of the top differentially expressed genes, which more intuitively show the differential expression results (Endoderm enriched and Somite enriched).

Finally, we have referenced several previously-reported instances in the literature where distinct subsets of endothelial precursors with unique gene expression programs were identified. Although in these cases 1) the embryo models were different (in [Rossi 2021, Rossi 2022] gastruloids made with a different protocol and treated with factors designed to promote blood development, and in [Veenlveit 2020] trunk-like structures) and 2) the methods were different (IF and 10x single-cell sequencing) this at least establishes a precedent for the observation of multiple types of endothelial precursors. In the case of [Veenlveit 2020] the authors specifically note that one subset is associated with somites, and we have updated the text to reflect these new results:

“We observed that in 5 out of the 26 gastruloids, there was a large central patch of endoderm cells intermixed with endothelial precursors; these samples also had unique spatial L-score clustering of endothelial and endoderm genes (Figure 5b). An example of one such gastruloid is shown in Figure 6a. Migration to and association with the endoderm is also a hallmark of endothelial development [47,48], and we were curious whether there were differences between these cells and the cells we observed forming anterior, somite-associated clusters. When we computed the cell type exposure index for just this gastruloid, we found that, consistent with our visual observations, in this particular sample, endothelial and endoderm cells were much more frequently found next to one another than on average (Figure 6b,c). To determine whether these spatial and organizational differences reflected gene expression differences, we divided the gastruloid normal to the anterior-posterior axis to separate the endothelial cells into endoderm-associated and somite-associated and looked for differentially expressed genes between the two groups in this gastruloid. To ensure we were focused on genes that truly varied in expression in endothelial cells and were not merely a reflection of spillover from surrounding cells, we pre-filtered genes on expression, so only genes that were present in at least 50% of the cells in either group at a greater than 2 count per cell level were considered. The significantly differentially expressed genes after filtering are shown in Figure 6d. As an additional check on the degree to which transcript mis-assignment affected our analysis of gene expression in these cells in particular, we varied the nuclear dilation in this gastruloid specifically, and calculated cell type score entropy as a function of nuclear dilation (Figure S6.1a). Because cell type score entropy of a cell reflects the degree to which that cell specificity expresses genes associated with a single cell type, our expectation was that if spillover between endoderm and endothelial cells was a significant issue, then decreasing the nuclear dilation should greatly decrease the entropy scores for both groups. Although we saw a slight increase in the spread of the distribution as nuclear dilation increased, the median cell type entropy stayed extremely low for both groups (Figure S6.1a). From this analysis we conclude that the genes we identify as differentially expressed are not due to spillover from surrounding cells, but instead are due to spatially-dependent differences in endothelial cell biology.

The genes with the highest fold-change in expression in endoderm-associated endothelial genes are shown on the left-hand side of Figure 6d. Two are endothelial genes: Pecam1 and Cdh5, both of which are associated with angiogenesis. Pecam1 also clustered uniquely in our L-metric analysis (Figure 4c), suggesting this differential expression is conserved across gastruloids. Spatial expression of these genes is shown in the top row of Figure 6e (larger version in Figure S6.1b). Notch1 is more expressed in endoderm-associated endothelial cells, and this could reflect an increase in Notch signaling in the posterior of the gastruloid. [Chan et al 2017] demonstrated that Notch signalling can be sensitive to shear stress, raising the possibility that the differences in cell state we observe may be driven by differences in mechanical forces in the anterior and posterior. Although most endothelial cells are thought to be of mesodermal origin, some evidence suggests that, in the organogenesis of specific tissues like the liver, the endoderm can give rise to endothelial cells [49]. Furthermore, in [Rossi 2022] the authors show that in a gastruloid-like model specifically designed to model blood development, there is strong spatial adjacency between endothelial and endoderm cells. They hypothesize that these may be a subset of endothelial cells, specifically hemogenic endothelial cells (which have the potential to become blood progenitors). Our data demonstrate a molecularly driven organization distinct from the clustering we observed in the anterior and suggest that multiple mechanisms of endothelial specification could be modeled in gastruloids, even simultaneously within the same structure, although further characterization is needed to determine exactly what processes these unique endodermal/endothelial structures model.

Several other endothelial genes are instead differentially expressed in somite-associated endothelial cells: Nrp2, Tek, Apoe, and Cldn5. Although these genes have less obvious functional distinctions than the endoderm-associated genes, Nrp2 enables semaphorin receptor activity, including nervous system development and ventral trunk neural crest cell migration and Tek negatively regulates endothelial cell apoptotic process and response to retinoic acid (RA), which is known to be higher in the gastruloid anterior. Furthermore, a specialized population of endothelial precursors associated with somites was also observed in trunk-like structures, which are more organized organoids than gastruloids [Veenvliet et al. 2020].

Although endothelial cells have consistently been observed in single-cell measurements of gastruloids, their relative rarity has precluded in-depth analysis of subtypes or inference of spatial location. Our results strongly suggest that endothelial precursor formation, migration, and organization may all be modeled in 3D gastruloids, even without treatment with additional factors as in [Rossi 2021, 2022]; recent advances in 2D gastruloids have allowed modeling of cardiac and hepatic vascularization [45], and our data suggest that 3D gastruloids may similarly be adapted to model more specific aspects of hematopoiesis and vascularization. Early specification from a pool of mesodermal precursors is a hallmark of the endothelial lineage [47]; given the consistency with which we observe endothelial precursors, we speculate that this behaviour is recapitulated in gastruloids, but further epigenetic measurements are required to validate this hypothesis” (See Revised Figure 6)

(6) Interpretation of gene-program modules may be overstated

Claims that the L-metric reveals "novel gene programs" should be softened:

(a) The seqFISH panel is an approx. 200-gene marker-enriched panel, already biased toward known cell-type markers.

This is true and we appreciate that this came through in the text since it’s important for the reader to understand the approach we took in this study.

(b) Strong blocks in Figure 4a and S4.1a may reflect panel design rather than newly discovered programs.

We agree with the reviewer that the panel design was not sufficiently highlighted, so we have made the following changes to the text to emphasize which patterns would be expected due to the genes we are probing for, and which findings were surprising given the known functional role of the gene.

At the end of the section titled ‘The L-metric captures the spatial distribution of gene expression despite being calculated without spatial information’:

“These analyses demonstrate that information contained within the hierarchical relationships between genes, determined by scL-score can reveal novel information about cell states within cell types, although we acknowledge that since cell type is determined by a limited panel of marker genes, results should be further functionally verified. scL-score analysis can also identify distinct spatial locations of cells in this cell state, all without explicit encoding of spatial information, but rather quantifying and clustering the degree to which genes are mutually exclusively expressed with one another.”

At the end of the section titled ‘Clustering scL-metric vectors clearly resolves cell types and reveals novel genetic interactions’:

“Finally, although the strong blocks we find in the heatmaps in Figures 4a and S4.1a largely reflect cell types, as is consistent with our panel design, we discovered some novel functions of genes in the panel through their location in the scL-score tree: although Tgfβ was initially included in our panel to generally detect inflammatory and growth signaling, clustering by expression patterns revealed its unique association with endothelial precursors.”

To further address the concern that the generality of clustering is due to gene selection, we performed random gene drop-out and assessed how well cell types clustered as a function of the number of genes removed:

“Given the amount of spatial and state information that was encoded in the scL-score heatmap for a subset of our gene panel, we expanded our analyses to all genes, hoping to discover new genetic interactions or refine existing ones. We first calculated the scL-score for all genes in all gastruloids, then averaged across gastruloids and clustered the resulting interaction vectors (see Methods for details). The heatmap is shown in Figure 4a (heatmap including cell cycle genes is shown in Figure S4.1a). We noted that just as when we clustered genes associated with NMPs and their direct descendants, genes associated with cell types tended to cluster together. Specifically, NMP, spinal cord, endoderm, and endothelial genes clustered very strongly together, while presomitic mesoderm genes again were split into two groups, one of which was more closely associated with genes involved in early somitogenesis. We quantified how well cell type specific genes clustered compared to a random null by first calculating the dispersion of cell types within the tree topology using cophenetic distance (see Methods), and then permuting the leaves of the tree to create a null distribution of the dispersion expected by random. The results produced by hierarchical clustering on scL-score vectors were significantly (p=0.0001) more clustered than would be expected by chance (Figure S4.2a,b). To assess cluster stability, we randomly selected subsets of the panel and repeated the clustering. Regardless of panel size, the tree produced by clustering on scL-score vectors was always significantly less dispersed than permuted nulls (Figure S4.2c). Although our method of calculating dispersion can only be compared between trees clustered on the same gene set, we noted that as we increased the number of genes, the gap between the dispersion of the real tree and the dispersion of the permuted trees increased (Figure S4.2d), indicating that, as would be expected, better clustering was achieved when more genes were considered.”

Finally, we performed scL-score analysis on an unbiased, scRNA-seq dataset without any panel selection, and were able to show similar groupings of cell type markers:

“scL-score analysis reveals cell type groupings and new transcription factor associations in a single-cell RNA-seq dataset

To test the generality of scL-score analysis, we analyzed a previously published dataset from [van den Brink 2020] where individual gastruloids (at the same stage as those used in this study) were pooled and subjected to single-cell RNA-seq analysis. After filtering for cell quality and common gene detection, we calculated the scL-score values using a cell-by-gene table of 14304 cells x 19075 genes.

To first test whether we could reproduce the results from this study, we performed hierarchical clustering on the scL-score difference vector (as previously described) on the set of 207 well-detected genes that were also present in our seqFISH gene panel. The resulting tree showed clustered cell types, similar to the tree produced with the expression data in this study (compare Figure S4.5a to Figure S4.4d). The cell types were clustered significantly more than expected by chance (Figure S4.5b).

Then, to test whether scL-score analysis would be effective for analyzing the entire dataset, we performed hierarchical clustering on all 19075 genes. We then examined the resulting heatmap (Figure S4.5c) for clusters of interest. We observed a cluster enriched for endothelial genes (Figure S4.5d), which contained some genes in our panel but many others which were not; this finding demonstrates that the clustering in Figure 4a is not solely due to the selection of genes in our seqFISH panel. We also observed a large cluster that contained genes associated with pluripotency or primordial germ cell fate (Figure S4.5e). Although some of the genes in this cluster were in our seqFISH panel, when we performed scL-score analysis we did not see them cluster with each other or with any other cell type genes. This lack of clustering implies that in our dataset cells that co-express these genes may be rare or too poorly detected to cluster strongly; however, the same analysis performed with more cells and genes showed association. This result demonstrates that clustering scL-score difference vectors can identify known cell-type-associated genes, even within transcriptome-scale data. Finally, we also found a small cluster showing strong co-expression of the transcription factor Gata4, a crucial regulator of the development of visceral and parietal endoderm, and two other genes: a predicted gene of unknown function (Gm43715) and Troponin C (Tnnc1) (Figure S4.5f). Intriguingly, Gata4 has been implicated in heart development (albeit in an indirect manner) [Watt 2004], and troponin C is important for cardiac muscle cell contraction and has been implicated in cardiomyopathy, although at a much later stage of development than that modeled by gastruloids [Li 2015].

Together, these results demonstrate that scL-score analysis is reproducible across datasets, even when different numbers of genes are compared. It effectively clusters genes associated with cell types, and can reveal developmental transitions. Moreover, increasing the number of cells and genes can reveal new clusters, some of which may predict novel regulatory interactions or spatial co-occurrence not previously observed.”

(c) Cluster robustness is not assessed (bootstrap, stability).

We appreciate the reviewer’s suggestion that cluster robustness should be assessed. To address this point, we performed a clustering-stability analysis on the common 202-gene panel that included cell cycle genes by asking to what extent hierarchical clustering could recapitulate cell type-based groupings of genes as the number of genes used for clustering was varied across progressively larger, randomly sampled panel subsets. For each resulting tree, we averaged cell types’ dispersion of genes across the tree using the framework described in our response to suggestion 6m and compared the observed value to a permutation-based distribution generated on the same tree.

This analysis showed that the observed cell-type dispersion remained consistently lower than the corresponding permutation distribution across all subset sizes examined. In other words, genes assigned to the same annotated cell type remained closer together in the dendrogram than expected by chance even when clustering was performed on reduced random subsets of the panel. We also observed that dispersion values increased as larger gene subsets were included, which is expected as the clustering problem becomes more complex with increasing panel size; however, the separation between the permuted distribution of average cell type dispersion and the observed dispersion value increased as the panel subset size increased. Taken together, these results indicate that the cell type-resolved organization captured by the scL-score derived hierarchy is not dependent on one particular subset of genes, but is instead a stable property of the broader gene panel. New Figure S4.2 addressing cluster stability:

We have added the following explanation in the text:

“Given the amount of spatial and state information that was encoded in the scL-score heatmap for a subset of our gene panel, we expanded our analyses to all genes, hoping to discover new genetic interactions or refine existing ones. We first calculated the scL-score for all genes in all gastruloids, then averaged across gastruloids and clustered the resulting interaction vectors (see Methods for details). The heatmap is shown in Figure 4a (heatmap including cell cycle genes is shown in Figure S4.1a). We noted that just as when we clustered genes associated with NMPs and their direct descendants, genes associated with cell types tended to cluster together. Specifically, NMP, spinal cord, endoderm, and endothelial genes clustered very strongly together, while presomitic mesoderm genes again were split into two groups, one of which was more closely associated with genes involved in early somitogenesis. We quantified how well cell type specific genes clustered compared to a random null by first calculating the dispersion of cell types within the tree topology using cophenetic distance (see Methods), and then permuting the leaves of the tree to create a null distribution of the dispersion expected by random. The results produced by hierarchical clustering on scL-score vectors were significantly (p=0.0001) more clustered than would be expected by chance (Figure S4.2a,b). To assess cluster stability, we randomly selected subsets of the panel and repeated the clustering. Regardless of panel size, the tree produced by clustering on scL-score vectors was always significantly less dispersed than permuted nulls (Figure S4.2c). Although our method of calculating dispersion can only be compared between trees clustered on the same gene set, we noted that as we increased the number of genes, the gap between the dispersion of the real tree and the dispersion of the permuted trees increased (Figure S4.2d), indicating that, as would be expected, better clustering was achieved when more genes were considered.”

(d) Agreement with cNMF (claimed in text) is not quantified (ARI, Jaccard, hypergeometric overlap).

We appreciate this suggestion offered by the reviewer as quantifying the agreement between cNMF-derived gene programs and our scL-score-determined clusters will allow readers to more rigorously assess the extent to which these two approaches recover similar groupings of genes. To address this, we compared the top 24 genes of K=7 clusters identified using cNMF to 7 clusters (average 24 genes) obtained from scL-score-based hierarchical clustering at the appropriate cophenetic distance threshold (as originally depicted in Figure S4.2). We computed the pairwise overlap between every scL cluster and every cNMF cluster and quantified each comparison using the Jaccard Index and Adjusted Rand Index. For each scL cluster, we plotted only the maximum value observed across its 7 possible cNMF cluster comparisons, thereby capturing the strongest correspondence between each scL cluster and the cNMF-defined programs for a given metric.

To establish a baseline for these overlap measures, we designed a reference simulation by preserving the same cNMF clusters while defining a “permuted” set of scL clusters obtained by randomly assigning genes to clusters of the same number (7 clusters) and set of sizes (average 24 genes) as the scL clusters. As above, for each permuted scL cluster and each metric, we retained only the maximum overlap value across 7 possible cNMF cluster comparisons. We note that under this framework, the same cNMF cluster can serve as the highest-overlap comparison for more than one scL cluster.

The following plots summarize the results of applying this approach. Higher values (closer to +1) for the Jaccard Index and Adjusted Rand Index correspond to greater overlap between observed or permuted scL clusters and cNMF clusters. Across both metrics, the observed scL clusters consistently exhibited substantially higher overlap with cNMF clusters compared to permuted scL clusters. For the Jaccard Index, the observed clusters showed markedly elevated values relative to the narrow distribution centered near 0 obtained under permutation, demonstrating that gene overlap between scL clusters and cNMF programs is greater than expected by chance. This similarly holds when gene overlap is assessed using the Adjusted Rand Index. Together, these results quantify how the scL-score can hierarchically derive clusters of genes that recapitulate major gene programs identified by cNMF to an extent beyond that expected under random clustering (See Revised Figure S4.2 (now S4.4)).

We have updated the text to reflect these quantitative comparisons:

“To validate the clustering produced by the scL-score, we compared our results to a state-of-the-art method for identifying gene programs in an unbiased fashion from single-cell data: consensus non-negative matrix factorization (cNMF) [39]. We pooled nuclei from all individual gastruloids and ran cNMF. We found that many of the resulting clusters (Figure S4.4a,b) corresponded to the clusters identified when the scL-score tree was truncated to produce exactly the same number of clusters (Figure S4.4c). The similarities were even greater when the scL-score clusters were hand-selected based on visual inspection of the tree and density of marker genes (Figure S4.4d). To quantify the overlap between clusters, we calculated both the Jaccard Index and the Adjusted Rand Index (ARI) between each scL-score cluster (Figure S4.4c) and the most similar cNMF cluster. These distributions are shown in Figure S4.4e (blue). We compared to a bootstrapped null where we permuted the genes found in the scL-score clusters, and found that permuted clusters were far less similar to the cNMF clusters than those derived from the real scL-score tree (Figure S4.4e). From these observations, we conclude that the two methods are capable of producing similar results at a high-level, but are different in their application. Individual cells receive component scores for cNMF gene programs, yielding more per-cell information, while the tree produced by L-score clustering reveals hierarchical information about gene programs, which quantifies their similarity in expression on a more global scale.”

(e) Testing the scL-metric on larger, unbiased scRNA-seq datasets would help demonstrate generality.

We agree with the reviewer that this would demonstrate generality, so we applied scL-score analysis to the scRNA-seq dataset from van den Brink 2020 — several gastruloids at the same stage of development as those used in this paper were pooled and sequenced. We added a figure, new Figure S4.5 with the results of this analysis, and a new section in the paper describing them:

“scL-score analysis reveals cell type groupings and new transcription factor associations in a single-cell RNA-seq dataset

To test the generality of scL-score analysis, we analyzed a previously-published dataset from [van den Brink 2020] where individual gastruloids (at the same stage as those used in this study) were pooled and subjected to single-cell RNA-seq analysis. After filtering for cell quality and common gene detection, we calculated the scL-score values using a cell-by-gene table of 14304 cells x 19075 genes.

To first test whether we could reproduce the results from this study, we performed hierarchical clustering on the scL-score difference vector (as previously described) on the set of 207 well-detected genes that were also present in our seqFISH gene panel. The resulting tree showed clustered cell types, similar to the tree produced with the expression data in this study (compare Figure S4.5a to Figure S4.4d). The cell types were clustered significantly more than expected by chance (Figure S4.5b).

Then, to test whether scL-score analysis would be effective for analyzing the entire dataset, we performed hierarchical clustering on all 19075 genes. We then examined the resulting heatmap (Figure S4.5c) for clusters of interest. We observed a cluster enriched for endothelial genes (Figure S4.5d), which contained some genes in our panel but many others which were not; this finding demonstrates that the clustering in Figure 4a is not solely due to the selection of genes in our seqFISH panel. We also observed a large cluster that contained genes associated with pluripotency or primordial germ cell fate (Figure S4.5e). Although some of the genes in this cluster were in our seqFISH panel, when we performed scL-score analysis we did not see them cluster with each other or with any other cell type genes. This lack of clustering implies that in our dataset cells that co-express these genes may be rare or too poorly detected to cluster strongly; however, the same analysis performed with more cells and genes showed association. This result demonstrates that clustering scL-score difference vectors can identify known cell-type-associated genes, even within transcriptome-scale data. Finally, we also found a small cluster showing strong co-expression of the transcription factor Gata4, a crucial regulator of the development of visceral and parietal endoderm, and two other genes: a predicted gene of unknown function (Gm43715) and Troponin C (Tnnc1) (Figure S4.5f). Intriguingly, Gata4 has been implicated in heart development (albeit in an indirect manner) [Watt 2004], and troponin C is important for cardiac muscle cell contraction and has been implicated in cardiomyopathy, although at a much later stage of development than that modeled by gastruloids [Li 2015].

Together, these results demonstrate that scL-score analysis is reproducible across datasets, even when different numbers of genes are compared. It effectively clusters genes associated with cell types, and can reveal developmental transitions. Moreover, increasing the number of cells and genes can reveal new clusters, some of which may predict novel regulatory interactions or spatial co-occurrence not previously observed.”

(7) Minor Comments

(a) We suggest including representative raw seqFISH images. The manuscript does not show raw images, which makes it difficult to evaluate the quality of the underlying data that all spatial analyses depend on. A figure showing raw fluorescence channels, detected spots, and nuclei segmentation masks for at least one anterior region, one posterior region, and one dense interface (e.g., endoderm-endothelial) would allow readers to assess spot intensity and background levels, signal-to-noise ratio, segmentation accuracy, and potential over/under-segmentation, channel cross-talk, and transcript crowding or dropouts in dense tissues. A small panel of raw images would improve the transferability to the spatial metrics.

This is an excellent suggestion and we have included examples of the raw images for a representative gene for all hybridizations, the spots as determined by the spot-finding algorithm distributed with the seqFISH instrument, deconvolved spots, and nuclear segmentation. These images can be found in Revised Figure S1.1d.

(b) The Introduction mentions 3D organization, which can confuse readers into thinking the seqFISH dataset is volumetric. The data shown and analyzed come from a single 2D plane per gastruloid, not from full 3D z-stacks. Since all spatial metrics rely on true adjacency, the manuscript should explicitly state early on that the dataset is 2D and briefly justify why a single plane is sufficient for the analyses.

We appreciate the reviewer bringing up this point and we have removed references to 3D so as not to confuse readers.

We also have included an extensive discussion of the 2D nature of the data. The current version of the manuscript addresses the limitations of 2D sampling in the following paragraph at the end of the section titled “Cell types’ locations and relative proportions are consistent across morphologically normal gastruloids”:

Our spatial transcriptomics is imaging-based, and the fact that we image transcripts in a single plane admits the possibility that, in any individual gastruloid, we may collect data from a different part of the gastruloid. We controlled for this to the extent possible within experimental limitations by imaging multiple gastruloids across several experiments and keeping our imaging parameters, particularly the instrument z-depth relative to the coverslip, nearly identical across experiments. The relatively wide distribution of mixing coefficients demonstrates that even within gastruloids with broadly similar morphologies and cell type proportions, the underlying organization of cell types can vary substantially.

To further emphasize the specific issues raised we have amended this paragraph to the following:

“The seqFISH technique is imaging-based, and the fact that we image transcripts in a single plane admits the possibility that due to rotational differences, different parts of the gastruloid are imaged in each sample. We controlled for this to the extent possible within experimental limitations by imaging multiple gastruloids across several experiments and keeping our imaging parameters, particularly the instrument z-depth relative to the coverslip, nearly identical across experiments. We also note that previous analysis of 2D and 3D distances has indicated that in many cases, 2D distances (as we use in this work) are preferable for making comparisons between cells in a sample [Finn 2018].”

The final sentence is derived from the abstract of the paper referenced by the reviewer, which states “We conclude that 2D distances are preferred for comparative analyses between cells, but 3D distances are preferred when comparing to theoretical models in large samples of cells. In general, 2D distance measurements remain preferable for many applications of analysis of spatial genome organization.” We thank the reviewer for bringing this paper to our attention.

(c) Results, first paragraph: "good agreement" along the AP axis should be quantified or defined.

In the first paragraph we compare the peak in gene expression along the (length-normalized) AP axis of each gene with a similar but orthogonally measured dataset from another group (van den Brink 2020). The text specifically reads:

“When we compared how gene expression varies along the AP axis, we saw good agreement at a coarse-grained level with a previous study that sectioned gastruloids along the axis and analyzed gene expression in each section.”

We have revised Figure S1.2 with a summary plot showing the distribution of correlation coefficients for all genes and for the Hox genes in our panel (which are known to be expressed sequentially along the AP axis).

We have updated the text as follows:

“To assess the quality of our data, we first assigned an AP axis to each gastruloid using the expression of T, a canonical marker for the posterior (Figure 1a). When we compared how gene expression varied along the AP axis, we saw good agreement at a coarse-grained level with a previous study that sectioned gastruloids along the axis and analyzed gene expression in each section [2] (Figure S1.2a). The colinearity of the peak expression of Hox genes in our panel was also consistent with this dataset, with a median Pearson correlation of 0.695 (compared to 0.663 for all genes (Figure S1.2b).”

(d) Clarify what "greater cell type distinction" means when using marker panels, and what metric demonstrates improvement?

By “greater cell type distinction” we meant that when using traditional clustering methods we were not able to individually resolve some cell types: there was a mixed differentiation front and presomitic mesoderm cluster, and NMP and spinal cord cells were also clustered together. Cluster labeling was performed by considering which genes showed up as being differentially expressed in each cluster using the same associations as were used to perform the cell type scoring with the marker gene panel. While we acknowledge that these results are subjective to clustering parameters, this is a general problem with clustering and not specific to this study. We have added additional clarification in the text and removed the phrase “greater cell type resolution” since it wasn’t clear what we were comparing to:

“To profile the spatial organization of individual gastruloids, we assigned a cell type to each nucleus using a cell type scoring method with known marker genes. We compared these results to those obtained with unsupervised clustering. We found that although clustering did produce clusters, they were not strongly separated and we were not able to individually resolve some cell types we expected to find: specifically, there was a mixed differentiation front and presomitic mesoderm cluster, and NMP and spinal cord cells were also clustered together (Figure S3.6a). Therefore, we proceeded with the scoring-based method; see Methods for additional details. A representative gallery of typed gastruloids is shown in Figure 1b; the full dataset is in Figure S1.3. Because each cell received a cell type score for each type, we could use the entropy of the cell type score probability distribution to assess confidence of our cell type assignment: a cell that received a similar score for multiple cell types would have a high entropy distribution, while one which scored highly for one type and low for the rest would have low entropy. On average, the cell type entropies for most cells in a given cell type were low, with the exception of paraxial mesoderm and cardiac mesoderm cells, which had intermediate values (see additional discussion below). The generally low entropies indicate that most of our cell type assignments were high-confidence (Figure S1.4a).”

(e) The variation in "none-typed" cells across gastruloids should be expanded and shown quantitatively.

We agree that this is important and we have added the proportion of none-typed cells to Revised Figure S1.4c.

(f) Figure S1.1b: explain the criteria used to decide which tissues "varied significantly." Showing the proportion of "None" per gastruloid would help.

We thank the reviewer for pointing this out. We agree that this is important and we have added the proportion of none-typed cells to Revised Figure S1.4c.

To justify the use of the phrase ‘varied significantly’ we have calculated the degree to which cell type pairs significantly co-vary (adjusted p value < 0.05) in their proportions and added the plot to Supplemental Figure 1.4, highlighting the significantly varying pairs.

We address said variation in the text:

“We sought to quantify variability in cell type composition between the 26 morphologically normal gastruloids profiled. Previous single-cell datasets relied on pooling multiple gastruloids, thus obscuring the degree to which the overall cell type distribution was reflected in each individual gastruloid. However, recent single-cell measurements of individual gastruloids have suggested substantial gastruloid-to-gastruloid variation in cell type proportions [13]. Figure 1c shows distributions of cell type proportions across samples, and Figure 1d shows the coefficient of variation of these proportions. Individual gastruloid cell type distributions, including the proportion of cells that had insufficient reads to be confidently assigned a type, are shown in Figures S1.4b and c. We found that cardiac mesoderm, endoderm, and spinal cord cells had the greatest coefficient of variation in proportion between gastruloids (Figure 1d). To calculate statistical significance, we first performed a centered log-ratio (CLR) transform on the proportions, then looked for covariation between cell types across gastruloids. We found there was a statistically significant inverse correlation between the proportion of endoderm and NMP, presomitic mesoderm, and differentiation front (Figure S1.4d). We did not observe gastruloids that were as strongly neurally-biased as those reported in [13], but we did see some gastruloids with a relatively high proportion of spinal cord precursor cells (Figure S1.32a ii., xv., b vii.), and overall the proportion of spinal cord had a negative covariation with the mesodermally-derived cell types, consistent with the anticorrelation also reported in [13] (Figure S1.41dc).”

“The proportion of somite cells was significantly positively correlated with the proportion of presomitic mesoderm cells (covariation = 0.63, Figure S1.41dc).”

(g) In Figure 1c, showing the variability of "None" cells is important.

We made this adjustment and added the plot to Figure S1.3c

(h) For Figure 1e, consider normalizing to T-expression proportion, not only AP length. This could clarify multimodality in the endoderm and spinal cord.

We thank the reviewer for this suggestion to normalize by molecular as well as physical features. We agree this could be a useful projection of the data, since expression of T is used in many contexts to define the posterior of gastruloids.

We incorporated T expression into the length normalization in the following way: we generated a cumulative distribution of all T spots along the AP axis, and when this value exceeded a threshold (specifically 90% of all spots) we defined this as the midpoint of the gastruloid, and linearly normalized space before and after it from 0-0.5 and 0.5-1 respectively. We then re-projected nuclei for each gastruloid onto this new coordinate system, and visualized in the same way as the main text figure (See Figure 1f).

To assess whether this increased or decreased variability in position, we calculated how much the location of the peak of each cell type in each gastruloid differed from the peak position of that cell type in all samples pooled together. We compared this difference to a bootstrapped null drawn from the pooled distribution. Interestingly, we found that nearly all cell types had statistically significant variation (meaning the average distance from the mean fell outside the bootstrapped distribution in the positive direction), however the effect size for most cell types was very small.

Notably, as the reviewer suggested, normalizing to T expression decreased the effect size of this variability in all cases but one, and particularly decreased spinal cord variability (despite not being a marker for spinal cord):

We interpret these findings in the following way — that most of the variation observed in cell type arrangement along the AP axis is due to morphological and molecular variability (likely due to stochasticity in initial cell number and differences in developmental timing between gastruloids), and that once these factors are taken into account, for most cell types the effect size of variability is very small. However for some cell types, notably the location of the differentiation front, the position varies among gastruloids. We hypothesize that this may be due to the rhythmic nature of somite development. We have updated the text to reflect these changes.

“Given that the proportions of cell types within each gastruloid were fairly consistent, to what degree did their spatial organization vary? We first projected each cell’s expression onto the AP axis and looked at the distribution of AP axis locations across gastruloids (Figure 1f, left-hand side). The most posterior cell types (NMP and presomitic mesoderm) showed wide distributions from 0-30% of the axis. Centred around 30%, spinal cord precursors and endoderm had distinctive peaks (clusters), the exact location of which varied between gastruloids. Using a threshold of T expression to define the midpoint of each gastruloid and uniformly length-normalizing each half caused these cell types to collapse into a single peak (Figure 1f, right-hand side). The differentiation front was similarly located in one peak (between 30-40% of the AP axis), and the variation in the location of this peak decreased with T-expression normalization. To quantify this variability, we bootstrapped a null distribution of cell type locations by pooling each cell type together across samples and using the resulting distribution to define a reference mean. When we compared how much peak variation there was among samples randomly drawn from this null to our observed data, we found that although nearly every cell type (excepting paraxial mesoderm) had more variability than expected by chance, the effect size of this variation was small (Figure S1.5a). Normalizing location to T-expression decreased variability in most cell types, most notably for differentiation front and spinal cord (Figure S1.5b).

We also asked how the order of cell types along the AP axis varied between gastruloids. When we ranked the peaks shown in Figure 1f per gastruloid, we found that Kendall’s W, an overall measure of rank coherence across independent samples that spans from 0 (no agreement) to 1 (complete agreement), was 0.834 (Figure S1.5c). We found that the cell types most likely to swap rank order were spinal cord and endoderm, and paraxial mesoderm and endothelium (Figure S1.5d).”

(i) Clarify "mixing coefficient values range from 0.29-0.58" (incomplete sentence). Section "Cell types' arrangement is consistent across gastruloids, but varies by type" second paragraph.

We appreciate that there was some confusion here and we thank the reviewer for pointing it out. We have updated the text with the new values and a more specific interpretation to aid the reader and address the reviewer’s comment:

“The mixing index values range from -0.50 to -0.22 (Figure 2a). While all values are negative, the range was large. This observation led us to conclude that while in all the gastruloids profiled cell types tended to cluster together, there was variation between gastruloids in the degree of coherent clustering between types.”

(j) The paragraph claiming organization consistency may be too strong, given the lack of statistical validation. "Overall, these data speak to the consistency of gastruloid organization...".

We agree that quantification of variability, which we only assessed qualitatively, would help readers better evaluate claims of organizational consistency, and we appreciate the reviewer pointing this out.

We have taken several steps to add quantification, including:

(1) Calculating the coefficient of variation for the proportion of each cell type

(2) Quantifying co-variation and assessing statistical significance

(3) Quantifying the variation in AP-axis location, including comparison to a bootstrapped null, significance testing, and a new form of normalization which decreases some of the variability.

(4) Quantifying order along the AP axis and calculating Kendall’s W to quantify how concordant this ordering is across gastruloids

(5) Overhauling the methods used to quantify local patterning and mixing into one unified metric

(6) Calculating significance for variation in exposure index across samples

To address the question of whether claims of organization consistency are appropriate, we have added the following summary paragraph at the end of the results for the first two figures, and have taken the reviewer’s suggestion of only discussing statistically significant or quantified results in listing both consistent and variable features. Now rather than making an argument about whether gastruloids are consistent or not, we merely provide the readers with our findings:

“Variation in cell type abundance and organization is structured and concentrated in specific cell types

We have demonstrated that some aspects of gastruloid composition and spatial organization are consistent across gastruloids, while others are more variable. Consistent features include proportions for NMP, presomitic mesoderm, somite, and paraxial mesoderm, whose coefficients of variation were lower than other cell types (Figure 1d). Organizationally, all cell types across gastruloids are more physically clustered than random (Figure 2a), and the order in which cell types are found along the AP axis has statistically significant high agreement between gastruloids as measured by Kendall’s W (Figure S1.5c). At the local neighbourhood scale, we found that most cell type interactions were conserved across gastruloids (Figure S2.1c). At the local scale, across individual gastruloids, we found many motifs of three cells that were statistically enriched over random, suggesting a conserved local order (Figure 2c). While the normalized distance along the AP-axis of all cell types significantly varied compared to a bootstrapped null (Figure S1.5a), the effect size was small, and decreased in almost all cases when normalized to gene expression (of T) in addition to morphology (Figure S1.5b).

However, there were also variable features. The proportion of cardiac mesoderm, endoderm, and spinal cord had the highest coefficient of variation between gastruloids (Figure 1d). Because proportions must sum to one, a change in the proportion of one cell type is necessarily linked to changes in others; we performed centred log transformation and looked for statistically significant covariation. Of all possible pairings, the following proportions had a significantly negative correlation across samples: endoderm/differentiation front, NMP/endoderm, presomitic mesoderm/endoderm, none/endothelial, and spinal cord/endothelium. This result shows that the proportions of these cell types predictably co-vary between samples, potentially suggesting some kind of biological trade-off in cell type specification or organization (Figure S1.4d).

Across gastruloids, intra-cell type interactions (degree of clustering) of spinal cord, endoderm, and differentiation front vary (Figure S2.1b). This variation suggests that these cell types may be patterned differently between gastruloids. For example, the local motif of 3 endoderm cells found next to one another was statistically enriched within some but not all individual gastruloids, and by definition is completely absent from gastruloids lacking endoderm (Figure 2c). We interpret this contrast to mean that when endoderm is found in a gastruloid, it is consistently patterned at a local level, but may vary more at a global level. This interpretation is concordant with the findings from [Farag 2024], which demonstrates several distinct classes of endoderm organization in gastruloids.

To summarize, while changes in the amount of individual cell types can vary, these changes are in most cases explained by variations in morphology and molecular characteristics (such as anterior: posterior ratio and the expression of morphogens like T). For patterning, we found that, in most cases, global patterns were conserved, but there were small variations in local patterning that may lead to variable meso-scale organization of specific cell types, particularly those found in the middle of the anterior-posterior axis.”

(k) Figure 3c: gene set sizes differ substantially; proportion-based normalization may be more appropriate.

We thank the reviewer for carefully noting the gene set differences. While the NMP-only and spinal cord-only gene sets each have 10 genes, PSM-only and NMP+PSM have 6 genes, and NMP+spinal cord has 5 genes. Given the relatively small N, we feel that normalization would likely introduce a layer of abstraction that would be more confusing for the reader, especially given the qualitative nature of the claims made about the shape of the plots in question. However, we agree this point is important, so we have now indicated the size of the gene sets on the plot (revised Figure 3b, previously c).

(l) The purpose of the first two graphs in Figure 3c is unclear.

We thank the reviewer for pointing out that this is unclear. The purpose of this visualization is to demonstrate correlation between genes shared between cell types and genes exclusive to only one of the cell types. The text reads:

“Figure 3b shows the total expression of each gene group versus the NMP genes for all cells in all gastruloids that we typed as NMP, presomitic mesoderm, or spinal cord. As expected, there was a clear correlation between the mixed categories and NMP genes, supporting the notion of a continuous differentiation process.”

We have added a correlation line to revised Figure 3b (former Figure 3c) to emphasize this point.

(m) Clustering of all genes: quantify whether clusters align with cell types.

We appreciate this suggestion offered by the reviewer as this analysis will allow readers to quantitatively assess the overlap between cell type-based groupings of genes and our scL-score-determined clusters for this dataset. To address this, we have used a bootstrapping approach to evaluate how well our scL-score-based hierarchies capture cell type-based groupings. Specifically, following hierarchical clustering of genes using the scL-score, for each cell type, we computed the average of the minimum cophenetic distance between each pair of genes associated with that cell type. After obtaining each cell type’s average, we took the mean of these averages which we refer to as the cell type dispersion for that tree. We note that the absolute value depends on the topology of the tree. To establish a reference for this measure, we bootstrapped a null distribution by preserving the same tree topology and randomly assigning genes to leaves and calculating the resulting dispersion. We performed this 10,000 times to create a reference null distribution. We compared the true dispersion value to this null, and computed a bootstrapped p-value. In both cases (with and without cell cycle genes), the observed average cell type dispersion is less than that for all permutations (p-value=0.0001, bootstrapping), indicating that genes associated with the same cell type were significantly more likely to cluster together on the observed scL-score hierarchy than would be expected under random clustering.

We have updated the text to reflect these quantitative comparisons:

“Given the amount of spatial and state information that was encoded in the scL-score heatmap for a subset of our gene panel, we expanded our analyses to all genes, hoping to discover new genetic interactions or refine existing ones. We first calculated the scL-score for all genes in all gastruloids, then averaged across gastruloids and clustered the resulting interaction vectors (see Methods for details). The heatmap is shown in Figure 4a (heatmap including cell cycle genes is shown in Figure S4.1a). We noted that just as when we clustered genes associated with NMPs and their direct descendants, genes associated with cell types tended to cluster together. Specifically, NMP, spinal cord, endoderm, and endothelial genes clustered very strongly together, while presomitic mesoderm genes again were split into two groups, one of which was more closely associated with genes involved in early somitogenesis. We quantified how well cell type-specific genes clustered compared to a random null by first calculating the dispersion of cell types within the tree topology using cophenetic distance (see Methods), and then permuting the leaves of the tree to create a null distribution of the dispersion expected by random. The results produced by hierarchical clustering on scL-score vectors were significantly (p=0.0001) more clustered than would be expected by chance (Figure S4.2a,b).

(n) Clarify how the distance between genes is computed for clustering; the current method is hard to interpret.

We appreciate the reviewer’s suggestion to clarify how distances between genes were computed for hierarchical clustering. To clarify this point, we have added the following to the text:

“Two genes that play similar regulatory or functional roles would be expected to have similar patterns of coexpression and exclusivity across the full gene panel and thus similar L-score vectors. We reasoned that the Euclidean distance between these vectors could be used instead, as it represents the degree to which A and B have a similar scL-score to all other genes considered and satisfies the requirements of a distance measure for the purposes of clustering. We performed hierarchical clustering using the distance between these vectors; the clustering therefore groups genes by the overall similarity of their coexpression profiles rather than by any single pairwise relationship. A heatmap of this clustering (with the pairwise scL-score values displayed between individual genes displayed for clarity) is shown in Figure 3g.”

(o) Quantify similarity between cNMF and L-metric clusters.

We appreciate this suggestion offered by the reviewer as quantifying the agreement between cNMF-derived gene programs and our scL-score-determined clusters will allow readers to more rigorously assess the extent to which these two approaches recover similar groupings of genes. To address this, we compared the top 24 genes of K=7 clusters identified using cNMF to 7 clusters (average 24 genes) obtained from scL-score-based hierarchical clustering at the appropriate cophenetic distance threshold (as originally depicted in Figure S4.2, now in updated Figure S4.4c). We computed the pairwise overlap between every scL cluster and every cNMF cluster and quantified each comparison using the Jaccard Index and Adjusted Rand Index. For each scL cluster, we plotted only the maximum value observed across its 7 possible cNMF cluster comparisons, thereby capturing the strongest correspondence between each scL cluster and the cNMF-defined programs for a given metric.

To establish a baseline for these overlap measures, we designed a reference simulation by preserving the same cNMF clusters while defining a “permuted” set of scL clusters obtained by randomly assigning genes to clusters of the same number (7 clusters) and set of sizes (average 24 genes) as the scL clusters. As above, for each permuted scL cluster and each metric, we retained only the maximum overlap value across 7 possible cNMF cluster comparisons. We note that under this framework, the same cNMF cluster can serve as the highest-overlap comparison for more than one scL cluster.

The following plots summarize the results of applying this approach. Higher values (closer to +1) for the Jaccard Index and Adjusted Rand Index correspond to greater overlap between observed or permuted scL clusters and cNMF clusters. Across both metrics, the observed scL clusters consistently exhibited substantially higher overlap with cNMF clusters compared to permuted scL clusters. For the Jaccard Index, the observed clusters showed markedly elevated values relative to the narrow distribution centered near 0 obtained under permutation, demonstrating that gene overlap between scL clusters and cNMF programs is greater than expected by chance. This similarly holds when gene overlap is assessed using the Adjusted Rand Index. Together, these results quantify how the scL-score can hierarchically derive clusters of genes that recapitulate major gene programs identified by cNMF to an extent beyond that expected under random clustering (See updated Figure S4.4 (formerly Figure S4.2)).

“To validate the clustering produced by the scL-score, we compared our results to a state-of-the-art method for identifying gene programs in an unbiased fashion from single-cell data: consensus non-negative matrix factorization (cNMF) [39]. We pooled nuclei from all individual gastruloids and ran cNMF. We found that many of the resulting clusters (Figure S4.4a,b) corresponded to the clusters identified when the scL-score tree was truncated to produce exactly the same number of clusters (Figure S4.4c). The similarities were even greater when the scL-score clusters were hand-selected based on visual inspection of the tree and density of marker genes (Figure S4.4d). To quantify the overlap between clusters, we calculated both the Jaccard Index and the Adjusted Rand Index (ARI) between each scL-score cluster (Figure S4.4c) and the most similar cNMF cluster. These distributions are shown in Figure S4.4e (blue). We compared to a bootstrapped null where we permuted the genes found in the scL-score clusters, and found that permuted clusters were far less similar to the cNMF clusters than those derived from the real scL-score tree (Figure S4.4e). From these observations, we conclude that the two methods are capable of producing similar results at a high-level, but are different in their application. Individual cells receive component scores for cNMF gene programs, yielding more per-cell information, while the tree produced by L-score clustering reveals hierarchical information about gene programs, which quantifies their similarity in expression on a more global scale.”

(p) In Figure 3e, clarify what the orange lines represent.

We have updated the text:

“Per-cell expression scatterplots of the two pairs of genes shown in b). The y-axis of each is the per-cell expression of Eogt. The x-axis is the per-cell expression of Pax6 (left) or Rfx4 (right). R is Pearson’s r, scL is scL-score. Count data is shown in black; smoothed 2D densities are shown in orange.”

(q) Ensure figures and panels follow the text order. For example, Figure S1.3a is referenced earlier than 1.1 and 1.2.

We appreciate the reviewer’s attention to detail. We have changed the order of these figures so that their reference in the text follows their numeric order.

(r) Typo in "by covariation with any other cell type (Figure S1.1e)" did you mean S1.1.c?

We have updated the text with this change.

(s) For circularity (Figure 6), justify the convex-hull-based measure; thin protrusions can distort interpretation. Consider the volume difference between the convex hull and the original shape.

We thank the reviewer for this helpful suggestion. We tested the difference method suggested by the reviewer, as well as several other methods of clustering and calculating circularity. We determined that the difference in spatial organization of endothelial cells was not robust to changes in method and parameters, so we have chosen to remove that section of the figure and text.

Summary

This work delivers a rich spatial dataset and introduces creative computational tools. The main limitations lie not in the data but in the clarity, justification, and validation of the quantitative methods. We suggest strengthening these aspects by adding formal definitions, parameter justification, benchmarking, robustness tests, and controlled interpretations. This will improve the manuscript's impact and reproducibility of the methods described, besides making it easier to understand for the readers.

We thank the reviewer for their kind assessment and also for their many insightful comments for improvement. We feel the revised manuscript is greatly improved because of them.

Reviewer #3 (Recommendations for the authors):

In my view, this manuscript is well-designed and clearly written, and supports all of the claims made. I have no suggestions for major revisions for this manuscript; rather, I would suggest the following as outstanding questions for future investigation:

We thank the reviewer for a careful reading of our manuscript and the several interesting suggestions and useful references in the literature. Including the discussion of these ideas in the text (see below) has improved the flow and scope of the manuscript.

(1) On the NMP fate, bifurcation has been studied extensively in gastruloids and related structures (see, for example, Underhill et al 2023; Bolondi et al 2024). Can this dataset from Triandafillou and colleagues reveal new regulatory hierarchies in this process? This seems possible in principle, but I was not able to reach this interpretation (for example, it seems the authors interpret the clustering in Figure 3H as reflecting spatial patterns rather than a regulatory hierarchy).

We agree that this is an exciting implication of the work, but we feel that with the current panel (which was originally chosen primarily to type cells and not to infer regulatory structure) we would not be able to comment on this. However we feel that with a larger set of genes such inference may be possible, so we took your suggestion in point 3) and applied the L-metric to a previously existing single-cell dataset. We looked for possible regulatory structure, and found at least one interesting case where a transcription factor showed strong co-expression with two previously unconnected genes. While more specific analyses and experimental validation would be required to establish a direct relationship, we feel that this suggestion by the reviewer represents an important potential future application of this methodology, so we’ve included a new figure and explanatory text to address this possibility:

“scL-score analysis reveals cell type groupings and new transcription factor associations in a single-cell RNA-seq dataset

To test the generality of scL-score analysis, we analyzed a previously-published dataset from [van den Brink 2020] where individual gastruloids (at the same stage as those used in this study) were pooled and subjected to single-cell RNA-seq analysis. After filtering for cell quality and common gene detection, we calculated the scL-score values using a cell-by-gene table of 14304 cells x 19075 genes.

To first test whether we could reproduce the results from this study, we performed hierarchical clustering on the scL-score difference vector (as previously described) on the set of 207 well-detected genes that were also present in our seqFISH gene panel. The resulting tree showed clustered cell types, similar to the tree produced with the expression data in this study (compare Figure S4.5a to Figure S4.4d). The cell types were clustered significantly more than expected by chance (Figure S4.5b).

Then, to test whether scL-score analysis would be effective for analyzing the entire dataset, we performed hierarchical clustering on all 19075 genes. We then examined the resulting heatmap (Figure S4.5c) for clusters of interest. We observed a cluster enriched for endothelial genes (Figure S4.5d), which contained some genes in our panel but many others which were not; this finding demonstrates that the clustering in Figure 4a is not solely due to the selection of genes in our seqFISH panel. We also observed a large cluster that contained genes associated with pluripotency or primordial germ cell fate (Figure S4.5e). Although some of the genes in this cluster were in our seqFISH panel, when we performed scL-score analysis we did not see them cluster with each other or with any other cell type genes. This lack of clustering implies that in our dataset cells that co-express these genes may be rare or too poorly detected to cluster strongly; however, the same analysis performed with more cells and genes showed association. This result demonstrates that clustering scL-score difference vectors can identify known cell-type-associated genes, even within transcriptome-scale data. Finally, we also found a small cluster showing strong co-expression of the transcription factor Gata4, a crucial regulator of the development of visceral and parietal endoderm, and two other genes: a predicted gene of unknown function (Gm43715) and Troponin C (Tnnc1) (Figure S4.5f). Intriguingly, Gata4 has been implicated in heart development (albeit in an indirect manner) [Watt 2004], and troponin C is important for cardiac muscle cell contraction and has been implicated in cardiomyopathy, although at a much later stage of development than that modeled by gastruloids [Li 2015].

Together, these results demonstrate that scL-score analysis is reproducible across datasets, even when different numbers of genes are compared. It effectively clusters genes associated with cell types, and can reveal developmental transitions. Moreover, increasing the number of cells and genes can reveal new clusters, some of which may predict novel regulatory interactions or spatial co-occurrence not previously observed.

(2) On the formation of endothelial clusters ('blood islands'), this process has been observed in gastruloids (Rossi et al 2022). The observation of endoderm-associated endothelium in gastruloids is interesting, but it was also not clear how the authors interpret this finding. Are there two separate endothelial differentiation paths captured here? Or is there one path, and only some migrate towards the endoderm? The authors seem to raise possibilities, and it was slightly unclear on my reading how they interpreted their findings.

We thank the reviewer for pointing us to this paper. We think it’s especially interesting that we also see close association between endoderm and endothelial precursors, especially given that the protocol used in the referenced paper was designed to generate blood precursors (treatment with VEGF, bFGF and ascorbic acid). In light of the comments of other we re-did this analysis — Author response image 7 shows the updated list of differentially expressed genes.

Some of the spatially differentially expressed genes are linked to signalling, and likely reflect overall signalling differences between the anterior (where the somite-associated endothelial cells are) and the posterior (where the endoderm-associated endothelial cells are). For example, Nrp2 enables semaphorin receptor activity, including nervous system development and ventral trunk neural crest cell migration and Tek negatively regulates endothelial cell apoptotic process and response to retinoic acid (RA is higher in the anterior). Pecam1 is involved in adhesion and cell morphology, and perhaps is higher in cells interacting with endoderm due to tighter packing/association; the same could also be true of Cdh5.

While none of these answer the reviewer’s questions about the origin of the cells (and whether it is common), we found another instance of differential endothelial populations in embryo models: in Veenvliet 2020, they find that some endothelial cells have a mesodermal (somitic) origin. Thus we may be seeing a similar phenomenon in our samples. We have updated the text to reflect these additional lines of evidence and to clarify how we think the two populations may differ (while acknowledging that we lack the tools to confidently assign cell of origin or functional differences with this technique):

“We observed that in 5 out of the 26 gastruloids, there was a large central patch of endoderm cells intermixed with endothelial precursors; these samples also had unique spatial L-score clustering of endothelial and endoderm genes (Figure 5b). An example of one such gastruloid is shown in Figure 6a. Migration to and association with the endoderm is also a hallmark of endothelial development [47,48], and we were curious whether there were differences between these cells and the cells we observed forming anterior, somite-associated clusters. When we computed the cell type exposure index for just this gastruloid, we found that, consistent with our visual observations, in this particular sample, endothelial and endoderm cells were much more frequently found next to one another than on average (Figure 6b). To determine whether these spatial and organizational differences reflected gene expression differences, we divided the gastruloid normal to the anterior-posterior axis to separate the endothelial cells into endoderm-associated and somite-associated and looked for differentially expressed genes between the two groups in this gastruloid. To ensure we were focused on genes that truly varied in expression in endothelial cells and were not merely a reflection of spillover from surrounding cells, we pre-filtered genes on expression, so only genes that were present in at least 50% of the cells in either group at a greater than 2 count per cell level were considered. The significantly differentially expressed genes after filtering are shown in Figure 6d. As an additional check on the degree to which transcript mis-assignment affected our analysis of gene expression in these cells in particular, we varied the nuclear dilation in this gastruloid specifically, and calculated cell type score entropy as a function of nuclear dilation (Figure S6.1a). Because cell type score entropy of a cell reflects the degree to which that cell specificity expresses genes associated with a single cell type, our expectation was that if spillover between endoderm and endothelial cells was a significant issue, then decreasing the nuclear dilation should greatly decrease the entropy scores for both groups. Although we saw a slight increase in the spread of the distribution as nuclear dilation increased, the median cell type entropy stayed extremely low for both groups (Figure S6.1a). From this analysis we conclude that the genes we identify as differentially expressed are not due to spillover from surrounding cells, but instead are due to spatially-dependent differences in endothelial cell biology.”

The genes with the highest fold-change in expression in endoderm-associated endothelial genes are shown on the left hand side of Figure 6d. Two are endothelial genes: Pecam1 and Cdh5, both of which are associated with angiogenesis. Spatial expression of these genes is shown in the top row of Figure 6e (larger version in Figure S6.1b). Notch1 is more expressed in endoderm-associated endothelial cells, and this could reflect an increase in Notch signaling in the posterior of the gastruloid. [Chan et al 2017] demonstrated that Notch signalling can be sensitive to shear stress, raising the possibility that the differences in cell state we observe may be driven by differences in mechanical forces in the anterior and posterior. Although most endothelial cells are thought to be of mesodermal origin, some evidence suggests that, in the organogenesis of specific tissues like the liver, the endoderm can give rise to endothelial cells [49]. Furthermore, in [Rossi 2022] the authors show that in a gastruloid-like model specifically designed to model blood development, there is strong spatial adjacency between endothelial and endoderm cells. They hypothesize that these may be a subset of endothelial cells, specifically hemogenic endothelial cells (which have the potential to become blood progenitors). Our data demonstrate a molecularly driven organization distinct from the clustering we observed in the anterior and suggest that multiple mechanisms of endothelial specification could be modeled in gastruloids, even simultaneously within the same structure, although further characterization is needed to determine exactly what processes these unique endodermal/endothelial structures model.

Several other endothelial genes are instead differentially expressed in somite-associated endothelial cells: Nrp2, Tek, Apoe, and Cldn5. Although these genes have less obvious functional distinctions than the endoderm-associated genes, Nrp2 enables semaphorin receptor activity, including nervous system development and ventral trunk neural crest cell migration and Tek negatively regulates endothelial cell apoptotic process and response to retinoic acid (RA), which is known to be higher in the gastruloid anterior. Furthermore, a specialized population of endothelial precursors associated with somites was also observed in trunk-like structures, which are more organized organoids than gastruloids [Veenvliet et al. 2020].

Although endothelial cells have consistently been observed in single-cell measurements of gastruloids, their relative rarity has precluded in-depth analysis of subtypes or inference of spatial location. Our results strongly suggest that endothelial precursor formation, migration, and organization may all be modeled in 3D gastruloids, even without treatment with additional factors as in [Rossi 2021, 2022]; recent advances in 2D gastruloids have allowed modeling of cardiac and hepatic vascularization [45], and our data suggest that 3D gastruloids may similarly be adapted to model more specific aspects of hematopoiesis and vascularization. Early specification from a pool of mesodermal precursors is a hallmark of the endothelial lineage [47]; given the consistency with which we observe endothelial precursors, we speculate that this behavior is recapitulated in gastruloids, but further epigenetic measurements are required to validate this hypothesis” (See Revised Figure 6).

(3) Finally, a broader question about the analytical framework. The authors emphasize that the L-metric is parameter-free; however, much of their analysis still appears to rely on baked-in priors about known marker genes for cell type assignment. Is there a way to extend their analysis to infer cell types directly from the structure of the L-metric? The hierarchical clustering in e.g., Figure 3G suggests something like this: the hierarchy mostly (but not exactly) follows the marker gene annotation. Does this suggest that the cell type labeling should be revisited?

Once of the initial motivations behind the creation of the L-metric (now L-score) was to have a more reliable and specific way of quantifying the interaction between genes that are known in the literature to be cell type markers — with more sophisticated (and noisy) methods of analysis like single-cell RNA sequencing, we found that there was substantial variation in the specificity and ubiquity of so-called ‘marker genes’, and that their usefulness often depended on context. We appreciate that the reviewer raises this point as well, and we think that scL-score analysis, such as that exemplified in Figure S4.4, can help identify new marker genes, or at the very least distinguish the biological context in which a marker gene is useful.

Version 2

Abstract

Gastruloids are stem-cell-based models that recapitulate key aspects of mammalian gastrulation, including formation of an anterior-posterior (AP) axis. However, we do not have detailed spatial information about gene expression and cell type organization, particularly at the level of individual gastruloids. Here, we report a spatially resolved, single-cell molecular catalog of the transcriptomes of 26 individual gastruloids. We found that cell type composition and tissue-scale spatial organization were largely consistent across gastruloids, but meso-scale patterning of specific cell types varied between samples. Posterior cell types formed distinct, organized clusters, while anterior cell types were more disorganized. To distinguish progressive differentiation from cell type differences, we developed the L-score, a parameter-free quantification of mutually exclusive gene expression. This analysis revealed spatial organization without explicit encoding, recapitulated known cell type relationships, and identified novel gene expression states and spatial subclusters within cell types. We confirmed that in gastruloids, NMP differentiation occurred through a continuous, spatially-coordinated process. We also showed that endothelial precursors exhibited unique spatial organization and had distinct gene expression profiles dependent on their association with anterior somitic or posterior endodermal tissues. This work enables the rigorous use of gastruloids as models for studying the molecular mechanisms underlying mammalian development and tissue organization, and introduces new computational tools for analyzing spatially-resolved single-cell datasets.

Introduction

Embryogenesis involves not only the creation but, crucially, the organization of new cells and new cell identities. Gastruloids model an important stage of embryogenesis: gastrulation. Gastruloids break symmetry, elongate along an anterior-posterior (AP) axis, and form many of the cell types found in embryos at a similar developmental stage, as shown by single-cell RNA-seq analyses (Beccari et al., 2018; Turner et al., 2017; van den Brink et al., 2020, 2014; van den Brink and van Oudenaarden, 2021; Veenvliet et al., 2020). These characteristics mean they could hold enormous potential to enable us to deeply characterize, perturb, and test developmental hypotheses, circumventing many practical and ethical constraints imposed by using in vivo models. Yet despite extensive single-cell characterization of gastruloids, we still lack detailed information about the patterning of cell types within individual gastruloids.

Sequencing-based gene expression analyses of gastruloids have at most been able to generate coarse-grained maps of the spatial arrangement of the constituent cells via sectioning (Moris et al., 2020; van den Brink et al., 2020). These methods cannot resolve the position of individual cells, nor can they measure how tissues are organized other than along the AP axis. Computational methods that measure gene expression and identity of individual cells can only estimate AP axis location through inference methods and do not measure organization in any other dimension (Harland et al., 2025; Yaman and Ramanathan, 2023). Additionally, detailed comparative spatial analyses of individual gastruloids have not been performed, as the experiments either require pooling gastruloids or are low-throughput, such that there are insufficient samples to make meaningful comparisons. While some studies point to stringent reproducibility of gene expression along the AP axis in gastruloids (Bennabi et al., 2024; Merle et al., 2024), others show that there is great variation, gastruloid-to-gastruloid, in the organization of specific tissues (Farag et al., 2023). Without high-resolution spatial characterization across multiple gastruloids, it is unclear what aspects of tissue development can be effectively modeled with this system.

The cell types present in gastruloids have largely been inferred from single-cell sequencing analyses. As is standard for these experiments, cells are clustered on the basis of their overall gene expression profiles, and types are assigned based on a qualitative assessment of the genes differentially expressed by the cluster (McNamara et al., 2023; Moris et al., 2020; Regalado et al., 2025; Rosen et al., 2022; van den Brink et al., 2020). However, it is unclear whether discrete binning cells into categories is the only, or even the most useful, approach to single-cell studies of gastruloids, a system nearly entirely composed of cells in the process of differentiating and transitioning between identities. A more gene-centric analysis, especially one coupled to a spatial measurement, could more naturally represent these transitions and how they are coupled to the arrangement and physical location of cells. Whether the diversity of cell types observed in sequencing experiments truly maps onto spatially organized collections of cells, and if this organization is consistent between gastruloids, remains an open question.

To address these gaps, and to create a systematic, high resolution dataset of gene expression in gastruloids considered to be morphologically normal, we developed a spatially resolved, single-cell molecular map of the location, identity, and gene expression of cells within 26 individual gastruloids with normal morphologies. We found that despite some morphological variability within the qualitative category of elongated and polarized, “normal” gastruloids had largely reproducible cell type composition. The order of types and expression of individual genes along the AP axis was also quite consistent. Leveraging our spatial single-cell expression data, we found that the posterior of the gastruloids was characteristically patterned with distinct clusters of progenitors and differentiated cell types. The degree of this clustering varied between samples, as did the mesoscale organization of cell types found near the middle of the AP axis. Some, but not all, of these differences were explainable by variation in morphology. We present novel computational techniques useful for identifying genes with mutually exclusive expression patterns, which illuminated the spatial component of a bipotent progenitor differentiation decision and revealed a spatial hierarchy of gene expression patterns. Using evidence generated from this measure, we show evidence of spatial and functional specialization of endothelial cells, suggesting that gastruloids may be used in the future to model early stages of vasculogenesis and organization.

Results

Cell types’ locations and relative proportions are consistent across morphologically normal gastruloids

To measure the spatial distribution of gene expression, we prepared gastruloids using mouse E14TG2a cells and a standard protocol (see Methods). We harvested mature gastruloids after 120 hours of growth. The experiment was performed 3 times on different days, so to ensure consistency we checked that the proportion of the gastruloids that formed correctly was the same or greater than the median of all experiments (Figure S1.1a). Although there was variation in the length, width, and relative amounts of anterior and posterior tissues in the gastruloids considered, they were within the range of what would qualitatively be considered a ‘morphologically normal’ gastruloid (Merle et al., 2024; van den Brink et al., 2014).

To address potential batch effects due to biological differences between runs, we examined brightfield images of all the gastruloids generated for each experiment (529 total gastruloids across 6 plates on 3 different days), segmented them, and quantified morphological characteristics. When we embedded all 529 gastruloids into PCA space, there was near complete overlap between all groups, with the exception of one plate from 9/1/2024, which was slightly higher in PC1. Figure S1.1b shows this embedding, and examples of gastruloids at the extreme ends of PCs 1 and 2. We note that the samples collected on 9/1/2024 were on average smaller than those from the other two experiments, but spanned the same range of elongation (Figure S1.1c). Interestingly, the final size as measured by cross sectional area of a brightfield image of the gastruloid did not correlate with the initial seeding number (the experiment on 4/4/2025 used 100 starting cells and the other two experiments used 300). Previous studies have demonstrated that the gene expression differences between gastruloids seeded with 100 and 300 cells is extremely small (Bennabi et al., 2025).

We embedded fixed gastruloids whole-mount in polyacrylamide gel and assayed with seqFISH (Eng et al., 2019). An example of raw images showing spots and nuclear quality, as well as deconvolved spots for an example gene Dll1, is shown in Figure S1.1d. To assess the quality of the data, we first assigned an AP axis to each gastruloid using the expression of T, a canonical marker for the posterior (Figure 1a). We compared how gene expression varied along the AP axis, and saw good agreement at a coarse-grained level with a previous study that sectioned gastruloids along the axis and analyzed gene expression in each section (van den Brink et al., 2020) (Figure S1.2a). The colinearity of the peak expression of Hox genes in our panel was also consistent with this dataset, with a median Pearson correlation of 0.695 (compared to 0.663 for all genes, Figure S1.2b).

Cell types’ locations and relative proportions are consistent across morphologically normal gastruloids.

a. Representative gastruloid with the expression of T shown in magenta. Each spot is one transcript (count), nuclei are shown in gray. Thick dashed line denotes the AP axis, thin lines outline the anterior and posterior halves of the gastruloid. Scale bar is 200 μm. b. Gallery of representative gastruloids. Scale bar is the same as a. c. Proportion of each cell type across all gastruloids, ordered by median proportion. d. Coefficient of variation for distributions shown in c. e. Correlation between the ratio of anterior area to posterior area across gastruloids versus the proportion of specific cell types. Cell types shown had a statistically significant (p < 0.05) correlation; all cell types are shown in Figure S1.4e. f. Smoothed density of each cell type along the AP axis. Nuclei in each gastruloid were projected onto the AP axis, and their position was normalized to the total length (left) or symmetrically around a midpoint defined by T expression (right). Traces from individual gastruloids are shown by individual curves.

To profile the spatial organization of individual gastruloids, we assigned a cell type to each nucleus using a cell type scoring method with known marker genes. We compared these results to those obtained with unsupervised clustering. We found that although clustering did produce clusters, they were not strongly separated and we were not able to individually resolve some cell types we expected to find: specifically, there was a mixed differentiation front and presomitic mesoderm cluster, and NMP and spinal cord cells were also clustered together (Figure S3.6a). Therefore, we proceeded with the scoring-based method; see Methods for additional details. A representative gallery of typed gastruloids is shown in Figure 1b; the full dataset is in Figure S1.3. Because each cell received a cell type score for each type, we could use the entropy of the cell type score probability distribution to assess confidence of our cell type assignment: a cell that received a similar score for multiple cell types would have a high entropy distribution, while one which scored highly for one type and low for the rest would have low entropy. On average, the cell type entropies for most cells in a given cell type were low, with the exception of paraxial mesoderm and cardiac mesoderm cells, which had intermediate values (see additional discussion below). The generally low entropies indicate that most of our cell type assignments were high-confidence (Figure S1.4a).

The seqFISH technique is imaging-based, and the fact that we image transcripts in a single plane admits the possibility that due to rotational differences, different parts of the gastruloid are imaged in each sample. We controlled for this to the extent possible within experimental limitations by imaging multiple gastruloids across several experiments and keeping our imaging parameters, particularly the instrument z-depth relative to the coverslip, nearly identical across experiments. We also note that previous analysis of 2D and 3D distances has indicated that in many cases, 2D distances (as we use in this work) are preferable for making comparisons between cells in a sample (Finn et al., 2017).

Once we had the cell type identity and spatial location of each cell in all the gastruloids, we first qualitatively examined where each cell type was found relative to other types and overall morphology. The posterior region, although variable in size (Figure S1.3a,b,c), mainly consisted of neuromesodermal precursors (NMP, turquoise), a bipotent cell type that contributes to both neural and mesodermal tissues (Gouti et al., 2014; Henrique et al., 2015; Tzouanacou et al., 2009). This localization is consistent with previous observations of the expression of NMP genes in the gastruloid posterior (Moris et al., 2020; Underhill and Toettcher, 2023; van den Brink et al., 2020). However, unlike these previous studies, which inferred NMPs from the expression of single genes and could not resolve other cell types, we were able to see that NMPs were intermixed in a patchy fashion with mesodermally-fated cells (presomitic mesoderm, PSM, red). In some instances, we saw PSM cells forming symmetric patches on either side of a central neurally-fated region (Figure 1b, center gastruloid, Figure S1.3a v., xi., xii, b iv.), mimicking the organization found in embryos (Gossler and Tam, 2002; Tam and Behringer, 1997; Xie et al., 2025) and thought to occur in gastruloids (Yamanaka et al., 2023; Yaman and Ramanathan, 2023). Generally, we saw 1-2 distinct clusters of neurally fated cells (spinal cord precursors, black), often adjacent to or opposite the region associated with somitic fate acquisition (“differentiation front”, light blue), although occasionally found in the center of the gastruloid (Figure S1.2a ii., vi., xvii). Although distinct organization of neurally fated cells has been observed in other embryo models (Veenvliet et al., 2020), the organization of spinal cord precursors in gastruloids in the context of other cell types has not yet been systematically characterized; staining for spinal cord-associated genes showed compelling patterning, but the patterns varied by gene, complicating interpretation of the organization of the tissue (Beccari et al., 2018). Consistent with (van den Brink et al., 2020), the differentiation front was nearly always found in one distinct cluster marking the boundary between the anterior and posterior (Figure S1.3a i., iii., vi., b iii., vi). The anterior regions consisted mainly of somitic cells, presomitic cells, and cardiac mesoderm (yellow, green, and purple) (Figure S1.3 a iii., v., x., xv., xvi.). We also detected endoderm cells, although not in all gastruloids, consistent with observations in (Farag et al., 2023). This simultaneous mapping of all cell types is qualitatively consistent with many observations in the literature about the expression of individual genes and integrates them into one consistent picture of cell type organization.

We sought to quantify variability in cell type composition between the 26 morphologically normal gastruloids profiled. Previous single-cell datasets relied on pooling multiple gastruloids, thus obscuring the degree to which the overall cell type distribution was reflected in each individual gastruloid. However, recent single-cell measurements of individual gastruloids have suggested substantial gastruloid-to-gastruloid variation in cell type proportions (Rosen et al., 2022). Figure 1c shows distributions of cell type proportions across samples, and Figure 1d shows the coefficient of variation of these proportions. Individual gastruloid cell type distributions, including the proportion of cells that had insufficient reads to be confidently assigned a type, are shown in Figure S1.4b,c. We found that cardiac mesoderm, endoderm, and spinal cord cells had the greatest coefficient of variation in proportion between gastruloids (Figure 1d). To calculate statistical significance, we first performed a centered log-ratio (CLR) transform on the proportions, then looked for covariation between cell types across gastruloids. We found there was a statistically significant inverse correlation between the proportion of endoderm and NMP, presomitic mesoderm, and differentiation front (Figure S1.4d). We did not observe gastruloids that were as strongly neurally-biased as those reported in (Rosen et al., 2022), but we did see some gastruloids with a relatively high proportion of spinal cord precursor cells (Figure S1.3a ii., xv., b vii.), and overall the proportion of spinal cord had a negative covariation with the mesodermally-derived cell types, consistent with the anticorrelation also reported in (Rosen et al., 2022) (Figure S1.4).

The proportion of presomitic mesoderm cells was significantly positively correlated with the proportion of somite cells (covariation = 0.63, Figure S1.4d). Presomitic mesoderm cells progressively take on somitic fate at and after the differentiation front; a positive correlation between these two would suggest that the amount of presomitic mesoderm is linked to how much somite is ultimately formed. How much presomitic mesoderm is formed depends on fate decisions made by NMPs: NMP progeny can become either mesodermal or neural, depending on their local signaling environment. This chain of correlation suggests that fate decisions made by NMPs may drive the overall composition of individual gastruloids and that, as in embryos, a balance between the amount of neural and mesodermal tissue is achieved, although likely to a less stringent degree (Garriock et al., 2015). Further experiments could delineate whether, as in embryos, Wnt3a drives this balance (Garriock et al., 2015), and the degree to which growth of differentiated progeny affects compositional variation (Underhill and Toettcher, 2023).

Finally, we quantified correlations between morphology and cell type proportion: we used the segmented images of the gastruloids to generate an anterior to posterior area ratio, and looked for correlation between this ratio and the proportion of individual cell types. We found that only cardiac mesoderm, endothelium, and paraxial mesoderm cells exhibited significant correlation with A:P ratio, and in all cases the correlation was positive (Figure 1e, all cell types in Figure S1.4e).

Cell types’ arrangement along the AP axis is highly consistent across gastruloids

Given that the proportions of cell types within each gastruloid were fairly consistent, to what degree did their spatial organization vary? We first projected each cell of each type onto the AP axis and looked at the distribution of AP axis locations across gastruloids (Figure 1f, left hand side). The most posterior cell types (NMP and presomitic mesoderm) showed wide distributions from 0-30% of the axis. Centered around 30%, spinal cord precursors and endoderm had distinctive peaks (clusters), the exact location of which varied between gastruloids. Using a threshold of T expression to define the midpoint of each gastruloid and uniformly length-normalizing each half caused these cell types to collapse into a single peak (Figure 1f, right hand side). The differentiation front was similarly located in one peak (between 30-40% of the AP axis), and the variation in the location of this peak decreased with T-expression normalization. To quantify this variability, we bootstrapped a null distribution of cell type locations by pooling each cell type together across samples and using the resulting distribution to define a reference mean. When we compared how much peak variation there was among samples randomly drawn from this null to our observed data, we found that although nearly every cell type (excepting paraxial mesoderm) had more variability than expected by chance, the effect size of this variation was small (Figure S1.5a). Normalizing location to T-expression decreased variability in most cell types, most notably for differentiation front and spinal cord (Figure S1.5b).

We also asked how the order of cell types along the AP axis varied between gastruloids. When we ranked the peaks shown in Figure 1f per gastruloid, we found that Kendall’s W, an overall measure of rank coherence across independent samples that spans from 0 (no agreement) to 1 (complete agreement), was 0.834 (Figure S1.5c). We found that the cell types most likely to swap rank order were spinal cord and endoderm, and paraxial mesoderm and endothelium (Figure S1.5d).

Overall cell type mixing varies between gastruloids

Several studies of gene expression in gastruloids have used pooled measurements to infer the AP axis-location of genes and cell types (Anlas et al., 2025; Moris et al., 2020; Rosen et al., 2022; Veenvliet et al., 2020; Yamanaka et al., 2023) and our data are largely consistent with these lower-resolution findings. Yet it is obvious from individual gene staining (Beccari et al., 2018; Braccioli et al., 2022; van den Brink et al., 2014; Yaman and Ramanathan, 2023) and our detailed 2D maps of cell identity and location that gastruloid organization is much more complex than the average order of cells along the AP axis. We thus needed an analytical method for quantifying spatial organization beyond distributions along the AP axis. To further characterize spatial organization, we sought to quantify the degree to which cells were mixed in each gastruloid, and how that mixing might vary between gastruloids. For each cell in each gastruloid, we counted the interactions between that cell and all its neighbors within a 16 μm radius (on average 5-6 neighbors per cell), and summarized all these interactions for all cells in the gastruloid in a matrix, normalizing each element by the frequency of the cell type considered to be the ‘neighbor’ in the interaction.

To quantify overall mixing, we calculated the sum (across types) of the frequency of self interactions, normalized to the total interactions and then took the inverse. We normalized this value to the expected cross-type interactions predicted from random mixing (i.e. the proportion of the neighboring cell type). This gave us, for each gastruloid, a single value that we call the mixing index. The mixing index ranged from -1 (totally segregated) to +1 (totally mixed with less frequent self-interactions than expected from chance); a score of 0 indicates a random distribution, i.e. neighbor frequency is exactly what would be predicted by that cell type’s frequency alone. The mixing index values range from -0.50 to -0.22 (Figure 2a). While all values were negative, the range was large. This observation led us to conclude that while in all the gastruloids profiled cell types tended to cluster together, there was variation between gastruloids in the degree of coherent clustering between types.

Pairwise cell type interactions quantify cell type mixing and interaction motifs.

a. Distribution of mixing indices across gastruloids. b. Exposure index, which represents how frequently the source cell type (row) is found next to the target cell type (column). The maximum possible value is 1, self-interactions excluded for readability. c. Significantly enriched and depleted triplet combinations of cell types, ordered by effect size from left to right. Bars are colored by the composition of the triplet they represent. Only triplets found in over half of the gastruloids are shown.

Pairwise cell type interactions suggest more cell type mixing in the gastruloid anterior than posterior

The interaction between long-range signaling gradients and short-range cell-cell interactions is thought to drive gastruloid symmetry breaking and patterning (Arias et al., 2022; Gupta et al., 2021; Simunovic and Brivanlou, 2017), and our spatial data allowed us to query both micro and meso-scale organization. Our simultaneous observation of variability in mixing index and consistent cell type organization along the AP axis led us to ask if mixing was driven largely by local topology and whether this effect was disproportionately due to the arrangement of a subset of cell types. We noted that the off-diagonal elements of the matrix we used to calculate the mixing index were informative about cell type-cell type interactions. Specifically they quantify the degree to which each cell type (source) is exposed to another cell type (neighbors). To assess the overall frequency of cell type-cell type interactions, we first pooled the data from all gastruloids together into one interaction matrix, the elements of which we refer to as the ‘exposure index’. (Figure 2b, Figure S2.1a,b). Higher values indicate a greater frequency of being found in close proximity (see Methods for full details). The matrix is asymmetric in that the exposure of cell type A to B may not be the same as the exposure of cell type B to A. For example, a cell type found in singlets within a larger tissue will be more frequently exposed to cells of that common tissue type, while the common type will see itself much more frequently than the intermixed cell. Thus, this asymmetry accounts for the fact that local patterning can influence the degree to which any one cell type is found next to another. For instance, somite and endothelial precursors have an asymmetric pattern: although endothelial cells are often exposed to somite cells (exposure index 0.21), somite cells are comparatively infrequently exposed to endothelial cells (0.07). This observation, coupled with a fairly strong endothelial-endothelial exposure index (Figure S2.1b), suggests small clusters of endothelial cells are found among somite cells, reflecting our visual impressions. In contrast, cardiac mesoderm cells were often exposed to all other anterior cell types, but their self-exposure index was low (Figure S2.1b), matching our observation that they are found in inconsistent and disorganized clumps or individually.

The overall interaction matrix (Figure 2b) also revealed that there were more inter-type exposures between the anterior cell types of the gastruloid than between the posterior types. This quantification matches the impression from visual inspection that the gastruloids contain both somitic cells and paraxial mesoderm cells that properly localize to the anterior (as per (van den Brink et al., 2020; Yaman and Ramanathan, 2023)) but are arranged in a disorganized fashion. Cells of the posterior, by contrast, were much less mixed, with individual cell types forming large, distinct clusters. These results demonstrate that our analytical methods were able to capture and quantify the degree of spatial disorder in these two compartments. The contrast between organized posterior clustering and disorganized anterior mixing matches expectations based on literature that self-organization mechanisms in gastruloids in the anterior vs. posterior are differentially sensitive to culture conditions, with somitic patterning requiring external matrix support (Beccari et al., 2018; van den Brink et al., 2014; Yamanaka et al., 2023; Yaman and Ramanathan, 2023), distinguishing it from the seemingly more autonomous organization observed in posterior cell types.

Having established the general expectations for cell type interactions, we then looked for deviation from these patterns. We quantified the variance across gastruloids in the exposure index by creating an interaction matrix for each individual gastruloid and comparing the distributions of values for each interaction. We found that for most, the gastruloid-to-gastruloid variation was not significant. However, there were several self-interactions that were significantly variable: specifically presomitic mesoderm, spinal cord, endoderm, and differentiation front (Figure S2.1c). This likely reflects differences in the size and shape of clusters of these cell types. There was only one pair of different cell types that significantly varied: endoderm and endothelium. We show additional evidence for the unique arrangement of these cell types in Figures 5 and 6.

One potential caveat to this finding is that differences in uncertainty in cell typing could be the primary driver of mixing and cell type interaction differences, both between the anterior and the posterior within an individual gastruloid, or overall between gastruloid. To control for this, we applied an entropy filter to our dataset. We filtered out cells that had entropy > 1.5 (see plot below for cutoff), which was chosen based on the distribution of entropy values for ‘none’ type cells, which effectively describe the upper limit of random transcript assignment (99.7% of ‘none’ type cells are removed with this filter, and about 50% of cardiac mesoderm cells and paraxial mesoderm cells, see Figure S2.2a). We first examined overall mixing; there was strong correlation between the per-gastruloid mixing indices before and after entropy filtering (Pearson r = 0.809, Figure S2.2b). Globally, mixing indices decreased with filtering, meaning that overall the cell types were more clustered. When we compared the absolute value of the change in mixing index pre and post filtering to the proportion of each cell type, the only significant correlation was with cardiac mesoderm (Figure S2.2c). From these analyses we conclude that ambiguous cardiac mesoderm cells drive some of the per-gastruloid mixing index. Exposure indices were overall quite similar after filtering, although the strength of somite-somite and somite-paraxial mesoderm interactions increased (Figure S2.2d).

We also examined how entropy filtering might affect the exposure index, given that the mixing index is calculated from the exposure index of across cell types. In general, the magnitude of the exposure index values increased when more uncertain cells were excluded, but the directionality and relative ordering was not affected. Although the magnitude of change in the posterior cells types was less than the anterior cell types, the cross-cell type exposure values, particularly between paraxial mesoderm/endothelium and somites, doubled. From these results we conclude that mixing in the posterior is driven mainly by NMP/presomitic mesoderm interactions, and is overall lower than mixing in the anterior, which is driven by rarer cell types like cardiac mesoderm, paraxial mesoderm, and endothelium, being interspersed within somite cells.

Micro-scale organization is preserved across gastruloids

Our analyses focused on exposure of one cell type to another, however given that some have reported consistency in gene expression in gastruloids to the resolution of single cells (Merle et al., 2024) we were curious how hyper-local interactions might vary. To measure such interactions, we first determined all the possible combinations of three cells (triplet motifs) from the 9 cell types we had identified (with replacement). Then, for each cell in the gastruloid, we identified the two nearest neighbors and determined which triplet type it was. We calculated the enrichment of each triplet motif relative to its expected frequency based on the amount of each of the constituent cell types. Figure 2c shows this distribution of triplet motifs that were found in over half the gastruloids and in each were significantly enriched or depleted based on overall cell type frequency. We found that motifs formed from 3 of the same cell type were overwhelmingly the most significantly enriched, and all were found in over half of the gastruloids. The enriched 2+1 (two same) motifs were composed of all posterior cell types or all anterior cell types. 3/4 of the most common significantly enriched motifs that were of three different cell types consisted of cardiac mesoderm with two other anterior cell types or differentiation front, again suggesting that the cardiac mesoderm cells are uniquely spread out as individuals.

We also identified motifs that were highly underrepresented compared to what would be expected given the frequency of occurrence of each of the individual cell types: nearly all of the significantly underrepresented motifs were of 3 different cell types, with the rarest being somite+NMP+presomitic mesoderm. This observation was initially surprising given that these cells are all thought (in gastruloids) to derive from the same lineage; however, the early mesodermal precursors are spatially segregated from the anterior by the differentiation front, which is where somite determination happens (van den Brink et al., 2020; Veenvliet et al., 2020; Yamanaka et al., 2023). The relative paucity of this triplet indicates that the spatial organization of the differentiation process occurs with very high fidelity, and that less differentiated precursors rarely make it into the anterior, where they would be exposed to somite and paraxial mesoderm cells.

Variation in cell type abundance and organization is structured and concentrated in specific cell types

We have demonstrated that some aspects of gastruloid composition and spatial organization are consistent across gastruloids, while others are more variable. Consistent features include proportions for NMP, presomitic mesoderm, somite, and paraxial mesoderm, whose coefficients of variation were lower than other cell types (Figure 1d). Organizationally, all cell types across gastruloids are more physically clustered than random (Figure 2a), and the order in which cell types are found along the AP axis has statistically significant high agreement between gastruloids as measured by Kendall’s W (Figure S1.5c). At the local neighborhood scale, we found that most cell type interactions were conserved across gastruloids (Figure S2.1c). At the local scale, across individual gastruloids, we found many motifs of three cells that were statistically enriched over random, suggesting a conserved local order (Figure 2c). While the normalized distance along the AP-axis of all cell types significantly varied compared to a bootstrapped null (Figure S1.5a), the effect size was small, and decreased in almost all cases when normalized to gene expression (of T) in addition to morphology (Figure S1.5b).

However, there were also variable features. The proportion of cardiac mesoderm, endoderm, and spinal cord had the highest coefficient of variation between gastruloids (Figure 1d). Because proportions must sum to one, a change in the proportion of one cell type is necessarily linked to changes in others; we performed centered log transformation and looked for statistically significant covariation. Of all possible pairings, the following proportions had a significantly negative correlation across samples: endoderm/differentiation front, NMP/endoderm, presomitic mesoderm/endoderm, none/endothelial, and spinal cord/endothelium. This result shows that the proportions of these cell types predictably co-vary between samples, potentially suggesting some kind of biological trade-off in cell type specification or organization (Figure S1.4d).

Across gastruloids, intra-cell type interactions (degree of clustering) of spinal cord, endoderm, and differentiation front vary (Figure S2.1b). This variation suggests that these cell types may be patterned differently between gastruloids. For example, the local motif of 3 endoderm cells found next to one another was statistically enriched within some but not all individual gastruloids, and by definition is completely absent from gastruloids lacking endoderm (Figure 2c). We interpret this contrast to mean that when endoderm is found in a gastruloid, it is consistently patterned at a local level, but may vary more at a global level. This interpretation is concordant with the findings from (Farag et al., 2023), which demonstrate several distinct classes of endoderm organization in gastruloids.

To summarize, while changes in the amount of individual cell types can vary, these changes are in most cases explained by variations in morphology and molecular characteristics (such as anterior:posterior ratio and the expression of morphogens like T). For patterning, we found that, in most cases, global patterns were conserved, but there were variations in local patterning that may lead to variable meso-scale organization of specific cell types, particularly those found in the middle of the anterior-posterior axis.

Coexpression of cell type marker genes delineates progressive NMP differentiation in gastruloids

One of the central goals of studying developmental processes is to determine the means by which cells make fate decisions and when these decisions are driven by the physical location of the cell as opposed to cell-intrinsic programs. A large part of gastruloid development is thought to be driven by the differentiation, growth, and migration of NMPs and their progeny, which are fated towards either a mesodermal or posterior neural (spinal cord) fate (Bolondi et al., 2024; Braccioli et al., 2022; Veenvliet et al., 2020). In embryos, this process is thought to happen in a continuous fashion; NMP progeny simultaneously move anteriorly and acquire either a mesodermal or spinal cord fate (Garriock et al., 2015; Gouti et al., 2017; Henrique et al., 2015; Jin et al., 2025). The differently fated cells are driven to different parts of the developing spinal column; lineage tracing evidence suggests that fate commitment is at least partially driven by physical location and the local signaling environment (Garriock et al., 2015), and is progressive in nature (Jin et al., 2025). We wondered how this coupling between space and differentiation was resolved in gastruloids, and whether the continuous transition from progenitor to more differentiated cell type was reflected in spatially continuous patterns of gene expression.

Our data represent a single slice of time during which NMPs, presomitic mesoderm, and spinal cord precursors exist simultaneously; if this process happened continuously, we would also expect to see a subset of cells somewhere on an identity continuum between these types (Figure 3a). For simplicity, when initially assigning cell types, we used differential expression of panels of marker genes to assign identity. However, many of the cells in the posterior had non-zero cell type scores in all three categories (Figure S3.1a). This observation motivated us to leverage the advantages of our dataset to delineate where individual genes were expressed in space with single-cell resolution and determine whether that additional level of resolution gave additional information about NMP differentiation trajectories.

Progressive NMP differentiation is revealed by the single cell L-score.

a. Illustration of NMP differentiation with predictions for where along the trajectory gene expression is shared or unique. b. Per cell expression of all NMP exclusive genes summed versus all other gene categories summed, one plot per type. n=25514 cells typed as NMP, presomitic mesoderm, or spinal cord in n=26 gastruloids. Pearson correlation is shown with colored lines. c. Expression plot showing the relative amount of Eogt vs Pax6 (left) and Eogt vs Rfx4 (right) per cell in a representative gastruloid. Pax6 and Rfx4 are both annotated as spinal cord-associated genes. d. Per cell expression of all spinal cord exclusive genes summed versus expression of all presomitic mesoderm genes summed. n=25514 cells typed as NMP, presomitic mesoderm, or spinal cord in n=26 gastruloids. e. Per cell expression scatterplots of the two pairs of genes shown in b). The y axis of each is the per-cell expression of Eogt. The x-axis is the per-cell expression of Pax6 (left) or Rfx4 (right). R is Pearson’s r, scL is scL-score. Count data is shown in black; smoothed 2D densities are shown in orange. f. Distribution of scL-score values for pairs of spinal cord and presomitic mesoderm genes (left), NMP and presomitic mesoderm genes (center), and NMP and spinal cord genes (right). g. Hierarchical clustering heatmap of scL-score vectors for NMP, presomitic mesoderm, spinal cord, and combined category genes. scL-score difference vectors were used for clustering; scL-score values are shown in the heatmap. Blocks highlighted are discussed in the text. h. Tree of the hierarchical relationships between genes resulting from the clustering shown in e). Color indicating the gene type is shown at the bottom, legend is the same as in g. The density plots shown are the summed, averaged gene expression of all genes in the leaves up until that node, smoothed with a 2D density kernel estimate (see Methods for details about smoothing).

Since both presomitic mesoderm and spinal cord cells derive from NMPs, we reasoned that it was possible, especially if the cells were still in the process of differentiating, for cells typed as presomitic mesoderm or spinal cord to residually express NMP genes. Indeed, there were genes in our panel that were shared between each type and the precursor (referred to here as NMP+presomitic mesoderm genes and NMP+spinal cord genes or “shared genes”). Figure 3b shows the total expression of each gene group versus the NMP genes for all cells in all gastruloids that we typed as NMP, presomitic mesoderm, or spinal cord. As expected, there was a clear correlation between the mixed categories and NMP genes, supporting the notion of a continuous differentiation process.

In an effort to locate cells that had fully committed to one lineage or the other, we looked at the expression of genes thought to be exclusively associated with presomitic mesoderm or spinal cord. Our expectation was that the expression of these genes would be mutually exclusive, and several genes did display this pattern (Figure 3c, right). However, we were surprised to find many examples where these supposed marker genes for different cell types were expressed in the same cell (Figure 3c, left). Spot assignment has the potential to mis-assign transcripts, which could, in principle, lead to artifactual mixed states. However, we don’t believe misassignment explained our observations. We used stringent nuclear dilation (small enough that ∼30% of transcripts were not assigned to a nucleus). Moreover, when we calculated the exposure index, we found very little exposure between presomitic mesoderm and spinal cord cells (0.08 and 0.05, Figure 2b), meaning the majority of the time these cells were not nearest neighbors, and thus had little potential for overlap and transcript mis-assignment.

Was this trend of coexpression true for the two groups of marker genes as a whole? When we compared the total per-cell expression of the genes in both categories (spinal cord and presomitic mesoderm), we noted that while they did show more mutually exclusive (i.e. L-shaped) expression than when compared to the NMP genes, there were clearly many cells that had substantial counts from both PSM and spinal cord gene categories (Figure 3d). One possible explanation for the coexpression of genes associated with both arms of the differentiation trajectory is that these cells had not yet fully committed to either lineage. It is also possible that a subset of the genes were better (or at least more exclusive) markers than others (as our observation of the heterogeneity in expression patterns suggested), and that there may in fact be a continuum of markers, some of which may distinguish certain cell types and not others. To determine if any of these possibilities were true, we needed a way to identify genes that were expressed in a mutually exclusive manner.

The L-score quantifies mutually exclusive gene expression

Marker genes for distinct lineages are often assumed to be mutually exclusive in their expression. However, the above results show that such assumptions may not hold. We were curious whether there were genes that were truly exclusively associated with each cell type, and if such genes in general would prove to be more effective marker genes than those that were merely associated with the cell type.

To this end, we developed a pairwise measure between genes that reported the degree of mutually exclusive expression. It is calculated by rank ordering cells by the expression of one gene and measuring the degree to which the expression of the other gene is anti-rank-ordered (see Methods for details and Figure S3.2 for a visual explanation of how the measure is calculated). We call this measure the “single-cell L-score” (scL-score) because when the per-cell expression of mutually exclusive genes is plotted against one another, the data make an L shape (Figure 3e, right). A value of -1 represents perfectly mutually exclusive expression, which only happens when the genes are never found in the same cell. Higher values indicate more coexpression. Genes that are ubiquitously expressed without a strong correlative relationship between them will have a score of ∼0. The maximum possible value is 1, which is obtained when two genes are expressed only in a shared subset of cells and are never found apart. We refer to this as ‘perfect coexpression’. Our expectation is that ubiquitously expressed genes like cell cycle and housekeeping genes will, due to the rank-ordered nature of the L-score calculation, have L-scores consistently close to 0 no matter which genes they are compared with, whereas genes that are specifically associated with a single cell type will have an scL-score value close to -1 when compared with genes specific to other types, but higher values when compared with genes associated with the same cell type. The results of our simulations confirmed these hypotheses (Figure S3.3a).

To benchmark this measure against existing exclusivity or coexpression measures, we calculated the Exclusively Expressed Index (EEI) (Nakajima et al., 2021) and Coefficient of Expression (COEX) (Galfrè et al., 2021) for the same simulated datasets (Figure S3.3b) and a subset of NMP/presomitic mesoderm/spinal cord genes (Figure S3.4a). All three methods were able, to some extent, to distinguish mutual exclusivity from coexpression, but the scL-score provided clearer separation between these different relationship types; a more detailed description of the analysis is included with Figure S3.3 and Figure S3.4.

We compared the distribution of L-score values between spinal cord and presomitic mesoderm genes; as our previous results suggested, there were many genes that had very strong mutually exclusive expression, while others were expressed in the same cells. For example, Pax6 (spinal cord) and Eogt (presomitic mesoderm) are often expressed in the same cells; they have a mild positive correlation as measured by Pearson’s R (0.24), and an scL-score of -0.55 (Figure 3e, left). However, Rfx4 (spinal cord) and Eogt (presomitic mesoderm) were expressed mostly in distinct cells; their expression was nearly uncorrelated when measured by Pearson’s R (-0.02), but they have an scL-score of -0.96 (Figure 3e, right). Thus, Rfx4 appears to be a good marker for differentiated spinal cord cells, while Pax6 might mark cells that are neurally fated but not yet committed. We also note that it is not possible to distinguish from these analyses whether all differentiated spinal cord cells express Rfx4, merely that it is exclusively associated with this type.

Overall, the distribution of spinal cord and presomitic mesoderm gene scL-score values was strongly skewed towards -1, with at least 20 pairs of genes strongly distinguishing between the two types (Figure 3f, left). From the 54 possible pairs of spinal cord and presomitic mesoderm genes, we found that 24 (44.4%) had scL-score values < -0.9, likely indicating that they mark different populations of cells.

We predicted that there would be more coexpression (i.e., higher scL-score) between NMP genes and the other two categories (presomitic mesoderm and spinal cord), since these two types of cells derive from NMPs. When we compared the distribution of scL-score values of NMP and spinal cord or NMP and presomitic mesoderm gene pairs, we did in fact see a shift of the distribution to higher values compared to the presomitic mesoderm and spinal cord pairs, indicating more coexpression (Figure 3f, right).

Despite this shift to higher values, some pairs still seemed to delineate the cell types. For example, Fgf9 (presomitic mesoderm) and Wnt3a (NMP) have a very low scL scores (-0.84). There were even more low-scL pairs when comparing NMP and spinal cord genes, such as Sox21 and Wnt3a (L=-0.98) and Scube2 (spinal cord) and Cdx1 (NMP) (scL=-0.91). Notably, the average scL-score of presomitic mesoderm and NMP gene pairs was much higher than spinal cord and NMP gene pairs, which suggests that perhaps commitment to a neural pathway involves a more systematic reorganization of gene expression.

Analysis using this measure allowed us to distinguish genes such as Eogt and Rfx4 that clearly mark distinct subsets of cells, from those like Eogt and Pax6 that partially overlap and may mark intermediates as well as more differentiated cells. We also observed a unique pattern when we compared the expression of Rfx4, which we identified as a good marker of spinal cord genes, with Nkx1-2, a shared gene annotated as being associated with both NMPs and spinal cord cells. The expression of Nkx1-2 was strongest in the posterior, but extended into the region marked by Rfx4. It seemed to be progressively downregulated in this region from posterior to anterior, producing a graded expression pattern with the anterior-most cells only expressing Rfx4 (Figure S3.1b, c). This observation suggests that shared genes can mark cells on their way to becoming spinal cord cells, and at least some are turned off as the cells differentiate, suggesting that the continuous differentiation pathway observed in embryos is also found in gastruloids.

These graded expression patterns led us to be more broadly interested in the patterns of spatial expression and scL-score values of genes that were annotated as belonging to more than one cell type. Up until now, we had excluded such genes from our analyses, but we reasoned that their distribution and coexpression might reveal more about the path that NMPs take on their way to becoming presomitic mesoderm or spinal cord cells. Shared NMP+presomitic mesoderm genes included T, Fgf17, and Hoxaas3, while shared NMP+spinal cord genes included Sox2, Gbx2, and Nkx1-2. When we compared both shared categories to their terminal cell type, we found on average that the distributions of L-score values were positively skewed compared to the NMP and terminal cell type distributions, indicating that, as expected, there was even more coexpression among these categories (Figure S3.1d). Interestingly, we found that both distributions were bimodal: a subset of genes were clustered around 0, indicating high coexpression with the terminal cell type genes (this included Dll1 (presomitic mesoderm) and T (NMP+presomitic mesoderm), and Cdh6 (spinal cord) and Nkx1-2 (spinal cord+NMP). Others were closer to -1, meaning the gene in the shared category is either more associated with the parental type (NMPs) or potentially is specific to intermediates. Examples of these pairs include Mesp2 (presomitic mesoderm) and Msgn1 (presomitic mesoderm+NMP), and Rfx4 (spinal cord) and Hoxb6 (spinal cord+NMP).

We wondered whether we could use the scL-scores to refine our marker gene panel. We reasoned that ‘good’ marker genes would have very low scL-scores with most other genes and have high scL-scores with a small subset of genes (presumably those belonging to the same type). However, filtering on these criteria did not markedly improve our cell type scoring, likely because the genes in our panel were hand-selected to be good type representatives, so most fit this criterion (Figure S3.5a, b and Methods). However, we found that scL-score analysis could be used to quantify how related cell types were in terms of marker gene expression (Figure S3.5c), and moreover was extremely effective at assessing the outcome of unsupervised clustering, including quantifying overlap in gene expression between clusters and identifying gene pairs that were mutually exclusively expressed between clusters, even when they were highly related (Figure S3.6).

This analysis demonstrates that we could use the scL-score to delineate, in an automated fashion, whether genes reported in the literature as associated with multiple clusters in single-cell data were truly associated with multiple cell types, without parameter tuning and using only the cell-by-gene table.

The scL-score captures the spatial distribution of gene expression despite being calculated without spatial information

The amount of information about individual gene pairs we were able to extract from this relatively simple measure led us to wonder whether it could also be used to generate a global picture of the relatedness of genes associated with NMPs and their derivatives.

For each gene annotated as belonging to any of the three cell types (NMP, PSM, or spinal cord), we calculated a vector of scL-score values with all other genes. Two genes that play similar regulatory or functional roles would be expected to have similar patterns of coexpression and exclusivity across the full gene panel and thus similar L-score vectors. We reasoned that the Euclidean distance between these vectors could be used for clustering, as it represents the degree to which A and B have a similar scL-score to all other genes considered and satisfies the requirements of a distance measure. We performed hierarchical clustering; a heatmap of the resulting relationships (with the pairwise scL-score values between individual genes displayed for clarity) is shown in Figure 3g. Broadly speaking, the genes associated with each cell type clustered together, and the shared category genes generally clustered closely to the NMP genes. Interestingly, although both the NMP and spinal-cord exclusive genes were each found in one large type-specific cluster (Figure 3g ii., v.), the presomitic mesoderm exclusive genes were split into two groups (Figure 3g i., iv.). There was also a subset of NMP-exclusive and spinal cord-exclusive genes that clustered together and separately from the rest of their group (Figure 3g iii.).

To determine what could be driving these differences, we plotted where the genes were expressed in space. Since the clustering was hierarchical, we first made a tree representing this hierarchical relationship, and then at each node of the tree, we made a summed density plot of all the genes contained below that node. The density plots at the top of the tree show broad distributions of expression, but as we followed the branches down closer to the leaves, we saw spatial patterns begin to emerge. We were surprised to find that the scL-score, which doesn’t inherently contain any spatial information, was able to capture a spatial tree of gene expression.

Most of the NMP-exclusive genes were found very close to the posterior and were diffusely expressed. However, in the mixed cluster, which contained both NMP-exclusive and spinal cord-exclusive genes, we saw a different pattern — here the NMP genes were more patchily expressed and extended far more anteriorly (Figure 3h i.), suggesting that genes in this group may mark cells destined to become part of the spinal cord. Indeed, one such gene was Epha5, an ephrin receptor differentially expressed in a cluster labeled as NMPs by single-cell sequencing (van den Brink et al., 2020) but which has separately been annotated to be involved in neural development (Akaneya et al., 2010). We also saw a clear division between more anteriorly and more posteriorly expressed spinal cord genes (Figure 3h ii..) Thus, clustering L-score values allowed us to delineate differences in behavior between cell type-associated genes: one cluster of spinal cord and NMP genes appears to mark neurally fated but uncommitted cells (Figure 3h i.), while another marks committed spinal-cord fated cells (Figure 3h ii.).

We also looked more deeply into the split presomitic mesoderm exclusive genes. Not only were they expressed in different parts of the gastruloid (compare Figure 3h iii., iv.), we realized that these genes were associated with presomitic mesoderm cells in different states: Mesp2 is turned on in a specific and spatially constrained location at the onset of somitogenesis (Veenvliet et al., 2020), a process which has only just started in this gastruloid, while Dll1 more generally marks the presomitic mesoderm. Thus, using only differences in scL-score values, we were able to identify both general presomitic mesoderm versus presomitic mesoderm that was starting to undergo somitogenesis.

These analyses demonstrate that information contained within the hierarchical relationships between genes, determined by scL-score can reveal novel information about cell states within cell types, although we acknowledge that since cell type is determined by a limited panel of marker genes, results should be further functionally verified. scL-score analysis can also identify distinct spatial locations of cells in this cell state, all without explicit encoding of spatial information, but rather quantifying and clustering the degree to which genes are mutually exclusively expressed with one another.

Clustering scL-score vectors clearly resolves cell types and reveals novel genetic interactions

Given the amount of spatial and state information that was encoded in the scL-score heatmap for a subset of our gene panel, we expanded our analyses to all genes, hoping to discover new genetic interactions or refine existing ones. We first calculated the scL-score for all genes in all gastruloids, then averaged across gastruloids and clustered the resulting interaction vectors (see Methods for details). The heatmap is shown in Figure 4a (heatmap including cell cycle genes is shown in Figure S4.1a). We noted that just as when we clustered genes associated with NMPs and their direct descendants, genes associated with cell types tended to cluster together. Specifically, NMP, spinal cord, endoderm, and endothelial genes clustered very strongly together, while presomitic mesoderm genes again were split into two groups, one of which was more closely associated with genes involved in early somitogenesis. We quantified how well cell type specific genes clustered compared to a random null by first calculating the dispersion of cell types within the tree topology using cophenetic distance (see Methods), and then permuting the leaves of the tree to create a null distribution of the dispersion expected by random. The results produced by hierarchical clustering on scL-score vectors were significantly (p=0.0001) more clustered than would be expected by chance (Figure S4.2a,b). To assess cluster stability, we randomly selected subsets of the panel and repeated the clustering. Regardless of panel size, the tree produced by clustering on scL-score vectors was always significantly less dispersed than permuted nulls (Figure S4.2c). Although our method of calculating dispersion can only be compared between trees clustered on the same gene set, we noted that as we increased the number of genes, the gap between the dispersion of the real tree and the dispersion of the permuted trees increased (Figure S4.2d), indicating that, as would be expected, better clustering was achieved when more genes were considered.

Clustering L-score vectors clearly resolves cell types and reveals novel genetic interactions.

a. Heatmap of all scL-score values for all genes in the panel (excluding poorly detected and cell cycle genes, see Methods, n=166). Colored bars on the top and right hand sides indicate if a gene is associated with a particular cell type. Heatmap was hierarchically clustered by row. b. Expansion of a presomitic mesoderm cluster (i.). Clustering relationships are indicated with the dendrogram, and summed densities for all genes in an example gastruloid show where the genes in each cluster are expressed spatially. Clustering relationships determined from the full gene set shown in a. c. Expansion of the posterior cell type cluster (ii.). Clustering relationships are indicated with the dendrogram, and summed densities for all genes in an example gastruloid show where the genes in each cluster are expressed spatially. Clustering relationships determined from the full gene set shown in a. d. Expansion of the endothelial cluster (iii.). Clustering relationships are indicated with the dendrogram, and summed densities for all genes in an example gastruloid show where the genes in each cluster are expressed spatially. Clustering relationships determined from the full gene set shown in a. e. Expression of two example pairs of genes from the endothelial cluster: Cldn5 and Tgfb1, and Cldn5 and Sox17.

Spatial L-score reveals tissue-level patterns of gene expression.

a. Illustration of how the density-based L-score (spatial L-score) is calculated, and how the value changes as a function of how much the density estimate is smoothed. b. Clustered heatmap of the spatial L-score for a representative gastruloid. Purple box indicates a mixed cluster of endothelial and endoderm genes. c. An image of the gastruloid used to generate the heatmap in b. Cell types are indicated by color, legend is the same as in b.

Endothelial precursors show unique organization and distinct, spatially-dependent cell states.

a. Representative gastruloid showing two distinct morphologies of endothelial cells: somite-associated (anterior) and endoderm-associated (posterior). b. Exposure index for endothelial cells compared to all other cell types. Dataset-wide average (n=26 gastruloids) shown in the colored bars, grey lines indicate the values for the gastruloid shown in a. c. Exposure index for endoderm cells compared to all other cell types. Dataset-wide average (n=26 gastruloids) shown in the colored bars, grey lines indicate the values for the gastruloid shown in a. d. Genes differentially expressed in somite-associated or endoderm-associated endothelial cells. Bar color represents the cell type associated with the gene (if any); genes that were not previously known to be associated with a cell type in gastruloids are shown in gray. Bars are ordered by significance (adjusted p value) from greatest (Pecam1, adjusted p=8.49e-22) to least (Top2a, adjusted p=7.65e-3). e. Spatial distribution of expression for example endoderm-associated genes shown in d). Each plot shows one gene, each dot is a single transcript, and cells typed as endothelial are outlined in black. f. Spatial distribution of expression for example somite-associated genes shown in d). Each plot shows one gene, each dot is a single transcript, and cells typed as endothelial are outlined in black.

To generate the heatmap shown in Figure 4a and S4.1a, we used the same clustering method as described earlier with the Euclidean distance between scL-score vectors. However, we also tried clustering directly on the scL-scores themselves, by transforming each scL-score from the original [-1,1] scale to a [0,1] distance-like scale using a (1-scL)/2 mapping (Figure S4.3a,b). The results were largely consistent, however the cophenetic distance scale was relatively compressed when the transformed values were used (Figure S4.3a,b). This shallow structure implies that many branches are separated by only modest distances, so fine-scale ordering within the dendrogram should be interpreted more cautiously than the larger-scale cell type block structure. We chose to continue using Euclidean distance-based hierarchical clustering of scL-score profiles, where cell type grouping is observed alongside larger cophenetic separations between clusters.

We examined the largest cluster of presomitic mesoderm genes (Figure 4a, i..). Just as before, the clustering reflected distinct spatial distributions; a highlight of one diagonal block with two subclusters is shown in Figure 4b, along with the expression patterns in a representative gastruloid. The right subcluster has most of the presomitic mesoderm genes, and is particularly enriched in the left side of this gastruloid, despite the fact that there is a cluster of presomitic mesoderm genes on both sides of the central axis (Figure S1.3a, xii). The other subcluster has genes associated with differentiation into somites, and contains Snai1, which is a gene associated with EMT and cells in a more migratory state, consistent with how somitogenesis is thought to occur in gastruloids (van den Brink et al., 2020; Veenvliet et al., 2020; Yamanaka et al., 2023). These genes cluster together because both sets are strongly expressed in a central region of presomitic mesoderm, one is more posteriorly biased, while the other is more anteriorly biased.

The genes associated with more anterior cell types, like somite, paraxial mesoderm, and the differentiation front (which marks the boundary between posterior and anterior), also clustered together in one large block, although the genes associated with the different types tended to be mixed (Figure 4c). When we plotted their spatial distributions, the reason for this separation became clear — one block was genes more strongly expressed in the most anterior of the gastruloid, while the other, including almost all of the differentiation front genes, was expressed more posteriorly; these genes were associated with the onset of somitogenesis. We saw similar spatial patterning in the cluster of NMP genes. One sub-cluster, which contained genes like Wnt3a, Hoxc10, and Evx1, was very specifically far-posteriorly localized, while the other, which contained genes like Fgf8, Cdx4, and some of the shared NMP+presomitic mesoderm genes, extended further into the anterior (Figure S4.1b).

We also observed interesting differences between the subclusters of endothelial genes (Figure 4d). Cldn5 had the strongest average association with other endothelial genes and strong negative associations with genes associated with other cell types, indicating that it is likely the best marker in our panel for generically marking all endothelial precursors. The cluster that contained Cldn5 showed strong intra-cluster associations (high scL-score values), and although the other cluster showed relatively high association with Cldn5, the average association within the cluster and with the genes in the other cluster was weaker. However, this cluster contained some interesting genes: Sox17, an endoderm marker, and Tgfb1 (Tgfβ), which is associated with reduction in inflammation, pro-growth signals, and differentiation (Massagué, 2012). Plotting the expression of these genes in representative gastruloids confirmed their coexpression with Cldn5, a consistent marker for endothelial cells (Figure 4e).

Tgfβ appeared to be exclusively associated with Cldn5-expressing cells, but only in some gastruloids (it was extremely lowly expressed in others and not clearly associated with any specific cell type). Tgfβ loss causes embryonic lethality due to hematopoiesis and vascular development failures (Dickson et al., 1995), specifically in extraembryonic tissues. Extraembryonic endoderm has been inconsistently reported in gastruloids; some studies find evidence of this tissue (van den Brink et al., 2020), while others do not (Rosen et al., 2022). Moreover, Tgfβ can drive the transition from endothelial precursors to hematopoietic stem cells (Monteiro et al., 2016), so it is particularly exciting to see it specifically expressed in endothelial tissue. While we note that the regions expressing Tgfβ have unique morphology (Figure S1.3a, xv..), further work is needed to determine the role this gene plays in the development of these tissues in gastruloids. These results demonstrate that the scL-score can be used to identify novel cell states in subsets of cell types.

To validate the clustering produced by the scL-score, we compared our results to a state-of-the-art method for identifying gene programs in an unbiased fashion from single-cell data: consensus non-negative matrix factorization (cNMF) (Kotliar et al., 2019). We pooled nuclei from all individual gastruloids and ran cNMF. We found that many of the resulting clusters (Figure S4.4a,b) corresponded to the clusters identified when the scL-score tree was truncated to produce exactly the same number of clusters (Figure S4.4c). The similarities were even greater when the scL-score clusters were hand-selected based on visual inspection of the tree and density of marker genes (Figure S4.4d). To quantify the overlap between clusters, we calculated both the Jaccard Index and the Adjusted Rand Index (ARI) between each scL-score cluster (Figure S4.4c) and the most similar cNMF cluster. These distributions are shown in Figure S4.4e (blue). We compared to a bootstrapped null where we permuted the genes found in the scL-score clusters, and found that permuted clusters were far less similar to the cNMF clusters than those derived from the real scL-score tree (Figure S4.4e). From these observations, we conclude that the two methods are capable of producing similar results at a high-level, but are different in their application. Individual cells receive component scores for cNMF gene programs, yielding more per-cell information, while the tree produced by L-score clustering reveals hierarchical information about gene programs, which quantifies their similarity in expression on a global scale.

Thus, by quantifying the similarity between genes based on the mutual exclusivity of their expression with all other genes, we were able to extract similarity in the expression of cell type genes without explicitly specifying them. Moreover, the spatial relationship between genes, including among genes associated with the same cell type, was encoded in the tree of relationships produced by clustering. Finally, although the strong blocks we find in the heatmaps in Figures 4a and S4.1a largely reflect cell types, as is consistent with our panel design, we discovered some novel functions of genes in the panel through their location in the scL-score tree: although Tgfβ was initially included in our panel to generally detect inflammatory and growth signaling, clustering by expression patterns revealed its unique association with endothelial precursors. Together, these data demonstrate the power of mutually exclusive expression relationships to uniquely quantify relationships between genes and the cells that express them.

scL-score analysis reveals cell type groupings and new transcription factor associations in a single-cell RNAseq dataset

To test the generality of scL-score analysis, we analyzed a previously-published dataset from (van den Brink et al., 2020) where individual gastruloids (at the same stage as those used in this study) were pooled and subjected to single-cell RNA-seq analysis. After filtering for cell quality and common gene detection, we calculated the scL-score values using a cell-by-gene table of 14304 cells x 19075 genes.

To first test whether we could reproduce the results from this study, we performed hierarchical clustering on the scL-score difference vector (as previously described) on the set of 207 well-detected genes that were also present in our seqFISH gene panel. The resulting tree showed clustered cell types, similar to the tree produced with the expression data in this study (compare Figure S4.5a to Figure S4.4d). The cell types were clustered significantly more than expected by chance (Figure S4.5b).

Then, to test whether scL-score analysis would be effective for analyzing the entire dataset, we performed hierarchical clustering on all 19075 genes. We then examined the resulting heatmap (Figure S4.5c) for clusters of interest. We observed a cluster enriched for endothelial genes (Figure S4.5d), which contained some genes in our panel but many others which were not; this finding demonstrates that the clustering in Figure 4a is not solely due to the selection of genes in our seqFISH panel. We also observed a large cluster that contained genes associated with pluripotency or primordial germ cell fate (Figure S4.5e). Although some of the genes in this cluster were in our seqFISH panel, when we performed scL-score analysis we did not see them cluster with each other or with any other cell type genes. This lack of clustering implies that in our dataset cells that co-express these genes may be rare or too poorly detected to cluster strongly; however, the same analysis performed with more cells and genes showed association. This result demonstrates that clustering scL-score difference vectors can identify known cell-type associated genes, even within transcriptome-scale data. Finally, we also found a small cluster showing strong coexpression of the transcription factor Gata4, a crucial regulator of the development of visceral and parietal endoderm, and two other genes: a predicted gene of unknown function (Gm43715) and Troponin C (Tnnc1) (Figure S4.5f). Intriguingly, Gata4 has been implicated in heart development (albeit in an indirect manner) (Watt et al., 2004), and troponin C is important for cardiac muscle cell contraction and has been implicated in cardiomyopathy, although at a much later stage of development than that modeled by gastruloids (Li and Hwang, 2015).

Together, these results demonstrate that scL-score analysis is reproducible across datasets, even when different numbers of genes are compared. It effectively clusters genes associated with cell types, and can reveal developmental transitions. Moreover, increasing the number of cells and genes can reveal new clusters, some of which may predict novel regulatory interactions or spatial co-occurrence not previously observed.

Calculating the L-score on smoothed gene expression spatial profiles reveals higher-order tissue organization and unique cell type arrangement in certain gastruloids

Our analysis of mutually exclusive gene expression on a per-cell basis both yielded the tools to identify highly specific marker genes and revealed gene expression states that were characteristic of subsets of our annotated cell types. However, there are examples where graded gene expression in tissue, independent of the individual cells expressing the gene, is biologically meaningful: zonation in tissues such as intestine (Moor et al., 2018), liver (Ben-Moshe and Itzkovitz, 2019) and kidney (Lindgren et al., 2017), and hot and cold areas of tumors (Li et al., 2018). Furthermore, many spatial analysis techniques are not resolved to a single cell. Therefore, we wondered whether we could adapt the scL-score to identify patterns of gene expression at different length scales independent of cellular assignment.

To do so, we first noted that in a spatially-resolved dataset, the assignment of spots to nuclei is essentially a non-linear partitioning of genes in space — by being assigned to a nucleus, an individual expression spot loses its specific spatial location and is binned into a region defined by the cell boundary. An example of this binning is shown on the left side of Figure 5a. We reasoned that we could instead capture relationships between the purely spatial expression of genes by mapping each transcript onto an evenly spaced grid and then smoothing the resulting 2-dimensional distribution using kernel density estimation. The bandwidth of smoothing can be set in any units, but we found the average (per gastruloid) nearest-neighbor distance to be an intuitive and useful length scale to use. Figure 5a shows an example of how choosing different distances for smoothing captures different scales of patterning—any amount of smoothing that causes gene expression in adjacent cells to be averaged would increase the L-score by increasing the degree of coexpression, and more smoothing yields a further increase in L. We chose to use density estimations with a distance value of 2 times the nearest-neighbor average distance, as this spacing seemed likely to capture meso-scale organizational patterns in gastruloids while preserving finer structures of spatial variance (Figure 5a).

The scL-score is calculated by comparing gene expression on a cell-by-cell basis. To adapt this measurement to spatial data, we calculated the L-score using the density estimation at each spatial bin across the entire gastruloid (i.e., each bin is treated the way a cell is treated when the scL-score is calculated). We denote this value as the “spatial L-score”.

We calculated the spatial L-score for all the consistently-detected genes in our panel across all gastruloids (202/220 genes, see Methods). When we averaged across gastruloids and clustered the values, the resulting heatmap (Figure S5.1a) had many properties similar to the scL-score heatmap shown in Figure 4a: For the most part, cell types clustered together; however, we noted a few unique patterns. The order of clustering seemed to more closely reflect overall position in the gastruloid: the more posterior cell types (NMP, presomitic mesoderm, spinal cord) were found close together, then endoderm, differentiation front, and a mixed cluster of paraxial mesoderm, somite, and endothelial genes. This ordering suggests that when the spatially smoothed relationships (spatial L-scores) of all genes in the panel are clustered, the resulting relationships reflect the overall spatial organization of the underlying cell types within the larger tissue structure. Individual cell types appeared to cluster more strongly when using scL-score, but the order and mixing of cell type genes with the spatial L-score reflects tissue-level organization—NMP and presomitic mesoderm genes were close together and mixed, reflecting the mixing we see of the cell types associated with these genes.

The other difference of note was that, unlike in the scL-score heatmap, the endothelial genes were much less strongly clustered, likely reflecting the diversity of locations they were found in within individual gastruloids (Figure S1.2). We looked at their patterns in the heatmaps for individual gastruloids and noticed that there was strong co-clustering of endothelial and endoderm genes in some of the gastruloids. This was hinted at earlier when we calculated significant differences in the interactions between these cell types between gastruloids (Figure S2.1c). An example is shown in Figure 5b, which is the clustered heatmap for an individual gastruloid (cell types and morphology shown in Figure 5c). A strong block of interaction on the diagonal (“endoderm/endothelium cluster”, outlined) indicates a strong association of these genes on a length scale beyond individual cells. When we looked at the cell type arrangement of this gastruloid, there were two clear populations of endothelial cells: those that were located in the anterior and those that were more posteriorly located and strongly intermixed with the endoderm tissue. Thus, the unique clustering behavior of genes in this gastruloid is reflective of a unique, tissue-level pattern that is only found in some of our samples. When we calculate the scL-score heatmap for the same gastruloid, although endothelial and endoderm genes cluster close together, they form two distinct diagonal clusters, compared to the single, high-L block found in the spatial-L heatmap (compare the outlined clusters in Figure 5b and Figure S5.2a).

These comparisons suggested that the spatial L-score could capture patterns on a tunable length scale. When we directly compared the intra-cell type L-score values calculated using both methods (Figure S5.1b), we saw that while the values for some cell types, like NMP, were quite similar, others, such as spinal cord, showed a dramatic shift towards +1 when calculated using a smoothing of 2 nearest neighbors. This shift is consistent with our observation of consistent expression of NMP genes and a more “salt-and-pepper” expression of spinal cord genes, and suggests that by calculating the spatial L-score over many length scales, one could learn the effective length scale of interaction for specific gene pairs.

Endothelial precursors show unique organization and distinct, spatially-dependent cell states

We observed that in 5 out of the 26 gastruloids, there was a large central patch of endoderm cells intermixed with endothelial precursors; these samples also had unique spatial L-score clustering of endothelial and endoderm genes (Figure 5b). An example of one such gastruloid is shown in Figure 6a. Migration to and association with the endoderm is also a hallmark of endothelial development (Dyer and Patterson, 2010; Sendra et al., 2024), and we were curious whether there were differences between these cells and the cells we observed forming anterior, somite-associated clusters. When we computed the cell type exposure index for just this gastruloid, we found that, consistent with our visual observations, in this particular sample endothelial and endoderm cells were much more frequently found next to one another than on average (Figure 6b,c). To determine whether these spatial and organizational differences reflected gene expression differences, we divided the gastruloid normal to the anterior-posterior axis to separate the endothelial cells into endoderm-associated and somite-associated and looked for differentially expressed genes between the two groups in this gastruloid. To ensure we were focused on genes that truly varied in expression in endothelial cells and were not merely a reflection of spillover from surrounding cells, we pre-filtered genes on expression, so only genes that were present in at least 50% of the cells in either group at a greater than 2 count per cell level were considered. The significantly differentially expressed genes after filtering are shown in Figure 6d. As an additional check on the degree to which transcript mis-assignment affected our analysis of gene expression in these cells in particular, we varied the nuclear dilation in this gastruloid specifically, and calculated cell type score entropy as a function of nuclear dilation (Figure S6.1a). Because cell type score entropy of a cell reflects the degree to which that cell specificity expresses genes associated with a single cell type, our expectation was that if spillover between endoderm and endothelial cells was a significant issue, then decreasing the nuclear dilation should greatly decrease the entropy scores for both groups. Although we saw a slight increase in the spread of the distribution as nuclear dilation increased, the median cell type entropy stayed extremely low for both groups (Figure S6.1a). From this analysis we conclude that the genes we identify as differentially expressed are not due to spillover from surrounding cells, but instead are due to spatially-dependent differences in endothelial cell biology.

The genes with the highest fold-change in expression in endoderm-associated endothelial genes are shown on the left hand side of Figure 6d. Two are endothelial genes: Pecam1 and Cdh5, both of which are associated with angiogenesis. Spatial expression of these genes is shown in Figure 6e (larger version in Figure S6.1b). Notch1 is more expressed in endoderm-associated endothelial cells, and this could reflect an increase in Notch signaling in the posterior of the gastruloid. (Chan et al., 2017) demonstrated that Notch signaling can be sensitive to shear stress, raising the possibility that the differences in cell state we observe may be driven by differences in mechanical forces in the anterior and posterior. Although most endothelial cells are thought to be of mesodermal origin, some evidence suggests that, in the organogenesis of specific tissues like the liver, the endoderm can give rise to endothelial cells (Goldman et al., 2014). Furthermore, in (Rossi et al., 2022) the authors show that in a gastruloid-like model specifically designed to model blood development, there is strong spatial adjacency between endothelial and endoderm cells. They hypothesize that these may be a subset of endothelial cells, specifically hemogenic endothelial cells (which have the potential to become blood progenitors). Our data demonstrate a molecularly driven organization distinct from the clustering we observed in the anterior and suggest that multiple mechanisms of endothelial specification could be modeled in gastruloids, even simultaneously within the same structure, although further characterization is needed to determine exactly what processes these unique endodermal/endothelial structures model.

Several other endothelial genes are instead differentially expressed in somite-associated endothelial cells: Nrp2, Tek, Apoe, and Cldn5. Spatial distributions of these genes are shown in Figure 6f (larger version in Figure S6.1b). Although these genes have less obvious functional distinctions than the endoderm-associated genes, Nrp2 enables semaphorin receptor activity, including nervous system development and ventral trunk neural crest cell migration and Tek negatively regulates endothelial cell apoptotic process and response to retinoic acid (RA), which is known to be higher in the gastruloid anterior. Furthermore, a specialized population of endothelial precursors associated with somites was also observed in trunk-like structures, which show more tissue-like organization than gastruloids (Veenvliet et al., 2020).

Although endothelial cells have consistently been observed in single-cell measurements of gastruloids, their relative rarity has precluded in-depth analysis of subtypes or inference of spatial location. Our results strongly suggest that endothelial precursor formation, migration, and organization may all be modeled in 3D gastruloids, even without treatment with additional factors as in (Rossi et al., 2022, 2021); recent advances in 2D gastruloids have allowed modeling of cardiac and hepatic vascularization (Abilez et al., 2025), and our data suggest that 3D gastruloids may similarly be adapted to model more specific aspects of hematopoiesis and vascularization. Early specification from a pool of mesodermal precursors is a hallmark of the endothelial lineage (Sendra et al., 2024); given the consistency with which we observe endothelial precursors, we speculate that this behavior is recapitulated in gastruloids, but further epigenetic measurements are required to validate this hypothesis.

Discussion

A central goal of developmental biology is to map how the identity and arrangement of cells are determined and coordinated. Although precise scaling of gene expression in gastruloids had been previously reported (Merle et al., 2024), the relationship between this scaling and the proportions of specific cell types or more complex tissue organization was not known. Towards that end, we have performed detailed measurements and analyses of the expression of hundreds of genes in tens of individual gastruloids, revealing how cell types are organized within and across gastruloids. Our results highlight that many aspects of embryonic development, such as the arrangement and differentiation trajectories of posterior cell types, are faithfully recreated on a per-gastruloid basis. Others, however, can vary, and conditions may need to be tuned to increase reproducibility if a specific aspect of their development needs to be modeled. For example, we reproduced observed variation in endodermal structures (Farag et al., 2023) and reported both the distinct organization of endothelial cells and variation in that organization between gastruloids. Although we do not fully reproduce the level of per-gastruloid cell type proportion variation reported in a previous single-cell sequencing dataset (Rosen et al., 2022), we do find evidence of a tradeoff between neural and mesoderm tissues; differences in the number of cells analyzed per gastruloid could potentially explain this discrepancy. Moreover, we observe many of the same cell types as in this study, and show that their organization is quite consistent.

We introduce the L-score, which quantifies mutually exclusive gene expression, and highlight two ways of using it that can be employed on different types of data. The cell-based L-score, scL-score, can be applied to any technique that generates a cell by gene table. Hierarchical clustering analysis of the resulting vectors can reveal how similar each gene is to others in terms of how it is co-expressed with all genes, and may reveal spatial information if the sample was collected from a spatially-organized tissue. We were able to verify the spatial information because our data were spatial, but the scL-score is calculated only using the cell-by-gene table, a standard output for any single-cell analysis technique.

The spatial L-score, which operates spatial bin-wise rather than cell-wise, can be used on spatial data even when cell assignment or nuclear segmentation isn’t possible — in principle, it can reveal which genes are patterned similarly at many different length scales, with the fundamental resolution limit set by the measurement technique. It was able to identify coexpression, even amongst lowly-expressed genes, such as the spinal cord markers, that had a more ‘salt-and-pepper’ expression pattern and were only strong markers when considered in aggregate. It is possible that some genes with bursty gene expression may appear not to express together, but spatial smoothing may allow for the identification of coexpression of these genes, which may be useful for cell type identification.

Although the identification of cell types has utility for high-level analyses of organization, our in-depth dissection using the L-score revealed that, particularly for cells in the process of differentiating, a gene-centric view may yield a more holistic picture of tissue organization. It is particularly exciting to imagine these analyses performed in a time-resolved fashion, as dynamic changes in gene expression across a tissue in response to physical rearrangement and broad signaling patterns could reveal mechanisms or spatial aspects of differentiation not visible when cells are merely binned into predetermined categories.

Methods

Cell culture

All gastruloids were generated from E14TG2a cells (a generous gift from the Toettcher lab). Cells were cultured in 2i+LIF media at 37 °C, 5% CO2 and grown on plates coated in 0.1% gelatin (Millipore/Sigma ES006B). 2i+LIF media consisted of GMEM (Gibco 11710-035), 10% stem-cell qualified FBS (R&D Systems S10250), 1x GlutaMAX (Gibco 35050–061), 1x MEM non-essential amino acids (Gibco 11140050), 1 mM sodium pyruvate (Gibco 11360–070), 100 μM 2-mercaptoethanol (Gibco 471 21985–023), and 100 units/mL penicillin/streptomycin (Gibco 15140–122), 1000 units/mL LIF (Millipore Sigma ESG1107), 2 μM PD0325901 (Tocris 4192), and 3 μM CHIR99021 (Tocris 4423). Cells were passaged every 2 days: cells were trypsinized (Life Technologies 12563011), counted using an automated cell counter (Invitrogen Countess II), and 5x10-5 — 8x10-5 cells were transferred to a 25cm2 dish; cells were passaged at least once after thawing before gastruloid aggregation.

Gastruloid aggregation

The aggregation protocol was as described in (van den Brink et al., 2020) (protocol here). Briefly, cells were trypsinized, rinsed twice with 1X PBS to remove media, and then resuspended in N2B27 medium. N2B27 consisted of equal parts neurobasal medium (Gibco 21103049) and DMEM/F12 (Gibco 11320033) supplemented with 1X N2 (Gibco 17502048) and B27 (Gibco 17504-044), 100 units/mL penicillin/streptomycin (Gibco 15140–122), and 100 μM 2-mercaptoethanol. Cells were resuspended at a concentration of 2500-7500 cells/mL; 40 μL of this mixture were placed in each well of round-bottom ultra non-adherent U-bottom 96-well plates (Costar 7007), yielding 100-300 cells/well. The first two seqFISH runs used 300-cell gastruloids (n=8 total) and the last used 100-cell gastruloids (n=18). Cells were allowed to grow at 37 °C, 5% CO2 for 48 hours and then supplemented with 150 μL fresh N2B27+3 μM CHIR99021 (Chiron, Wnt agonist). After 24 hours, 150 μL of media was removed and replaced with N2B27 without Chiron; this media change was repeated daily until the gastruloids reached maturity at 120 hours. All wells were scored for elongation at 120 hours just before fixation. The gastruloids selected for fixation were elongated and uniaxial; the experiments that generated them had an 80-94% success rate (see Figure S1.1a for full distribution).

Sample fixation

After gastruloids had grown for 120 hours, they were harvested by pooling together all successful (i.e., elongated and uniaxial) wells in a plate, washed with 1X PBS, and then fixed for 10 minutes in 4% PFA in PBS. After fixation, gastruloids were washed with 1X PBS and stored in methanol for up to a month before preparation for seqFISH. 3 separate seqFISH runs were performed on 6 different aggregations (one experiment pooled 4 plates of gastruloids), producing a total of 26 individual samples.

Embedding and sample preparation for seqFISH

∼30 gastruloids were placed in a Lo-Bind tube (Eppendorf 0030108523). After removing as much methanol as possible, the gastruloids were washed with 2XSSC, then incubated in 2XSSC + 1 μg/mL DAPI overnight at 4 °C. Stained gastruloids were permeabilized with 1X PBS + 0.2% Tween20, arrayed on a gelatin-coated slide (Spatial Genomics 00200001) with an eyelash tool (Electron Microscopy Sciences 11852), covered in liquid acrylamide gel, and allowed to polymerize for 2 hours. Polymerized samples were washed with 8% SDS, then treated with 8x10-3 units/mL Proteinase K (NEB P8107S). After this wash, the primary probes were hybridized for 16-40 hours, after which the samples were washed and loaded onto a GenePS machine (Spatial Genomics).

Detailed protocol available here.

seqFISH and raw data handling

Sequential rounds of secondary probe addition and imaging were performed in an automated fashion by the instrument (Eng et al., 2019). The GenePS software that controlled the instrument was v1.6.0. The instrument was run with a 1 or 2 μm z-offset on 5-18 individual regions of interest, highlighting each gastruloid. The genes chosen for the seqFISH panel were based on differential expression in single-cell RNA sequencing experiments in references (Rosen et al., 2022; van den Brink et al., 2020) or by consensus in the literature. 200 of the genes were detected via a sequential barcoding scheme (see (Eng et al., 2019) for details), and 20 were serially detected as in traditional smRNA-FISH (Raj et al., 2008). The serial genes tended to vary in quality between gastruloids; genes with high background or in which expression was too high for the spot detection software to confidently assign spots were excluded. The list of excluded genes for each gastruloid is included with the raw data in a file called ‘bad-genes.txt’.

Gastruloid annotation and spot assignment

Detected spots were decoded and exported without nuclear assignment by SG Navigator software version 0.9.0. At this stage, gastruloids without even and consistent expression of the housekeeping gene Eef2, which indicated focus or technical issues or sample variability, were discarded. 5 out of 6 gastruloids were discarded from the first run, and 3 out of 11 in the second run. None of the samples from the third run were discarded. Raw TIFF files of the DAPI images were annotated using the NimbusImage platform, which is available here: https://github.com/arjunrajlaboratory/NimbusImage/ and hosted here: https://www.nimbusimage.com/. Expression of the gene T was used to locate the posterior end and corroborated by consistent differences in nuclear morphology between the posterior and anterior. An AP axis was inferred by drawing a line between that pole and the extreme opposite end of the gastruloid along the center of the gastruloid parallel to the edges and connecting with the furthest extended point to define the anterior pole. Segmentation was performed using cellposeSAM (Pachitariu et al., 2025) on the high-resolution .tiff files of the DAPI-stained nuclei from the 3rd round of hybridization. Once the nuclear masks had been created, spots were assigned to nuclei using a custom software package called SGAnalysis (available here). Nuclei were dilated by 8 pixels (0.86 μm), and all spots that fell within this dilated nuclear area were assigned to that nucleus. In cases where the dilated regions overlapped, spots were assigned to the nucleus for which the edge of the dilated area was furthest away. We were able to resolve 131,112 nuclei total, with an average of 5043/gastruloid. 79,607 nuclei had enough mRNA counts to be typed (mean 3062/gastruloid). Nuclei without enough mRNA were generally small, uniformly bright (in contrast to the usual mottled appearance of mouse nuclei), and lacked expression of the highly-expressed housekeeping gene Eef2, indicating the corresponding cell was likely dead or fragmented. We primarily found these nuclei in the anterior tissues and classified them as “none”-type, excluding them from our analyses unless noted. The average reads/nucleus per gastruloid ranged from 47-80; for typed nuclei the value was 105-265.

Morphological characterization of gastruloids

To assess gastruloid morphology, 4x images taken with an Incucyte, were segmented in Nimbus Image (https://www.nimbusimage.com/). Gastruloids were imaged directly in a 96-well U-bottom dish: 1 plate from 1/19/2024, 1 plate from 9/1/2024, and 4 plates from 4/4/2025 containing 88, 86, 93, 84, 88, and 90 gastruloids respectively. The Nimbus feature ‘Calculate blob metrics’ was used to quantify the following: "solidity”, "rectangularity", "perimeter", "elongation", "eccentricity", "convexity", "circularity", and "area" (code can be found at https://github.com/arjunrajlaboratory/NimbusImage/). These metrics were then decomposed by PCA. All images and annotations can be found at the following link: https://app.nimbusimage.com/#/project/69d3fa8f1be4701f5fab6359.

Cell type scoring

To assign cell type, we used a panel of marker genes for 9 consensus cell types thought to exist in gastruloids at this stage of development: Neuromesodermal precursor (NMP), presomitic mesoderm (PSM), spinal cord precursor (spinal cord), endoderm, differentiation front, cardiac mesoderm, paraxial mesoderm, somitic, and endothelial precursor (endothelium). Genes were chosen based on differential expression in references (Rosen et al., 2022; van den Brink et al., 2020), or consensus in the literature. Each nucleus received a score for each cell type using the score_genes function in scanpy (Wolf et al., 2018), and the highest-scoring cell type category was assigned to that cell. If none of the scores met a universally applied cutoff (0.1), then the cell was assigned ‘none’.

Cell type entropy score

The degree of uncertainty in cell typing was quantified by calculating the entropy of the cell type score distribution. The scores can be negative or zero, so they were first converted to probabilities using a softmax function with temperature scaling: P(i)=exp⁡(s′_i)/Σ□exp⁡(s′j)Wheresi′=(si−max⁡(s))/τ A temperature of 0.1 was used. Once the scores had been converted, the Shannon entropy of the distribution was calculated: H(P)=−∑ip_ilog2⁡(p_i)

Cell type proportion variation

Since the proportion of each cell type in a gastruloid must sum to 1, changes in one cell type will affect at least one other cell type within the same sample. To properly account for this when applying statistical tests, we calculated the center log ratio (CLR)-transformed proportion and tested all pairs of cell types for significant variation. Statistical significance of pairwise correlations between CLR-transformed cell type proportions was assessed using Pearson correlation coefficients with associated p-values. To account for multiple comparisons, p-values were adjusted using the Benjamini-Hochberg procedure to control the false discovery rate (FDR). Correlations with an adjusted p-value < 0.05 were considered statistically significant.

Computation of AP axis location

Two methods of normalization were used to define position along the AP axis. Both use a hand-drawn annotation of the axis from posterior to anterior as inferred from morphology and T-expression. All annotations are available; see Data Availability section below. For the first method (length-based), the drawn AP axis was normalized from 0 to 1. Each nucleus was projected onto the closest point on this normalized line. For the second method, the ends of the axis were still defined as 0 and 1, but rather than uniformly binning the space, a midpoint (0.5) was defined as the 90th percentile of T expression. The total amount of T spots was used, rather than the per-nucleus amount. The length from 0 to the T-midpoint was uniformly binned, as was the length from the midpoint to 1. The same projection method was used to find the AP axis location of each nucleus.

Quantification of Cell Type Spatial Relationships: Exposure index

To characterize the spatial organization of cell types, we computed two related measures: a pairwise exposure index matrix capturing type-specific spatial relationships, and a scalar mixing index summarizing overall spatial integration. For each cell, we identified neighbors as all cells whose centroids fell within a specified radius r of the focal cell’s centroid. We chose a value of r of 16 μm, which gave an average of 5-6 neighbors per cell. We chose this value as it captures local interactions, which was the primary goal of these analyses. For each ordered pair of cell types (s, t), we calculated the exposure index as the proportion of type s cells’ neighbors that are type t: Es→t=Ns→tNs where Ns→t denotes the count of neighbor pairs in which the focal cell is type *s* and the neighbor is type *t*, and Ns denotes the total number of neighbors across all type *s* cells. Each row of the resulting exposure matrix sums to unity and represents a probability distribution over neighbor types for a given source type. To account for differences in cell type abundance, we normalized exposure indices relative to the expectation under random spatial arrangement: Es→tnorm=Es→t−ptpt where pt is the proportion of cells that are type *t*. Normalized values of zero indicate exposure consistent with random mixing, positive values indicate spatial attraction (co-localization), and negative values indicate spatial avoidance, with a minimum of −1 representing complete exclusion.

Quantification of Cell Type Spatial Relationships: Mixing index

To summarize overall spatial integration across all cell types, we computed a mixing index defined as the fraction of neighbor pairs involving different cell types: M=NcrossNtotal where N_cross is the number of neighbor pairs involving cells of different types and N_total is the total number of neighbor pairs. For normalization, we compared the observed mixing to the expectation under random spatial arrangement: Mnorm=M−MexpectedMexpected where Mexpected=1−∑tpt2 is the expected cross-type interaction rate given cell type proportions. Normalized values of zero indicate random spatial mixing, positive values indicate greater integration than expected (hyper-mixing), and negative values indicate spatial segregation.

Exclusion of untyped cells

When computing the mixing index, cells lacking confident type assignments were optionally excluded from both the numerator and denominator, ensuring the measure reflects only spatial relationships among typed cells. These cells were retained in the exposure matrix to quantify how typed cells interact with unclassified cells.

Statistical analysis of variance

To identify cell type pairs whose spatial relationships varied significantly across samples, we computed the variance in exposure indices across samples for each type pair. To account for the expected relationship between mean exposure and variance, we regressed log-variance against log-absolute-mean across all type pairs and computed residuals. Type pairs with residual variance exceeding the 97.5th percentile (two-tailed α = 0.05) were considered significantly variable, indicating spatial relationships that differ across samples beyond what is expected from sampling variation and composition differences.

Triplet identification and enrichment

Spatial enrichment of cell type triplets was assessed using a computational framework based on Delaunay triangulation. Each gastruloid was processed independently to account for potential inter-sample variability in cell density and spatial organization. Spatial neighbors were identified using Delaunay triangulation, which partitions the coordinate space into triangular tessellations connecting nearest neighbors. A distance cutoff of a maximum of 200 pixels was used; neighbors with greater than this distance were not considered. All groups of 3 nearest neighbors were used to evaluate triplet frequency. Statistical significance was evaluated through a permutation testing framework. For each sample, the observed triplet frequencies were compared against null distributions generated through random reassignment of cell type labels while preserving spatial coordinates. The number of permutations was set to 1,000 per sample. The data for all gastruloids were compiled together such that the significance of each triplet in each sample, and how many samples each triplet was found in, could be assessed. Only significantly enriched or depleted triplets were analyzed; significant triplets found in at least half of the gastruloids are shown in Figure 2c.

Kernel density estimation for generating smoothed spatial profiles of gene expression

We performed kernel density estimation (KDE) to generate smoothed spatial profiles of gene expression patterns. Using the nonparametric.KDEmultivariate method from the statsmodels package in Python, a given gene’s KDE of gene expression in a given gastruloid was calculated using a two-dimensional Gaussian kernel for each RNA spot fitted to a bandwidth equal to the average nearest neighbor distance between cell centroids in that gastruloid multiplied by 2 (Table S1) in both dimensions. To produce an expression profile for a gene from its KDE on a given gastruloid, we first partitioned the gastruloid’s annotated hull to produce a grid containing square elements, each with side length equal to half of the average cell diameter, and then evaluated the KDE on the center of each grid element. Visualizations of multiple genes’ combined expression in a given gastruloid were obtained by summing individual genes’ normalized KDE values for each grid element.

Calculating the single-cell L-score

The single-cell L-score (scL-score) was computed for each ordered gene pair (gene A, gene B) within a single gastruloid. The cell-by-gene expression matrix was filtered to retain only cells with at least 2 detected transcripts for both genes. Cells were sorted in descending order by gene A’s expression values; gene A thus serves as the reference distribution, and the score is asymmetric with respect to gene order.

Three reference distributions were constructed for gene B: (1) perfect coexpression, in which gene B’s values were sorted in the same descending order as gene A; (2) perfect mutual exclusivity, in which gene B’s values were sorted in ascending order; and (3) independence, in which every cell was assigned the mean expression value of gene B.

Cumulative sums of expression values were computed for gene A, for the observed expression of gene B, and for each reference distribution. The cumulative sum of each gene B distribution (observed and reference) was then plotted against the cumulative sum of gene A. This cumulative-sum-versus-cumulative-sum representation captures how gene B’s expression accumulates relative to gene A’s: if gene B’s expression is concentrated in the same high-expressing cells as gene A, gene B’s cumulative curve rises steeply at first; if concentrated in opposite cells, the curve rises steeply at the end. The area under each curve was calculated using trapezoidal integration and normalized by the product of gene A’s and gene B’s total expression, yielding four normalized areas: A_observed (observed relationship), A_positive (perfect coexpression), A_negative (perfect mutual exclusivity), and A_uniform (independence). This normalization ensures that scores are comparable across gene pairs with different overall expression levels (Figure S3.2a,b). The scL-score was then defined as follows:

If A_observed > A_uniform: scL-score = (A_observed − A_uniform) / (A_positive − A_uniform), yielding values in (0, 1].

If A_observed = A_uniform: scL-score = 0.

If A_observed < A_uniform: scL-score = −(A_observed − A_uniform) / (A_negative − A_uniform), yielding values in [−1, 0).

A score of +1 indicates perfect coexpression, −1 indicates perfect mutual exclusivity, and 0 indicates independence.

The scL-score was computed for all gene pairs in each gastruloid from the 05/07/2025 dataset (n = 18 gastruloids). The other two datasets (n = 8 gastruloids) were excluded due to lower transcript detection quality. To generate an average scL-score matrix, the analysis was restricted to a common set of 202 genes well-detected across all 18 gastruloids, and per-gastruloid matrices were averaged. Unless otherwise noted, a symmetrized scL-score was used: scL-score_sym(A, B) = [scL-score(A, B) + scL-score(B, A)] / 2.

Calculating the spatial L-score

The spatial L-score extends the scL-score to spatial regions. For each gene, a kernel density estimate (KDE) was fitted over all detected transcript spots and evaluated on a regular square grid spanning the gastruloid. Bin side length was set to twice the median nearest-neighbor distance between detected spots, calculated separately for each gastruloid. Spatial bins were ranked by KDE-derived density and processed identically to the scL-score calculation. Low-density bins were not filtered, as KDE smoothing produced non-zero density values throughout the imaging area. The spatial L-score was symmetrized as for the scL-score, except when displaying asymmetric heatmaps.

Hierarchical clustering of L-score matrices

Gene-gene distances were defined as Euclidean distances between L-score vectors. Agglomerative hierarchical clustering was performed using Ward’s linkage criterion (scipy.cluster.hierarchy.linkage, method=’ward’, metric=’euclidean’). This approach operates on L-score vector differences rather than on pairwise L-score values directly, and therefore does not require the L-score itself to satisfy the properties of a mathematical distance metric; the Euclidean distance between L-score vectors is non-negative and symmetric by construction, satisfying the requirements of Ward’s method. Heatmaps display pairwise L-score values, not vector distances.

We applied this clustering procedure to the following gene sets:

A subset of NMP, presomitic mesoderm, and spinal cord marker genes in one gastruloid (n = 36 genes; Figure 3g).

All well-detected genes excluding cell cycle genes, averaged across all gastruloids (n = 166 genes; Figure 4 and Figure S4.3a).

All well-detected genes common to all gastruloids, averaged across gastruloids (n = 202 genes; Figures S4.1a, S4.3b, S5.1a).

All well-detected genes excluding cell cycle genes in one example gastruloid (n = 171 genes; Figures 5b, S5.2a).

All well-detected genes common between our seqFISH panel and those that were detected in > 3 cells in scRNA-seq data from (van den Brink et al., 2020) (n=207 genes; Figure S4.5a).

All genes detected in > 3 cells in scRNA-seq data from (van den Brink et al., 2020) (n=19075 genes, Figure S4.5c-f).

In Figure S4.3a,b a transformation of the L-score values was used to cluster genes. The scL-scores were averaged across gastruloids as described above, and then each pairwise scL-score was transformed to a distance-like value with(1 - scL)/2. The matrix was then symmetrized as described previously. Agglomerative hierarchical clustering was performed directly on this transformed gene-gene distance matrix using Ward’s linkage criterion (scipy.cluster.hierarchy.linkage, method=’ward’, metric=’euclidean’).

Cell type analysis via unsupervised clustering

We used scanpy (Wolf et al., 2018) to process the dataset from 05/07/2025. All nuclei were pooled together, nuclei with less than 40 total gene counts were filtered out, and then PCA (50 PCs) and UMAP embedding were performed using the standard pipeline. We performed Leiden clustering using scanpy.tl.leiden, and used visual inspection to choose a resolution at which to cluster. A resolution of 0.8 produced fairly well-separated clusters and roughly corresponded to the expected number of cell types. Clustering results and spatial projection are shown in Figure S4.4a.

Filtering marker genes

We re-scored nuclei using a refined marker gene panel. We retained only genes that met the following criteria: more than 10% of all scL-score values < -0.8 (73% of all genes) and more than 12% of all scL-score values > -0.3 (90% of all genes). 70% of the genes in our panel met both. Scores with this filtered gene panel are shown in Figure S3.5a.

Assessing marker gene panels and cluster-associated genes

To visually represent the relatedness of the genes either differentially expressed in clusters identified with unsupervised clustering, or the groups of genes associated with cell types in a marker gene panel, we calculated the average L-score of all possible combinations where the first gene in the L-score score belonged to the first type or cluster, and the second gene belonged to the second group or cluster. In the case of the marker gene panel, we first filtered genes on expression level, considering only genes that had > 2.3 counts / nucleus in nuclei where they were expressed; this filtering retained 84% of the genes in the panel.

Using cNMF to find gene clusters

All nuclei were pooled together, and cNMF (Kotliar et al., 2019) was run on all nuclei and all non cycle-genes. The default parameters were used and 5-10 components were calculated. 7 components produced the highest stability, and the top 24 most positively differential (by ordered z-score) genes were used for visualization of qualitative clusters. To quantify the overlap between the scL-score derived clusters and cNMF clusters, we computed the pairwise overlap between every scL-score cluster and every cNMF cluster and quantified each comparison using the Jaccard Index and Adjusted Rand Index. For each scL-score cluster, we used the maximum value observed across its 7 possible cNMF cluster comparisons, thereby capturing the strongest correspondence between each scL cluster and the cNMF-defined programs for a given metric. To establish a baseline for these overlap measures, we designed a reference simulation by preserving the same cNMF clusters while defining a permuted set of scL clusters obtained by randomly assigning genes to clusters of the same number (7 clusters) and set of sizes (average 24 genes) as the scL clusters. As above, for each permuted scL cluster and each metric, we retained only the maximum overlap value across 7 possible cNMF cluster comparisons. We note that under this framework, the same cNMF cluster can serve as the highest-overlap comparison for more than one scL cluster.

Calculating the spatial L-score

The spatial L-score was computed analogously to the scL-score except that rather than rank ordering the expression of genes in individual cells, we instead fitted a KDE over the detected spots for a given gene and evaluated it on a square grid of ’spatial bins’ that spanned the entire gastruloid (as described above), then rank-ordered these spatial bins by the resulting probability densities of expression. We treated these bins the same way we had treated cells in the previous calculation, so that the resulting L value reflected mutually exclusive expression in space rather than per-nucleus. The length scale of the interactions considered was determined by the chosen size of the spatial bin. Throughout, we report values for bins at 2 times the nearest-neighbor distance (which was calculated on a per-gastruloid basis). We did not filter low-expressing bins the way we had for cells for 2 reasons: 1) we calculated the KDE on all spots detected, not just those that fell within a nuclear area, so there were fewer areas of low expression, and 2) smoothing resulted in a non-zero density throughout the imaging area. Just as for the scL-score, we symmetrized the spatial L-score by averaging the 2 values calculated for each gene pair, except when displaying the values as a heatmap.

Evaluation of cell type specific gene clustering in scL-score trees

To determine how well genes associated with specific cell types clustered within a given tree produced by clustering L-score vectors, we first created a measure that we called ‘cell type dispersion’. For all the genes associated with a given cell type, we calculate the average pairwise cophenetic distance within the tree, with the expectation that a tree topology that closely clusters cell type specific genes will produce a lower value than one which spreads cell type specific genes widely among the leaves. To summarize a specific tree, values were averaged across cell types to produce a single value. To test for significance, a permutation-based bootstrap test was performed with 10,000 permutations. p < 0.01 was considered significant.

Assessment of L-score cluster robustness

To quantify how robust the clusters produced by the L-score were to which genes were used to cluster, we randomly selected subsets of the panel from 60-180 genes, and performed the same gene dispersion analysis described in the section titled ‘Evaluation of cell type specific gene clustering in scL-score trees’. The true values and distributions were compared for each subset, with the full panel of 202 genes well-detected in all gastruloids included as a reference.

Analysis of single-cell RNA-seq data

Single-cell RNA-seq data from (van den Brink et al., 2020) of pooled 120 hour gastruloids made from the LfngT2AVenus cell line were downloaded from GEO (GSE123187). 4 individual samples were pooled together to generate a dataset of 19075 genes expressed in 14304 cells. Cells with <6000 detected genes were removed, as were genes detected in <3 cells. Cells with >5% mitochondrial counts were also discarded. After this filtering, scL-score values were calculated as previously described, and hierarchical clustering was performed using the Euclidean Distance between scL-score vectors for each gene pair as a distance measure.

This clustering was first performed only using genes common to the panel in this work (207), and then on the entire dataset. Clusters along the diagonal of the resulting matrix were chosen by hand.

Benchmarking

To demonstrate the effects of noise and expression levels, and to compare to existing methods of calculating exclusivity and coexpression, five scenarios were simulated: (1) two ubiquitously expressed genes with similarly high mean expression, (2) two ubiquitously expressed genes with differing mean expression, (3) two ubiquitously expressed genes with similarly low mean expression, (4) two genes coexpressed across a subset of cells, and (5) two genes expressed in generally opposite subsets of cells. For scenarios 1–3, two genes were independently sampled across 50 cells from Poisson distributions with mean expression set to 100 or 50 (high or low average expression, respectively), with no expression bias toward any cell subset. For scenarios 4 and 5, gene 1 was sampled across 50 cells from a Poisson distribution with mean expression set to 4; gene 2 was then generated by sampling from cell-specific Poisson distributions parameterized to either generally match or oppose the transcript count of gene 1 in each cell.

The scL-score, Exclusive Expression Index (EEI) (Nakajima et al., 2021), and COEX (coexpression coefficient, as calculated by the COTAN package) (Galfrè et al., 2021) were determined for the gene pair in each simulation. As further comparison on real data, the three values were also calculated on 4 representative gene pairs from 2025-05-07_roi2 (Pax6-Eogt, Rfx4-Eogt, Nkx1-2-Rfx4, Cdx4-Cdx2).

Differential gene expression in endothelial subtypes

Nuclei associated with subpopulations were defined spatially by defining a horizontal cutoff along the AP axis in one gastruloid. The cells typed as endothelial above and below this line were compared to one another. Gene expression values were normalized to Eef2 expression (with a pseudocount of 1% of the mean Eef2 expression added to avoid division by zero). Genes were retained for testing if they were detected (≥2 counts) in at least 50% of cells in one or both groups. Differential expression between groups was assessed using Welch’s t-test on the normalized expression values, and p-values were corrected for multiple comparisons using the Benjamini-Hochberg procedure to control the false discovery rate. Genes with an adjusted p-value <

0.05 were considered differentially expressed. Log2 fold changes were calculated as the ratio of mean normalized expression between groups.